---
title: "Verifying the Data-Generating Process"
subtitle: "Comparing simulated ground-truth x decision crosstabs to an Ames-II-family study"
bibliography: ../output/paper/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
    code-fold: true
---

# Goal

Before building the two case studies for the talk, we want to check that the
data-generating process (DGP) behind `icfis/app.R` — at its current default
settings — produces data that look at least roughly like a real black-box
firearms study, not just internally-consistent simulated data. This document
generates 100 replicate studies under the app's defaults and compares the
resulting **ground truth x decision** crosstab (the same quantity now shown
on the app's Summary tab) to numbers reconstructed from a real Ames-II-family
study.

This is a sanity check, not a formal calibration exercise. If the simulated
proportions are wildly different from the benchmark, that is the concrete
starting point for tuning the DGP's baseline rates and/or heterogeneity
parameters later.

# The benchmark: reconstructing a real crosstab from Cuellar et al. (2024)

We do not have the raw data from any Ames-II-family study on hand, so we
cannot build a benchmark crosstab directly. What @cuellar2024methodological
*do* report (Section 5.4, discussing Flaw D) is the result of a bounding
exercise applied to the cartridge-case data from Monson, Smith, and Bajic
(2023a):

> "applying this procedure to the cartridge case data from Monson, Smith,
> and Bajic (2023a) result in false positive error rates of 51.4% and 0.92%,
> respectively; the corresponding false negative error rates are 1.76% and
> 25.6%, respectively." [@cuellar2024methodological, p. 8]

The "procedure" recodes every inconclusive response first as an
identification, then as an elimination, and recomputes the error rate each
way. That is enough information to back out the full three-way split of
responses within each ground-truth category, since:

- **Different-source pairs** (denominator for the false positive rate):
  - Inconclusive -> identification: $\text{FPR} = (m_{id} + m_{inc}) / n_{diff} = 51.4\%$
  - Inconclusive -> elimination: $\text{FPR} = m_{id} / n_{diff} = 0.92\%$
- **Same-source pairs** (denominator for the false negative rate):
  - Inconclusive -> identification: $\text{FNR} = m_{elim} / n_{same} = 1.76\%$
  - Inconclusive -> elimination: $\text{FNR} = (m_{elim} + m_{inc}) / n_{same} = 25.6\%$

```{r reconstruct-benchmark}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)

# Different-source pairs
diff_identification <- 0.0092
diff_id_plus_inc <- 0.514
diff_inconclusive <- diff_id_plus_inc - diff_identification
diff_elimination <- 1 - diff_id_plus_inc

# Same-source pairs
same_elimination <- 0.0176
same_elim_plus_inc <- 0.256
same_inconclusive <- same_elim_plus_inc - same_elimination
same_identification <- 1 - same_elim_plus_inc

ames_benchmark <- tribble(
  ~ground_truth_label, ~response,         ~benchmark_proportion,
  "different-source",  "identification",  diff_identification,
  "different-source",  "inconclusive",    diff_inconclusive,
  "different-source",  "elimination",     diff_elimination,
  "same-source",       "identification",  same_identification,
  "same-source",       "inconclusive",    same_inconclusive,
  "same-source",       "elimination",     same_elimination
)

knitr::kable(
  ames_benchmark %>%
    mutate(benchmark_proportion = percent(benchmark_proportion, accuracy = 0.1)),
  col.names = c("Ground truth", "Decision", "Reconstructed proportion")
)
```

This lines up with the false-positive point estimate reported elsewhere in
the same paper — "the point estimate of the false positive error rate for
cartridge cases in the Monson, Smith, and Bajic (2023a) study is 0.933%"
[@cuellar2024methodological, p. 9] — which is within rounding of the 0.92%
used above, so the two reported figures corroborate each other.

**Caveats on this benchmark, to keep in view for the rest of this document:**

- It is *reconstructed* from a bounding illustration, not copied from a
  results table in the primary source. If we get access to Monson, Smith,
  and Bajic (2023a) directly, this should be replaced with the real
  crosstab.
- It reflects one study (cartridge-case comparisons only, one Ames-II-family
  analysis) — not a general "true" rate for firearms examination.
- We do not know $n_{diff}$ or $n_{same}$ (the number of different-source
  and same-source test cases), so we can only compare **proportions**, not
  counts, and we cannot run a formal statistical test against this
  benchmark.

# The data-generating process

The model is implemented once, in `icfis/R/simulate_dgp.R`, and sourced
from here, from `icfis/app.R`, and from the case-study documents. (Earlier
drafts of this document copied the functions inline and needed to be kept
in sync with `app.R` by hand — that's what this refactor removes.)

```{r define-dgp}
source("R/simulate_dgp.R")

# Keep a copy of the shared file's canonical (already-retuned) defaults
# before this document overrides them below with the original,
# pre-calibration values -- used later to sanity-check that this document's
# derivation still lands on the same numbers as the shared file.
canonical_defaults <- dgp_defaults
canonical_incomplete_defaults <- incomplete_defaults
```

## Defaults: starting from the original, pre-calibration values

`R/simulate_dgp.R`'s `dgp_defaults` has already been retuned (see the
"Retuning the baseline rates" section below). This document exists to show
*how* we got there, so from here on it deliberately overrides those
canonical defaults with the original, naive starting values and works
forward from them.

```{r set-defaults}
dgp_defaults <- list(
  n_examiners = 50,
  n_comparisons = 100,
  match_rate = 0.5,
  false_positive_rate = 0.02,
  false_negative_rate = 0.05,
  inconclusive_rate = 0.10,
  examiner_skill_sd = 0.7,
  examiner_inconclusive_sd = 0.35,
  question_sd = 0.8
)

knitr::kable(
  tibble(parameter = names(dgp_defaults), value = unlist(dgp_defaults)),
  col.names = c("Parameter", "Default value")
)
```

# Simulating 100 studies: fully-crossed design

Each replicate is an independent draw of a full study (50 examiners x 100
comparisons = 5,000 responses) under the defaults above, using a distinct
seed per replicate. This assumes every examiner evaluates every comparison
— i.e., a fully-crossed design. **The real study did not use this design**
(see the next section), so treat this as a reference point, not the final
comparison to the benchmark.

```{r simulate-100-studies, cache=TRUE}
# study_crosstab() comes from R/simulate_dgp.R (sourced above).
n_replications <- 100
set.seed(2026)
replicate_seeds <- sample.int(1e6, n_replications)

sim_studies <- purrr::map_dfr(seq_len(n_replications), function(r) {
  study_data <- generate_sim_data(
    seed = replicate_seeds[r],
    n_examiners = dgp_defaults$n_examiners,
    n_comparisons = dgp_defaults$n_comparisons,
    match_rate = dgp_defaults$match_rate,
    false_positive_rate = dgp_defaults$false_positive_rate,
    false_negative_rate = dgp_defaults$false_negative_rate,
    inconclusive_rate = dgp_defaults$inconclusive_rate,
    examiner_skill_sd = dgp_defaults$examiner_skill_sd,
    examiner_inconclusive_sd = dgp_defaults$examiner_inconclusive_sd,
    question_sd = dgp_defaults$question_sd
  )

  study_crosstab(study_data, replicate_id = r)
})
```

# Results: fully-crossed design

## Across-study summary

For each of the six crosstab cells, this summarizes the proportion across
the 100 replicate studies (each study contributes one proportion per cell),
and also pools all 100 studies together into one combined crosstab (500,000
responses) for a single-number comparison to the benchmark.

```{r summarize-results}
sim_summary <- sim_studies %>%
  group_by(ground_truth_label, response) %>%
  summarise(
    mean_proportion = mean(proportion),
    sd_proportion = sd(proportion),
    min_proportion = min(proportion),
    max_proportion = max(proportion),
    .groups = "drop"
  )

pooled_crosstab <- sim_studies %>%
  group_by(ground_truth_label, response) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  group_by(ground_truth_label) %>%
  mutate(pooled_proportion = total_n / sum(total_n)) %>%
  ungroup() %>%
  select(-total_n)

comparison_table <- sim_summary %>%
  select(ground_truth_label, response, mean_proportion) %>%
  left_join(pooled_crosstab, by = c("ground_truth_label", "response")) %>%
  left_join(ames_benchmark, by = c("ground_truth_label", "response")) %>%
  mutate(
    difference = pooled_proportion - benchmark_proportion,
    ground_truth_label = factor(
      ground_truth_label,
      levels = c("same-source", "different-source")
    ),
    response = factor(
      response,
      levels = c("identification", "inconclusive", "elimination")
    )
  ) %>%
  arrange(ground_truth_label, response)

knitr::kable(
  comparison_table %>%
    mutate(across(c(mean_proportion, pooled_proportion, benchmark_proportion, difference), ~ percent(.x, accuracy = 0.1))),
  col.names = c("Ground truth", "Decision", "Mean (100 studies)", "Pooled (500k responses)", "Ames-II-family benchmark", "Pooled - benchmark")
)
```

## Simulated distribution vs. benchmark

```{r plot-comparison, fig.width=9, fig.height=5}
sim_studies_plot <- sim_studies %>%
  mutate(
    ground_truth_label = factor(
      ground_truth_label,
      levels = c("same-source", "different-source")
    ),
    response = factor(
      response,
      levels = c("identification", "inconclusive", "elimination")
    )
  )

ames_benchmark_plot <- ames_benchmark %>%
  mutate(
    ground_truth_label = factor(
      ground_truth_label,
      levels = c("same-source", "different-source")
    ),
    response = factor(
      response,
      levels = c("identification", "inconclusive", "elimination")
    )
  )

ggplot(sim_studies_plot, aes(x = response, y = proportion)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_point(
    data = ames_benchmark_plot,
    aes(y = benchmark_proportion),
    color = "#D55E00",
    shape = 18,
    size = 4
  ) +
  facet_wrap(~ ground_truth_label) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Simulated ground truth x decision proportions across 100 studies",
    subtitle = "Boxplots: 100 simulated studies at current app defaults. Diamond: Ames-II-family benchmark",
    x = "Decision",
    y = "Proportion within ground-truth category"
  ) +
  theme_minimal()
```

# Incorporating the study's actual design: incomplete assignment

The fully-crossed design above isn't what the real study did. Examiners
were randomly assigned roughly 30 comparisons each, drawn from a much
larger pool of about 300 comparisons — not every examiner evaluating every
item. That's a form of *designed* missingness: it's built into the study
design on purpose (unlike Flaw F's missing-not-at-random dropout, which
happens after assignment), but it still means most examiner-item pairs are
never observed.

The response model itself doesn't change — a pair that gets assigned is
generated exactly as before. Only the *assignment* mechanism changes: each
examiner now gets a random subset of the item pool instead of the whole
pool. `generate_sim_data_incomplete()`, already sourced from
`R/simulate_dgp.R` above, implements this.

## Defaults for the incomplete design

Same examiner panel size and rate/heterogeneity parameters as before; only
the item pool size and per-examiner assignment are new.

```{r set-defaults-incomplete}
incomplete_defaults <- modifyList(
  dgp_defaults,
  list(n_comparisons = 300, items_per_examiner = 30)
)

knitr::kable(
  tibble(parameter = names(incomplete_defaults), value = unlist(incomplete_defaults)),
  col.names = c("Parameter", "Default value")
)
```

## Simulating 100 studies under the incomplete design

Each replicate now produces 50 examiners x 30 items = 1,500 responses (not
5,000), since each examiner only sees a fraction of the pool.

```{r simulate-100-studies-incomplete, cache=TRUE}
set.seed(2026)
replicate_seeds_incomplete <- sample.int(1e6, n_replications)

sim_studies_incomplete <- purrr::map_dfr(seq_len(n_replications), function(r) {
  study_data <- generate_sim_data_incomplete(
    seed = replicate_seeds_incomplete[r],
    n_examiners = incomplete_defaults$n_examiners,
    n_comparisons = incomplete_defaults$n_comparisons,
    items_per_examiner = incomplete_defaults$items_per_examiner,
    match_rate = incomplete_defaults$match_rate,
    false_positive_rate = incomplete_defaults$false_positive_rate,
    false_negative_rate = incomplete_defaults$false_negative_rate,
    inconclusive_rate = incomplete_defaults$inconclusive_rate,
    examiner_skill_sd = incomplete_defaults$examiner_skill_sd,
    examiner_inconclusive_sd = incomplete_defaults$examiner_inconclusive_sd,
    question_sd = incomplete_defaults$question_sd
  )

  study_crosstab(study_data, replicate_id = r)
})
```

## Results: incomplete design

```{r summarize-results-incomplete}
sim_summary_incomplete <- sim_studies_incomplete %>%
  group_by(ground_truth_label, response) %>%
  summarise(
    mean_proportion = mean(proportion),
    sd_proportion = sd(proportion),
    min_proportion = min(proportion),
    max_proportion = max(proportion),
    .groups = "drop"
  )

pooled_crosstab_incomplete <- sim_studies_incomplete %>%
  group_by(ground_truth_label, response) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  group_by(ground_truth_label) %>%
  mutate(pooled_proportion = total_n / sum(total_n)) %>%
  ungroup() %>%
  select(-total_n)

comparison_table_incomplete <- sim_summary_incomplete %>%
  select(ground_truth_label, response, mean_proportion) %>%
  left_join(pooled_crosstab_incomplete, by = c("ground_truth_label", "response")) %>%
  left_join(ames_benchmark, by = c("ground_truth_label", "response")) %>%
  mutate(
    difference = pooled_proportion - benchmark_proportion,
    ground_truth_label = factor(
      ground_truth_label,
      levels = c("same-source", "different-source")
    ),
    response = factor(
      response,
      levels = c("identification", "inconclusive", "elimination")
    )
  ) %>%
  arrange(ground_truth_label, response)

knitr::kable(
  comparison_table_incomplete %>%
    mutate(across(c(mean_proportion, pooled_proportion, benchmark_proportion, difference), ~ percent(.x, accuracy = 0.1))),
  col.names = c("Ground truth", "Decision", "Mean (100 studies)", "Pooled (150k responses)", "Ames-II-family benchmark", "Pooled - benchmark")
)
```

```{r plot-comparison-incomplete, fig.width=9, fig.height=5}
sim_studies_incomplete_plot <- sim_studies_incomplete %>%
  mutate(
    ground_truth_label = factor(
      ground_truth_label,
      levels = c("same-source", "different-source")
    ),
    response = factor(
      response,
      levels = c("identification", "inconclusive", "elimination")
    )
  )

ggplot(sim_studies_incomplete_plot, aes(x = response, y = proportion)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_point(
    data = ames_benchmark_plot,
    aes(y = benchmark_proportion),
    color = "#D55E00",
    shape = 18,
    size = 4
  ) +
  facet_wrap(~ ground_truth_label) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Simulated proportions across 100 studies, incomplete design",
    subtitle = "Boxplots: 100 simulated studies, 30 of 300 items per examiner. Diamond: Ames-II-family benchmark",
    x = "Decision",
    y = "Proportion within ground-truth category"
  ) +
  theme_minimal()
```

## Does the design itself change the picture?

The response model is identical between the two designs, so any difference
in the *pooled* (bias) estimates should be sampling noise, not a systematic
shift. What should change is the *spread* across the 100 replicate studies
— each incomplete-design study has 1,500 responses instead of 5,000, so
individual-study estimates should be noticeably less stable.

```{r compare-designs}
design_comparison <- comparison_table %>%
  select(ground_truth_label, response, fully_crossed_pooled = pooled_proportion) %>%
  left_join(
    comparison_table_incomplete %>%
      select(ground_truth_label, response, incomplete_pooled = pooled_proportion),
    by = c("ground_truth_label", "response")
  ) %>%
  left_join(ames_benchmark, by = c("ground_truth_label", "response"))

knitr::kable(
  design_comparison %>%
    mutate(across(c(fully_crossed_pooled, incomplete_pooled, benchmark_proportion), ~ percent(.x, accuracy = 0.1))),
  col.names = c("Ground truth", "Decision", "Fully-crossed (pooled)", "Incomplete (pooled)", "Ames-II-family benchmark")
)

sd_comparison <- sim_summary %>%
  select(ground_truth_label, response, sd_fully_crossed = sd_proportion) %>%
  left_join(
    sim_summary_incomplete %>% select(ground_truth_label, response, sd_incomplete = sd_proportion),
    by = c("ground_truth_label", "response")
  )

knitr::kable(
  sd_comparison %>%
    mutate(across(c(sd_fully_crossed, sd_incomplete), ~ percent(.x, accuracy = 0.1))),
  col.names = c("Ground truth", "Decision", "SD across studies (fully-crossed)", "SD across studies (incomplete)")
)
```

# Interpretation

## Bias: the design doesn't change which benchmark gap we need to fix

The incomplete design's pooled proportions land close to the fully-crossed
design's pooled proportions (see the comparison table above) — as expected,
since the same per-pair response model generates both. So the benchmark
mismatch identified with the fully-crossed design is not an artifact of
that (unrealistic) design choice; incorporating the real assignment
mechanism doesn't fix it. Using the pooled incomplete-design proportions:

- **Inconclusive rate is far too low, by a wide margin.** Simulated:
  11.4% (different-source) / 11.6% (same-source). Benchmark: 50.5%
  (different-source) / 23.8% (same-source). The simulation is missing
  roughly 4/5 of the different-source inconclusive rate and about half of
  the same-source inconclusive rate. This is the single biggest mismatch
  and is worth prioritizing before this DGP is used to argue anything about
  inconclusive-handling (Flaw D) specifically.
- **False positive rate is too high by roughly 3.5x.** Simulated 3.2% vs.
  benchmark 0.9%, even though the *baseline* parameter is
  `false_positive_rate = 0.02` (2%) — already above the benchmark before
  heterogeneity is even added. The realized rate is higher still because
  the logit link means the baseline parameter and the realized marginal
  rate are not the same thing (Jensen's inequality: averaging
  `plogis(logit(p) + noise)` over symmetric noise pulls the mean above `p`)
  — see the talk's calibration slide.
- **False negative rate is too high by roughly 4.3x.** Simulated 7.7% vs.
  benchmark 1.8%, again starting from a baseline parameter
  (`false_negative_rate = 0.05`) already well above the benchmark.

**Bottom line:** all three quantities are biased in the same direction —
too many errors, too few inconclusives — relative to this benchmark, and
that bias is essentially unchanged by moving from the fully-crossed design
to the real incomplete design. Before using these defaults in the talk's
case studies, `inconclusive_rate` should be raised substantially, and
`false_positive_rate` / `false_negative_rate` lowered, with the
baseline-vs-realized gap (Jensen's inequality) accounted for rather than
assumed away. This document is set up to make that check repeatable —
rerunning it after changing `dgp_defaults` (and correspondingly `app.R`)
will show whether the gap closes.

## Variance: the design does matter here

The SD-across-studies table shows a modest but consistent increase in
study-to-study variability under the incomplete design — e.g., the
different-source inconclusive-rate SD rises from 1.1% (fully-crossed) to
1.4% (incomplete), and the different-source elimination-rate SD from 1.4%
to 1.7%. That's smaller than the roughly 1.8x increase a naive
$\sqrt{5000/1500}$ binomial-sampling argument would predict, likely because
examiner- and item-level heterogeneity (which don't shrink just because
fewer pairs are observed per study) already dominate the variance in the
fully-crossed case. Still, the direction is right: a sparser design produces
less stable single-study estimates, which is exactly the argument the
talk's sample-size case study (Flaw A) will want to make quantitatively —
this incomplete-design machinery is a natural building block for that.

# Retuning the baseline rates

## Why an exact fit isn't possible

The model has three scalar baseline-rate parameters
(`false_positive_rate`, `false_negative_rate`, `inconclusive_rate`), but the
benchmark effectively specifies four largely-independent targets: the
false positive rate, the false negative rate, and *two* inconclusive
rates — one for different-source pairs (50.5%) and one for same-source
pairs (23.8%). `inconclusive_probability` in `simulate_response()` doesn't
condition on ground truth at all, so the model structurally cannot produce
different inconclusive rates for same- vs. different-source pairs no matter
how `inconclusive_rate` is set — only one shared value is possible.

Given that ceiling, we searched for the baseline rates that come closest
to all four targets at once, accepting that the two inconclusive-rate
targets can only be compromised between, not matched individually. (We
decided against extending the model to give same-source and
different-source pairs separate inconclusive baselines, to avoid changing
the response-probability equation already written into the talk — see the
Interpretation below for exactly how large the resulting gap is.)

```{r calibration-search, cache=TRUE}
# A vectorized reimplementation of simulate_response(), used only for this
# search. Mathematically identical to the row-wise version defined above --
# vectorizing just avoids per-row function-call overhead so the optimizer
# can evaluate many candidate parameter sets quickly. The row-wise version
# above (matching icfis/app.R) is what actually generates the studies used
# everywhere else in this document.
simulate_responses_vec <- function(ground_truth, decision_challenge, examiner_inconclusive_tendency,
                                   false_positive_rate, false_negative_rate, inconclusive_rate) {
  error_probability <- ifelse(
    ground_truth == 1,
    plogis(safe_logit(false_negative_rate) + decision_challenge),
    plogis(safe_logit(false_positive_rate) + decision_challenge)
  )
  inconclusive_probability <- plogis(safe_logit(inconclusive_rate) + 0.5 * decision_challenge + examiner_inconclusive_tendency)
  total_probability <- error_probability + inconclusive_probability
  scaling_factor <- ifelse(total_probability >= 0.95, 0.95 / total_probability, 1)
  error_probability <- error_probability * scaling_factor
  inconclusive_probability <- inconclusive_probability * scaling_factor

  draw <- runif(length(ground_truth))
  error_label <- ifelse(ground_truth == 1, "elimination", "identification")
  correct_label <- ifelse(ground_truth == 1, "identification", "elimination")
  ifelse(
    draw < error_probability, error_label,
    ifelse(draw < error_probability + inconclusive_probability, "inconclusive", correct_label)
  )
}

# One large, fixed-seed, fully-crossed dataset per candidate parameter set.
# Fully-crossed (not incomplete) is fine here since we already showed design
# doesn't affect the pooled/bias estimates, only their variance.
simulate_large_reference <- function(false_positive_rate, false_negative_rate, inconclusive_rate,
                                     n_examiners = 300, n_comparisons = 300, seed = 42) {
  set.seed(seed)

  comparison_set <- tibble(
    question_id = seq_len(n_comparisons),
    ground_truth = rbinom(n_comparisons, 1, dgp_defaults$match_rate),
    question_difficulty = rnorm(n_comparisons, mean = 0, sd = dgp_defaults$question_sd)
  )
  examiner_panel <- tibble(
    examiner_id = paste0("E", seq_len(n_examiners)),
    examiner_skill = rnorm(n_examiners, mean = 0, sd = dgp_defaults$examiner_skill_sd),
    examiner_inconclusive_tendency = rnorm(n_examiners, mean = 0, sd = dgp_defaults$examiner_inconclusive_sd)
  )

  tidyr::crossing(examiner_panel, comparison_set) %>%
    mutate(decision_challenge = question_difficulty - examiner_skill) %>%
    mutate(response = simulate_responses_vec(
      ground_truth, decision_challenge, examiner_inconclusive_tendency,
      false_positive_rate = false_positive_rate,
      false_negative_rate = false_negative_rate,
      inconclusive_rate = inconclusive_rate
    ))
}

calibration_targets <- c(
  diff_id = diff_identification,
  diff_inc = diff_inconclusive,
  same_elim = same_elimination,
  same_inc = same_inconclusive
)

calibration_objective <- function(par) {
  rates <- plogis(par)
  d <- simulate_large_reference(rates[1], rates[2], rates[3])

  diff <- d %>% filter(ground_truth == 0)
  same <- d %>% filter(ground_truth == 1)

  sim <- c(
    diff_id = mean(diff$response == "identification"),
    diff_inc = mean(diff$response == "inconclusive"),
    same_elim = mean(same$response == "elimination"),
    same_inc = mean(same$response == "inconclusive")
  )

  sum((sim - calibration_targets)^2)
}

calibration_fit <- optim(
  par = qlogis(c(dgp_defaults$false_positive_rate, dgp_defaults$false_negative_rate, dgp_defaults$inconclusive_rate)),
  fn = calibration_objective,
  method = "Nelder-Mead",
  control = list(maxit = 300, reltol = 1e-10)
)

calibrated_rates <- plogis(calibration_fit$par)
names(calibrated_rates) <- c("false_positive_rate", "false_negative_rate", "inconclusive_rate")

knitr::kable(
  tibble(parameter = names(calibrated_rates), fitted_value = round(calibrated_rates, 4)),
  col.names = c("Parameter", "Fitted value")
)
```

Rounding those fitted values to cleaner numbers (`0.01`, `0.02`, `0.36`)
barely changes the fit, so those are the retuned defaults used from here
on.

```{r update-defaults}
dgp_defaults$false_positive_rate <- 0.01
dgp_defaults$false_negative_rate <- 0.02
dgp_defaults$inconclusive_rate <- 0.36

incomplete_defaults$false_positive_rate <- 0.01
incomplete_defaults$false_negative_rate <- 0.02
incomplete_defaults$inconclusive_rate <- 0.36

knitr::kable(
  tibble(parameter = names(dgp_defaults), value = unlist(dgp_defaults)),
  col.names = c("Parameter", "Retuned default value")
)
```

## Verifying the retuned defaults under the incomplete design

Re-running the same 100-replicate, incomplete-design simulation as before
(50 examiners, 30 of 300 items each), now with the retuned rates.

```{r simulate-100-studies-retuned, cache=TRUE}
set.seed(2026)
replicate_seeds_retuned <- sample.int(1e6, n_replications)

sim_studies_retuned <- purrr::map_dfr(seq_len(n_replications), function(r) {
  study_data <- generate_sim_data_incomplete(
    seed = replicate_seeds_retuned[r],
    n_examiners = incomplete_defaults$n_examiners,
    n_comparisons = incomplete_defaults$n_comparisons,
    items_per_examiner = incomplete_defaults$items_per_examiner,
    match_rate = incomplete_defaults$match_rate,
    false_positive_rate = incomplete_defaults$false_positive_rate,
    false_negative_rate = incomplete_defaults$false_negative_rate,
    inconclusive_rate = incomplete_defaults$inconclusive_rate,
    examiner_skill_sd = incomplete_defaults$examiner_skill_sd,
    examiner_inconclusive_sd = incomplete_defaults$examiner_inconclusive_sd,
    question_sd = incomplete_defaults$question_sd
  )

  study_crosstab(study_data, replicate_id = r)
})

pooled_crosstab_retuned <- sim_studies_retuned %>%
  group_by(ground_truth_label, response) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  group_by(ground_truth_label) %>%
  mutate(pooled_proportion = total_n / sum(total_n)) %>%
  ungroup() %>%
  select(-total_n)
```

```{r plot-comparison-retuned, fig.width=9, fig.height=5}
sim_studies_retuned_plot <- sim_studies_retuned %>%
  mutate(
    ground_truth_label = factor(ground_truth_label, levels = c("same-source", "different-source")),
    response = factor(response, levels = c("identification", "inconclusive", "elimination"))
  )

ggplot(sim_studies_retuned_plot, aes(x = response, y = proportion)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_point(
    data = ames_benchmark_plot,
    aes(y = benchmark_proportion),
    color = "#D55E00",
    shape = 18,
    size = 4
  ) +
  facet_wrap(~ ground_truth_label) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Simulated proportions across 100 studies, retuned defaults",
    subtitle = "Boxplots: 100 simulated studies, incomplete design, retuned rates. Diamond: Ames-II-family benchmark",
    x = "Decision",
    y = "Proportion within ground-truth category"
  ) +
  theme_minimal()
```

## Before vs. after retuning

```{r before-after-retuning}
before_after <- comparison_table_incomplete %>%
  select(ground_truth_label, response, before = pooled_proportion) %>%
  left_join(
    pooled_crosstab_retuned %>% select(ground_truth_label, response, after = pooled_proportion),
    by = c("ground_truth_label", "response")
  ) %>%
  left_join(ames_benchmark, by = c("ground_truth_label", "response")) %>%
  mutate(
    ground_truth_label = factor(ground_truth_label, levels = c("same-source", "different-source")),
    response = factor(response, levels = c("identification", "inconclusive", "elimination"))
  ) %>%
  arrange(ground_truth_label, response)

knitr::kable(
  before_after %>%
    mutate(across(c(before, after, benchmark_proportion), ~ percent(.x, accuracy = 0.1))),
  col.names = c("Ground truth", "Decision", "Before retuning", "After retuning", "Benchmark")
)
```

## Sanity check: does this match `R/simulate_dgp.R`?

```{r check-canonical-defaults}
rate_names <- c("false_positive_rate", "false_negative_rate", "inconclusive_rate")

knitr::kable(
  tibble(
    parameter = rate_names,
    this_document = unlist(dgp_defaults[rate_names]),
    shared_file = unlist(canonical_defaults[rate_names])
  ),
  col.names = c("Parameter", "Retuned here", "R/simulate_dgp.R")
)
```

This document's derived values match the shared file's canonical defaults
— `R/simulate_dgp.R` already reflects the outcome of this calibration, so
nothing further needs to be copied over by hand.

## Interpretation of the retuned fit

FPR and FNR are now close to the benchmark: false positive rate
~1.7% vs. a 0.9% benchmark, false negative rate ~3.4% vs. 1.8% — both much
closer than the roughly 3.5x/4.3x gaps before retuning, though still
running a bit high in the same direction as before.

The inconclusive-rate gap is smaller than before retuning but, as expected
from the structural argument above, doesn't close — it's now split roughly
evenly in both directions: simulated inconclusive rates land around
37% for both ground-truth categories, against a benchmark of 50.5%
(different-source, still ~13 pp low) and 23.8% (same-source, now ~13 pp
*high*). That residual asymmetry is a known, accepted limitation of these
defaults, not an oversight — closing it would require giving same-source
and different-source pairs separate inconclusive baselines, which was
deliberately deferred to keep the model matching what's already on the
slides.

# Next steps

- ~~`icfis/app.R`'s slider defaults should be updated to match these retuned
  rates~~ — done: `R/simulate_dgp.R` now holds the canonical, already-retuned
  defaults, and both `icfis/app.R` and this document source it.
- Track down the actual Monson, Smith, and Bajic (2023a) crosstab (or
  another Ames-II-family table) to replace the reconstructed benchmark with
  real reported proportions, including sample sizes and the actual
  examiner/item-pool counts (we used 50 examiners / 30 of 300 items as a
  stand-in for "roughly 30 questions from a pool of roughly 300" — worth
  replacing with the real numbers if available).
- Re-run this comparison against at least one more study (e.g., AMES I /
  Baldwin et al.) to check whether "true" inconclusive/error rates are
  reasonably stable across studies or vary enough that a single target
  isn't meaningful.
- If the residual same-source/different-source inconclusive-rate gap ends
  up mattering for the talk (e.g., it visibly weakens the inconclusive-
  handling material), revisit the decision to keep a single shared
  `inconclusive_rate` and consider the model extension noted above.
- Decide whether `icfis/app.R` itself should offer the incomplete-design
  option (it currently only supports full crossing), since that's the more
  realistic design and is now implemented here.

# References

::: {#refs}
:::
