---
title: "Case Study: A Non-Remediable Flaw (Non-Representative Sampling)"
subtitle: "Flaw B -- why reweighting can't fix a sample with no data on most of the population"
bibliography: ../output/paper/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
    code-fold: true
---

# Goal

This is case study 2 for the talk, matching the three-beat structure
already on the "Simulation design" slide in `icfis/draft.qmd`:

1. A convenience sample of firearm types produces a biased estimate of the
   population-average error rate.
2. *If* a study had covered every type (even non-proportionally) and
   recorded type as a covariate, reweighting by the true population
   distribution would recover something close to the true value — this is
   Chin et al.'s "moderator" / big-team-science suggestion, and it works.
3. The actual convenience sample has **zero** observations for most of the
   population's weight. Applying that same reweighting idea to it doesn't
   work — not because nobody tried, but because there's no data to weight.

Non-representative sampling in real black-box studies operates at two
nested levels, both documented in @cuellar2024methodological: which
firearm **makes/models** get tested at all (between-model), and, even for
a model that is tested, which specific **individual firearms** of that
model happen to be in the lab (within-model). This document models both,
stacked, since a real convenience sample is short on both dimensions at
once.

This builds on the same shared DGP and canonical (retuned) defaults as
`icfis/analysis.qmd` and `icfis/case_study_inconclusive.qmd`.

```{r setup}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)

source("R/simulate_dgp.R")
source("R/simulate_dgp_types.R")
```

# A population of firearm types (between-model)

"Type" here means firearm **make/model** (e.g., Beretta M9A3 vs. Glock 19)
— not an individual firearm. Eight types, each with its own baseline
classification difficulty. @cuellar2024methodological (p. 10) cite
evidence from Guyll et al. (2023) and Bajic et al. (2020, "AMES II") that
different makes/models (their example: Jimenez vs. Beretta) differ in how
easy they are to match — that's the between-model heterogeneity modeled
here.

```{r type-population}
type_population <- make_type_population()

knitr::kable(
  type_population %>%
    arrange(type_id) %>%
    mutate(
      population_weight = percent(population_weight, accuracy = 1),
      type_difficulty = round(type_difficulty, 2)
    ),
  col.names = c("Type", "Population weight (casework)", "Difficulty (latent scale)")
)

easiest_types <- type_population %>%
  arrange(type_difficulty) %>%
  slice(1:2) %>%
  pull(type_id)
```

The two easiest types to classify are types `r paste(easiest_types, collapse = " and ")`,
together accounting for
`r percent(sum(type_population$population_weight[type_population$type_id %in% easiest_types]), accuracy = 1)`
of real casework. The convenience sample below is restricted to exactly
these two — a stand-in for "the lab used whatever well-maintained,
easy-to-compare firearms it had on hand," echoing the paper's discussion of
newly-manufactured vs. damaged/corroded firearms affecting difficulty.

# Individual firearms within each type (within-model)

Even firearms of the *same* model differ. @cuellar2024methodological cite
a 0%–62% "hard" range across 27 otherwise-identical Beretta pistols in
Ames II — a big enough range to take seriously as a second, additive
source of difficulty on top of the type-level effect above. Each type's
"world population" here has 50 individual firearms, each with its own
difficulty offset (independent of its type's difficulty and of how common
that type is).

```{r firearm-population}
firearm_population <- make_firearm_population(type_population)

firearm_population_summary <- firearm_population %>%
  group_by(type_id) %>%
  summarise(
    n_firearms = n(),
    mean_firearm_difficulty = mean(firearm_difficulty),
    sd_firearm_difficulty = sd(firearm_difficulty),
    .groups = "drop"
  )

knitr::kable(
  firearm_population_summary %>% mutate(across(c(mean_firearm_difficulty, sd_firearm_difficulty), ~ round(.x, 2))),
  col.names = c("Type", "Firearms in world population", "Mean firearm difficulty", "SD firearm difficulty")
)
```

A real study only ever uses a handful of specific firearms per type it
includes — Ames-II-family studies used on the order of 27–37 firearms
total, not the full population of each model in circulation. Every
scenario below draws only 5 specific firearms per included type, so even
the "full-coverage" study in beat 2 only samples a small, possibly
unrepresentative slice of each type's own difficulty range.

# The true population-average error rate

A large reference simulation: items are assigned to types proportional to
the *true* population weights, and within each item's type, to an
individual firearm drawn from that type's full 50-firearm population
(with replacement, since the reference is meant to approximate the full
diversity of firearms in casework, not one study's limited draw). This is
the estimand the talk cares about: the population-average error rate
under casework-representative firearms.

```{r true-reference}
true_reference_data <- generate_population_reference_by_firearm(
  seed = 1,
  type_population = type_population,
  firearm_population = firearm_population,
  sampling_weights = type_population$population_weight,
  n_items_total = 2000,
  n_examiners = dgp_defaults$n_examiners,
  match_rate = dgp_defaults$match_rate,
  false_positive_rate = dgp_defaults$false_positive_rate,
  false_negative_rate = dgp_defaults$false_negative_rate,
  inconclusive_rate = dgp_defaults$inconclusive_rate,
  examiner_skill_sd = dgp_defaults$examiner_skill_sd,
  examiner_inconclusive_sd = dgp_defaults$examiner_inconclusive_sd,
  question_sd = dgp_defaults$question_sd
)

true_reference_rates <- compute_pooled_rates(true_reference_data)

knitr::kable(
  true_reference_rates %>% mutate(rate = percent(rate, accuracy = 0.01)),
  col.names = c("Metric", "True population-average rate")
)
```

# The three scenarios

Every scenario now draws 5 specific firearms per included type (held
fixed across scenarios, so the comparison isolates *which types* get
included, not how many firearms per type) and generates several
comparisons per firearm.

```{r define-scenarios}
# Beat 1: convenience sample -- only the two easiest types.
conv_weights <- if_else(type_population$type_id %in% easiest_types, 1, 0)

# Beat 2: hypothetical better study -- every type represented (not
# necessarily in population-correct proportions; equal allocation is a
# common, defensible design choice as long as type is recorded and the
# analyst reweights afterward).
better_weights <- rep(1, nrow(type_population))

n_firearms_per_type <- 5
n_replications <- 500
set.seed(4042)
replicate_seeds <- sample.int(1e6, n_replications)

simulate_scenario <- function(seed, sampling_weights, items_per_firearm) {
  generate_sim_data_by_type_and_firearm(
    seed = seed,
    type_population = type_population,
    firearm_population = firearm_population,
    sampling_weights = sampling_weights,
    n_firearms_per_type = n_firearms_per_type,
    items_per_firearm = items_per_firearm,
    n_examiners = dgp_defaults$n_examiners,
    match_rate = dgp_defaults$match_rate,
    false_positive_rate = dgp_defaults$false_positive_rate,
    false_negative_rate = dgp_defaults$false_negative_rate,
    inconclusive_rate = dgp_defaults$inconclusive_rate,
    examiner_skill_sd = dgp_defaults$examiner_skill_sd,
    examiner_inconclusive_sd = dgp_defaults$examiner_inconclusive_sd,
    question_sd = dgp_defaults$question_sd
  )
}
```

## Beat 1: the convenience sample is biased

`r n_replications` replicate studies, each using 5 firearms from each of
the two easiest types (10 firearms, 10 items per firearm = 100 items
total — same overall scale as before the firearm layer was added).

```{r beat-1, cache=TRUE, cache.extra=n_replications}
convenience_results <- purrr::map_dfr(seq_len(n_replications), function(r) {
  study_data <- simulate_scenario(replicate_seeds[r], conv_weights, items_per_firearm = 10)
  compute_pooled_rates(study_data) %>%
    mutate(replicate_id = r, scenario = "Convenience sample (naive)")
})
```

## Beat 2: the fix that could work

`r n_replications` replicate studies, each using 5 firearms from every one
of the eight types (40 firearms, 4 items per firearm = 160 items total) —
then reweighted by the *true* population distribution.

```{r beat-2, cache=TRUE, cache.extra=n_replications}
better_study_results <- purrr::map_dfr(seq_len(n_replications), function(r) {
  study_data <- simulate_scenario(replicate_seeds[r], better_weights, items_per_firearm = 4)
  type_rates <- compute_type_level_rates(study_data)
  reweight_by_population(type_rates, type_population) %>%
    rename(rate = reweighted_rate) %>%
    mutate(replicate_id = r, scenario = "Full-coverage study (reweighted)")
})
```

## Beat 3: the fix that doesn't exist

The *same* reweighting procedure applied to the convenience sample from
beat 1 — renormalizing the population weights over just the two types
actually observed.

```{r beat-3, cache=TRUE, cache.extra=n_replications}
convenience_fixed_results <- purrr::map_dfr(seq_len(n_replications), function(r) {
  study_data <- simulate_scenario(replicate_seeds[r], conv_weights, items_per_firearm = 10)
  type_rates <- compute_type_level_rates(study_data)
  reweight_by_population(type_rates, type_population) %>%
    rename(rate = reweighted_rate) %>%
    mutate(replicate_id = r, scenario = "Convenience sample (\"reweighted\")")
})
```

# Results

```{r combine-results}
all_results <- bind_rows(convenience_results, better_study_results, convenience_fixed_results) %>%
  mutate(
    scenario = factor(
      scenario,
      levels = c(
        "Convenience sample (naive)",
        "Full-coverage study (reweighted)",
        "Convenience sample (\"reweighted\")"
      )
    ),
    metric = factor(metric, levels = c("FPR", "FNR"))
  )

summary_table <- all_results %>%
  group_by(metric, scenario) %>%
  summarise(mean_rate = mean(rate), sd_rate = sd(rate), .groups = "drop") %>%
  mutate(se_rate = sd_rate / sqrt(n_replications)) %>%
  left_join(
    true_reference_rates %>% rename(true_rate = rate) %>% mutate(metric = factor(metric, levels = c("FPR", "FNR"))),
    by = "metric"
  ) %>%
  arrange(metric, scenario)

knitr::kable(
  summary_table %>%
    mutate(
      mean_rate = percent(mean_rate, accuracy = 0.01),
      sd_rate = percent(sd_rate, accuracy = 0.01),
      se_rate = percent(se_rate, accuracy = 0.001),
      true_rate = percent(true_rate, accuracy = 0.01)
    ),
  col.names = c("Metric", "Scenario", sprintf("Mean (%d studies)", n_replications), "SD", "SE of the mean", "True population rate")
)
```

```{r plot-results, fig.width=9, fig.height=5}
true_reference_plot <- true_reference_rates %>%
  mutate(metric = factor(metric, levels = c("FPR", "FNR")))

ggplot(all_results, aes(x = scenario, y = rate)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_hline(
    data = true_reference_plot,
    aes(yintercept = rate),
    color = "#D55E00",
    linetype = "dashed",
    linewidth = 1
  ) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Naive vs. reweighted estimates against the true population rate",
    subtitle = "Dashed line: true population-average rate. Reweighting only helps when the sample covers the population.",
    x = NULL,
    y = "Rate"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

# Monte Carlo error

Each scenario's reported mean is itself an estimate from a finite number
of simulated studies, so it carries its own uncertainty — separate from
the substantive bias/variance story the case study is about. The
`SE of the mean` column above is $SD/\sqrt{`r n_replications`}$ for each
scenario/metric; a rough 95% CI for the *mean* estimate itself is
`r n_replications`-replicate mean $\pm\ 1.96 \times$ that SE.

```{r mc-error-table}
knitr::kable(
  summary_table %>%
    mutate(
      ci_low = mean_rate - 1.96 * se_rate,
      ci_high = mean_rate + 1.96 * se_rate
    ) %>%
    mutate(across(c(mean_rate, se_rate, ci_low, ci_high), ~ percent(.x, accuracy = 0.01))) %>%
    select(metric, scenario, mean_rate, se_rate, ci_low, ci_high),
  col.names = c("Metric", "Scenario", "Mean", "SE of the mean", "95% CI (low)", "95% CI (high)")
)
```

These CIs are narrow relative to the gaps the case study rests on (the
full-coverage vs. convenience-sample means differ by roughly 1.8
percentage points on FPR; the widest CI half-width above is under 0.1
percentage points), so Monte Carlo noise isn't a plausible explanation for
any of the three-beat comparisons.

## Was 500 replicates actually necessary?

Comparing the `r n_replications`-replicate means reported here against the
100-replicate means from the version of this document before the replicate
count was raised:

```{r mc-error-comparison}
previous_100_results <- tribble(
  ~metric, ~scenario, ~mean_rate_100,
  "FPR", "Convenience sample (naive)", 0.0041,
  "FPR", "Full-coverage study (reweighted)", 0.0232,
  "FPR", "Convenience sample (\"reweighted\")", 0.0046,
  "FNR", "Convenience sample (naive)", 0.0077,
  "FNR", "Full-coverage study (reweighted)", 0.0393,
  "FNR", "Convenience sample (\"reweighted\")", 0.0085
) %>%
  mutate(
    metric = factor(metric, levels = c("FPR", "FNR")),
    scenario = factor(
      scenario,
      levels = c(
        "Convenience sample (naive)",
        "Full-coverage study (reweighted)",
        "Convenience sample (\"reweighted\")"
      )
    )
  )

mc_comparison <- summary_table %>%
  select(metric, scenario, mean_rate_500 = mean_rate, se_rate) %>%
  left_join(previous_100_results, by = c("metric", "scenario")) %>%
  mutate(shift = mean_rate_500 - mean_rate_100)

knitr::kable(
  mc_comparison %>%
    mutate(across(c(mean_rate_100, mean_rate_500, shift), ~ percent(.x, accuracy = 0.01))) %>%
    select(metric, scenario, mean_rate_100, mean_rate_500, shift),
  col.names = c("Metric", "Scenario", "Mean at n=100", "Mean at n=500", "Shift")
)
```

The means moved by at most a few hundredths of a percentage point between
100 and 500 replicates — well inside the SE at n=100 (which was
$\sqrt{5} \approx 2.24\times$ larger than the SE at n=500 shown above, for
the same underlying SD). In other words, **100 replicates already gave a
stable answer** for every conclusion this case study draws; going to 500
tightens the estimate's own uncertainty but doesn't change any of the
three-beat comparisons. Worth knowing before spending the extra ~2.5
minutes of render time on every future edit — 100 may be the better
default while still iterating on the narrative, with 500 reserved for a
final check before the talk.

# Interpretation

Using the pooled/mean rates from the `r n_replications`-replication table
above (true population rate: FPR 2.13%, FNR 3.67%):

- **The convenience sample is badly biased, in the optimistic direction.**
  Restricting the study to the two easiest types (and only 5 of each
  type's 50 firearms) produces FPR 0.40% and FNR 0.77% — roughly **5x too
  low** on both metrics.
- **Reweighting works when the sample has full coverage.** The
  full-coverage study (all eight types, 5 firearms each) reaches FPR
  2.27% and FNR 3.98% once reweighted — within about 6–8% of the true rate
  on both metrics, even though it too only samples 5 of each type's 50
  firearms. Within-model firearm sampling adds noise (see the SD column),
  not bias, as long as every type is represented.
- **The same correction fails on the actual convenience sample.**
  Reweighting the two-type convenience sample by population share moves
  the estimate only to FPR 0.44% and FNR 0.86% — still roughly **4.5x too
  low** on both metrics, barely different from doing nothing. There's no
  information at all about the six unsampled types, which together hold
  79% of the population's weight. Reweighting can't fix that, no matter
  how many (or how few) firearms of the *included* types were tested.
- **Spread is visibly wider than before the firearm layer was added.**
  Compare SDs to the type-only version of this case study: convenience
  sample FPR SD is 0.22% here vs. 0.13% there; full-coverage reweighted
  FPR SD is 0.71% here vs. 0.40% there. Which specific 5 firearms got
  drawn for a given type is now an extra source of study-to-study
  variability, on top of the between-type sampling question beats 1–3 are
  about — visible in the boxplots' width, not just their position. (As the
  Monte Carlo section above shows, none of this spread is an artifact of
  too few replications — it's the actual between-study variability the
  design produces.)

# Why this counts as non-remediable

Nothing about the *analysis* is broken here — the reweighting procedure in
beat 2 is statistically valid and works, even with the added within-model
noise from only sampling 5 firearms per type. What fails in beat 3 is
specific to the convenience sample's *design*: it never observed most of
the population it claims to generalize to, at the type level. No amount
of reanalysis of that particular dataset can fix a positivity violation
this severe — and adding a more realistic within-model layer doesn't
change that conclusion, it just makes the simulation's design closer to
how an actual black-box study is built (a handful of types, a handful of
firearms per type, many comparisons per firearm).

# Next steps

- Swap the updated numbers into `icfis/draft.qmd`'s Case Study 2 slides.
- Consider whether the "hypothetical better study" (beat 2) should use
  equal-per-type allocation or a different defensible design (e.g.,
  oversampling rare types further for power) — equal allocation was
  chosen for simplicity, not because it's the only valid choice.
- `firearm_difficulty_sd = 0.9` (in `make_firearm_population()`) is
  illustrative, not calibrated to the 0%–62% Beretta figure or anything
  else specific — worth a sensitivity check once there's time.
- If useful, add a version of beat 3 using *partial* type coverage (e.g.,
  4 of 8 types, not just the 2 easiest) to show the bias shrinking
  gradually as coverage improves, rather than only the two extremes.

# References

::: {#refs}
:::
