---
title: "Case Study: A Non-Remediable Flaw (Non-Representative Sampling)"
subtitle: "Flaw B -- why reweighting can't fix a sample with no data on most of the population"
bibliography: ../output/paper/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
    code-fold: true
---

# Goal

This is case study 2 for the talk, matching the three-beat structure
already on the "Simulation design" slide in `icfis/draft.qmd`:

1. A convenience sample of firearm types produces a biased estimate of the
   population-average error rate.
2. *If* a study had covered every type (even non-proportionally) and
   recorded type as a covariate, reweighting by the true population
   distribution would recover something close to the true value — this is
   Chin et al.'s "moderator" / big-team-science suggestion, and it works.
3. The actual convenience sample has **zero** observations for most of the
   population's weight. Applying that same reweighting idea to it doesn't
   work — not because nobody tried, but because there's no data to weight.

Non-representative sampling in real black-box studies operates at two
nested levels, both documented in @cuellar2024methodological: which
firearm **makes/models** get tested at all (between-model), and, even for
a model that is tested, which specific **individual firearms** of that
model happen to be in the lab (within-model). This document models both,
stacked, since a real convenience sample is short on both dimensions at
once.

This builds on the same shared DGP and canonical (retuned) defaults as
`icfis/analysis.qmd` and `icfis/case_study_inconclusive.qmd`.

```{r setup}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)

source("R/simulate_dgp.R")
source("R/simulate_dgp_types.R")
```

# A population of firearm types (between-model)

"Type" here means firearm **make/model** (e.g., Beretta M9A3 vs. Glock 19)
— not an individual firearm. Eight types, each with its own baseline
classification difficulty. @cuellar2024methodological (p. 10) cite
evidence from Guyll et al. (2023) and Bajic et al. (2020, "AMES II") that
different makes/models (their example: Jimenez vs. Beretta) differ in how
easy they are to match — that's the between-model heterogeneity modeled
here.

```{r type-population}
type_population <- make_type_population()

knitr::kable(
  type_population %>%
    arrange(type_id) %>%
    mutate(
      population_weight = percent(population_weight, accuracy = 1),
      type_difficulty = round(type_difficulty, 2)
    ),
  col.names = c("Type", "Population weight (casework)", "Difficulty (latent scale)")
)

easiest_types <- type_population %>%
  arrange(type_difficulty) %>%
  slice(1:2) %>%
  pull(type_id)

# Design constants for the three-beat comparison below, defined here (ahead of
# their first use in prose) so later sections can reference them inline instead
# of repeating literals.
world_firearms_per_type <- 4000
total_firearms <- 80
n_firearms_convenience <- total_firearms / length(easiest_types)
n_firearms_full_coverage <- total_firearms / nrow(type_population)
items_per_examiner <- 30
```

The two easiest types to classify are types `r paste(easiest_types, collapse = " and ")`,
together accounting for
`r percent(sum(type_population$population_weight[type_population$type_id %in% easiest_types]), accuracy = 1)`
of real casework. The convenience sample below is restricted to exactly
these two — a stand-in for "the lab used whatever well-maintained,
easy-to-compare firearms it had on hand," echoing the paper's discussion of
newly-manufactured vs. damaged/corroded firearms affecting difficulty.

# Individual firearms within each type (within-model)

Even firearms of the *same* model differ. @cuellar2024methodological cite
a 0%–62% "hard" range across 27 otherwise-identical Beretta pistols in
Ames II — a big enough range to take seriously as a second, additive
source of difficulty on top of the type-level effect above. Each type's
"world population" here has `r comma(world_firearms_per_type)` individual
firearms, each with its own difficulty offset (independent of its type's
difficulty and of how common that type is) — large enough that the
`r n_firearms_convenience` and `r n_firearms_full_coverage` firearms drawn
per type below (`r percent(n_firearms_convenience / world_firearms_per_type, accuracy = 0.1)`
and `r percent(n_firearms_full_coverage / world_firearms_per_type, accuracy = 0.1)`
of the world population, respectively) are genuinely small samples, unlike
an earlier version of this document whose 50-firearm world population meant
the convenience sample's 40 firearms/type covered 80% of it.

```{r firearm-population}
firearm_population <- make_firearm_population(type_population, firearms_per_type = world_firearms_per_type)

firearm_population_summary <- firearm_population %>%
  group_by(type_id) %>%
  summarise(
    n_firearms = n(),
    mean_firearm_difficulty = mean(firearm_difficulty),
    sd_firearm_difficulty = sd(firearm_difficulty),
    .groups = "drop"
  )

knitr::kable(
  firearm_population_summary %>% mutate(across(c(mean_firearm_difficulty, sd_firearm_difficulty), ~ round(.x, 2))),
  col.names = c("Type", "Firearms in world population", "Mean firearm difficulty", "SD firearm difficulty")
)
```

A real study only ever uses a handful of specific firearms per type it
includes — Ames-II-family studies used on the order of 27–37 firearms
total, not the full population of each model in circulation. Every
scenario below draws from the same total pool of 80 firearms, so the
comparison isolates *which types* get included, not how large the study
is overall: the convenience sample spreads those 80 firearms across only
2 types (40 each), while the full-coverage study spreads them across all
8 types (10 each).

# The true population-average error rate

A large reference simulation: items are assigned to types proportional to
the *true* population weights, and within each item's type, to an
individual firearm drawn from that type's full `r comma(world_firearms_per_type)`-firearm population
(with replacement, since the reference is meant to approximate the full
diversity of firearms in casework, not one study's limited draw). This is
the estimand the talk cares about: the population-average error rate
under casework-representative firearms.

```{r true-reference}
true_reference_data <- generate_population_reference_by_firearm(
  seed = 1,
  type_population = type_population,
  firearm_population = firearm_population,
  sampling_weights = type_population$population_weight,
  n_items_total = 2000,
  n_examiners = dgp_defaults$n_examiners,
  match_rate = dgp_defaults$match_rate,
  false_positive_rate = dgp_defaults$false_positive_rate,
  false_negative_rate = dgp_defaults$false_negative_rate,
  inconclusive_rate = dgp_defaults$inconclusive_rate,
  inconclusive_rate_same = dgp_defaults$inconclusive_rate_same,
  examiner_skill_sd = dgp_defaults$examiner_skill_sd,
  examiner_inconclusive_sd = dgp_defaults$examiner_inconclusive_sd,
  question_sd = dgp_defaults$question_sd
)

true_reference_rates <- compute_pooled_rates(true_reference_data)

knitr::kable(
  true_reference_rates %>% mutate(rate = percent(rate, accuracy = 0.01)),
  col.names = c("Metric", "True population-average rate")
)
```

# The three scenarios

**Total firearm count held fixed at 80 across scenarios, instead of
firearms-per-type held fixed at 5** (an earlier version of this document
used 5 firearms/type regardless of how many types were included, so the
convenience sample used only 10 firearms total against the full-coverage
study's 40 — this version checks whether the story changes once total
study size is put on equal footing). The convenience sample spreads 80
firearms across its 2 types (40 each); the full-coverage study spreads
the same 80 across all 8 types (10 each).

**Incomplete design, matching case study 1.** Each examiner sees only
`r items_per_examiner` of the scenario's items, assigned independently at
random without replacement — the same mechanism
`case_study_inconclusive.qmd` uses (`generate_sim_data_incomplete()`), now
extended to the type/firearm structure
(`generate_sim_data_by_type_and_firearm_incomplete()` in
`R/simulate_dgp_types.R`). `r items_per_examiner` is held fixed across all
three beats, so only the firearm-sampling design differs between scenarios,
not each examiner's workload. An earlier version of this document had every
examiner evaluate every item (fully crossed); this checks whether the story
changes once examiners see only a fraction of the pool, as they do in a real
study.

```{r define-scenarios}
# Beat 1: convenience sample -- only the two easiest types.
conv_weights <- if_else(type_population$type_id %in% easiest_types, 1, 0)

# Beat 2: hypothetical better study -- every type represented (not
# necessarily in population-correct proportions; equal allocation is a
# common, defensible design choice as long as type is recorded and the
# analyst reweights afterward).
better_weights <- rep(1, nrow(type_population))

# total_firearms, n_firearms_convenience, n_firearms_full_coverage,
# world_firearms_per_type, and items_per_examiner are defined earlier, in the
# type-population chunk.

n_replications <- 500
set.seed(4042)
replicate_seeds <- sample.int(1e6, n_replications)

simulate_scenario <- function(seed, sampling_weights, n_firearms_per_type, items_per_firearm) {
  generate_sim_data_by_type_and_firearm_incomplete(
    seed = seed,
    type_population = type_population,
    firearm_population = firearm_population,
    sampling_weights = sampling_weights,
    n_firearms_per_type = n_firearms_per_type,
    items_per_firearm = items_per_firearm,
    items_per_examiner = items_per_examiner,
    n_examiners = dgp_defaults$n_examiners,
    match_rate = dgp_defaults$match_rate,
    false_positive_rate = dgp_defaults$false_positive_rate,
    false_negative_rate = dgp_defaults$false_negative_rate,
    inconclusive_rate = dgp_defaults$inconclusive_rate,
    inconclusive_rate_same = dgp_defaults$inconclusive_rate_same,
    examiner_skill_sd = dgp_defaults$examiner_skill_sd,
    examiner_inconclusive_sd = dgp_defaults$examiner_inconclusive_sd,
    question_sd = dgp_defaults$question_sd
  )
}
```

## Beat 1: the convenience sample is biased

`r n_replications` replicate studies, each using
`r n_firearms_convenience` firearms from each of the two easiest types
(`r total_firearms` firearms total, 10 items per firearm =
`r total_firearms * 10` items in the pool, `r items_per_examiner` seen per
examiner).

```{r beat-1, cache=TRUE, cache.extra=n_replications}
convenience_results <- purrr::map_dfr(seq_len(n_replications), function(r) {
  study_data <- simulate_scenario(replicate_seeds[r], conv_weights, n_firearms_convenience, items_per_firearm = 10)
  compute_pooled_rates(study_data) %>%
    mutate(replicate_id = r, scenario = "Convenience sample (naive)")
})
```

## Beat 2: the fix that could work

`r n_replications` replicate studies, each using
`r n_firearms_full_coverage` firearms from every one of the eight types
(`r total_firearms` firearms total, 4 items per firearm =
`r total_firearms * 4` items in the pool, `r items_per_examiner` seen per
examiner) — then reweighted by the *true* population distribution.

```{r beat-2, cache=TRUE, cache.extra=n_replications}
better_study_results <- purrr::map_dfr(seq_len(n_replications), function(r) {
  study_data <- simulate_scenario(replicate_seeds[r], better_weights, n_firearms_full_coverage, items_per_firearm = 4)
  type_rates <- compute_type_level_rates(study_data)
  reweight_by_population(type_rates, type_population) %>%
    rename(rate = reweighted_rate) %>%
    mutate(replicate_id = r, scenario = "Full-coverage study (reweighted)")
})
```

## Beat 3: the fix that doesn't exist

The *same* reweighting procedure applied to the convenience sample from
beat 1 — renormalizing the population weights over just the two types
actually observed.

```{r beat-3, cache=TRUE, cache.extra=n_replications}
convenience_fixed_results <- purrr::map_dfr(seq_len(n_replications), function(r) {
  study_data <- simulate_scenario(replicate_seeds[r], conv_weights, n_firearms_convenience, items_per_firearm = 10)
  type_rates <- compute_type_level_rates(study_data)
  reweight_by_population(type_rates, type_population) %>%
    rename(rate = reweighted_rate) %>%
    mutate(replicate_id = r, scenario = "Convenience sample (\"reweighted\")")
})
```

# Results

```{r combine-results}
all_results <- bind_rows(convenience_results, better_study_results, convenience_fixed_results) %>%
  mutate(
    scenario = factor(
      scenario,
      levels = c(
        "Convenience sample (naive)",
        "Full-coverage study (reweighted)",
        "Convenience sample (\"reweighted\")"
      )
    ),
    metric = factor(metric, levels = c("FPR", "FNR"))
  )

summary_table <- all_results %>%
  group_by(metric, scenario) %>%
  summarise(mean_rate = mean(rate), sd_rate = sd(rate), .groups = "drop") %>%
  mutate(se_rate = sd_rate / sqrt(n_replications)) %>%
  left_join(
    true_reference_rates %>% rename(true_rate = rate) %>% mutate(metric = factor(metric, levels = c("FPR", "FNR"))),
    by = "metric"
  ) %>%
  arrange(metric, scenario)

knitr::kable(
  summary_table %>%
    mutate(
      mean_rate = percent(mean_rate, accuracy = 0.01),
      sd_rate = percent(sd_rate, accuracy = 0.01),
      se_rate = percent(se_rate, accuracy = 0.001),
      true_rate = percent(true_rate, accuracy = 0.01)
    ),
  col.names = c("Metric", "Scenario", sprintf("Mean (%d studies)", n_replications), "SD", "SE of the mean", "True population rate")
)
```

```{r plot-results, fig.width=9, fig.height=5}
true_reference_plot <- true_reference_rates %>%
  mutate(metric = factor(metric, levels = c("FPR", "FNR")))

ggplot(all_results, aes(x = scenario, y = rate)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_hline(
    data = true_reference_plot,
    aes(yintercept = rate),
    color = "#D55E00",
    linetype = "dashed",
    linewidth = 1
  ) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Naive vs. reweighted estimates against the true population rate",
    subtitle = "Dashed line: true population-average rate. Reweighting only helps when the sample covers the population.",
    x = NULL,
    y = "Rate"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

# Monte Carlo error

Each scenario's reported mean is itself an estimate from a finite number
of simulated studies, so it carries its own uncertainty — separate from
the substantive bias/variance story the case study is about. The
`SE of the mean` column above is $SD/\sqrt{`r n_replications`}$ for each
scenario/metric; a rough 95% CI for the *mean* estimate itself is
`r n_replications`-replicate mean $\pm\ 1.96 \times$ that SE.

```{r mc-error-table}
knitr::kable(
  summary_table %>%
    mutate(
      ci_low = mean_rate - 1.96 * se_rate,
      ci_high = mean_rate + 1.96 * se_rate
    ) %>%
    mutate(across(c(mean_rate, se_rate, ci_low, ci_high), ~ percent(.x, accuracy = 0.01))) %>%
    select(metric, scenario, mean_rate, se_rate, ci_low, ci_high),
  col.names = c("Metric", "Scenario", "Mean", "SE of the mean", "95% CI (low)", "95% CI (high)")
)
```

```{r summary-helpers, include=FALSE}
conv <- "Convenience sample (naive)"
full <- "Full-coverage study (reweighted)"
conv_fix <- "Convenience sample (\"reweighted\")"

pick <- function(m, sc) summary_table %>% filter(metric == m, scenario == sc)
pc <- function(x) percent(x, accuracy = 0.01)
too_low <- function(m, sc) { r <- pick(m, sc); sprintf("%.1fx", r$true_rate / r$mean_rate) }
off_pct <- function(m, sc) { r <- pick(m, sc); sprintf("%.0f%%", 100 * abs(r$mean_rate / r$true_rate - 1)) }
widest_halfwidth <- max(1.96 * summary_table$se_rate)
gap_fpr <- pick("FPR", full)$mean_rate - pick("FPR", conv)$mean_rate
```

These CIs are narrow relative to the gaps the case study rests on (the
full-coverage vs. convenience-sample means differ by roughly
`r sprintf("%.1f", 100 * gap_fpr)` percentage points on FPR; the widest CI half-width
above is `r sprintf("%.2f", 100 * widest_halfwidth)` percentage points), so Monte Carlo noise
isn't a plausible explanation for any of the three-beat comparisons.

## Was 500 replicates actually necessary?

```{r mc-error-comparison, results='asis'}
# This section only has something to compare when n_replications > 100 (i.e.
# a subset of these same replicates can stand in for "what a 100-replicate run
# would have reported"). At n_replications == 100 exactly -- as in this
# reduced check-in run of the incomplete-design switch -- there's no smaller
# subset to compare against, so print a note instead of a degenerate,
# self-vs-self comparison.
if (n_replications > 100) {
  first_100 <- all_results %>%
    filter(replicate_id <= 100) %>%
    group_by(metric, scenario) %>%
    summarise(mean_rate_100 = mean(rate), .groups = "drop")

  mc_comparison <- summary_table %>%
    select(metric, scenario, mean_rate_full = mean_rate, se_rate) %>%
    left_join(first_100, by = c("metric", "scenario")) %>%
    mutate(shift = mean_rate_full - mean_rate_100, se_at_100 = se_rate * sqrt(n_replications / 100))

  cat(sprintf(
    "Comparing the %d-replicate means reported here against the means of just the first 100 of those replicates (what a 100-replicate run with the same seeds would have reported):\n\n",
    n_replications
  ))

  print(knitr::kable(
    mc_comparison %>%
      mutate(across(c(mean_rate_100, mean_rate_full, shift), ~ percent(.x, accuracy = 0.01))) %>%
      select(metric, scenario, mean_rate_100, mean_rate_full, shift),
    col.names = c("Metric", "Scenario", "Mean at n=100", sprintf("Mean at n=%d", n_replications), "Shift")
  ))

  max_shift <- max(abs(mc_comparison$shift))
  max_shift_in_se <- max(abs(mc_comparison$shift) / mc_comparison$se_at_100)
  true_gap_min <- summary_table %>% filter(scenario != full) %>% summarise(g = min(true_rate - mean_rate)) %>% pull(g)

  cat(sprintf(
    "\n\nThe means moved by at most %.2f percentage points between 100 and %d replicates (at most %.1f standard errors of the 100-replicate mean, which is $\\sqrt{%d} \\approx %.2f\\times$ larger than the SE at n=%d shown above) — far smaller than the bias the case study is about (the smallest gap between a convenience-sample mean and the truth is %.2f percentage points). In other words, a 100-replicate run already gives a stable answer for every conclusion this case study draws; going to %d tightens the estimate's own uncertainty but doesn't change any of the three-beat comparisons. Worth knowing before spending the extra render time on every future edit — 100 may be the better default while still iterating on the narrative, with %d reserved for a final check before the talk.\n",
    100 * max_shift, n_replications, max_shift_in_se, n_replications / 100, sqrt(n_replications / 100), n_replications,
    100 * true_gap_min, n_replications, n_replications
  ))
} else {
  cat(sprintf(
    "This run already uses the reduced `n_replications = %d` (see the note in the `define-scenarios` chunk above) — there's no smaller subset of it to compare against. A previous 500-replicate run of this case study (before the incomplete-design switch) found the mean estimates stable to within 0.1 percentage points at n = 100, well inside the bias the case study is about; re-run at 500 replicates for a final Monte Carlo check once the design settles.\n",
    n_replications
  ))
}
```

# Interpretation

Using the mean rates from the `r n_replications`-replication table above (true population
rate: FPR `r pc(pick("FPR", conv)$true_rate)`, FNR `r pc(pick("FNR", conv)$true_rate)`):

- **The convenience sample is badly biased, in the optimistic direction.**
  Restricting the study to the two easiest types (`r n_firearms_convenience`
  of each type's `r comma(world_firearms_per_type)` firearms — `r total_firearms` total, the same total as
  the full-coverage study below) produces FPR `r pc(pick("FPR", conv)$mean_rate)`
  and FNR `r pc(pick("FNR", conv)$mean_rate)` — roughly **`r too_low("FPR", conv)`
  and `r too_low("FNR", conv)` too low**.
- **Reweighting works when the sample has full coverage.** The
  full-coverage study (all eight types, `r n_firearms_full_coverage`
  firearms each — the same `r total_firearms` total, just spread across
  4x as many types) reaches FPR `r pc(pick("FPR", full)$mean_rate)` and FNR
  `r pc(pick("FNR", full)$mean_rate)` once reweighted — within about
  `r off_pct("FPR", full)` and `r off_pct("FNR", full)` of the true rate.
  Within-model firearm sampling adds noise (see the SD column), not bias,
  as long as every type is represented.
- **The same correction fails on the actual convenience sample.**
  Reweighting the two-type convenience sample by population share moves
  the estimate only to FPR `r pc(pick("FPR", conv_fix)$mean_rate)` and FNR
  `r pc(pick("FNR", conv_fix)$mean_rate)` — still roughly **`r too_low("FPR", conv_fix)` and
  `r too_low("FNR", conv_fix)` too low**, barely different from doing nothing. There's no
  information at all about the six unsampled types, which together hold
  79% of the population's weight. Reweighting can't fix that, no matter
  how many (or how few) firearms of the *included* types were tested —
  matching the total firearm count across scenarios doesn't change this:
  a bigger convenience sample of the wrong types is still a sample of the
  wrong types.
- **Spread is wide relative to the rates being estimated.** Full-coverage reweighted
  FPR has an SD of `r pc(pick("FPR", full)$sd_rate)` around a mean of `r pc(pick("FPR", full)$mean_rate)`.
  Which specific `r n_firearms_full_coverage` firearms got drawn for a given type is an
  extra source of study-to-study variability on top of the between-type sampling question
  beats 1–3 are about — visible in the boxplots' width, not just their position. (As the
  Monte Carlo section above shows, none of this spread is an artifact of too few
  replications — it's the actual between-study variability the design produces.)

# Why this counts as non-remediable

Nothing about the *analysis* is broken here — the reweighting procedure in
beat 2 is statistically valid and works, even with the added within-model
noise from sampling only a fraction of each type's firearms. What fails in beat 3 is
specific to the convenience sample's *design*: it never observed most of
the population it claims to generalize to, at the type level. No amount
of reanalysis of that particular dataset can fix a positivity violation
this severe — and adding a more realistic within-model layer doesn't
change that conclusion, it just makes the simulation's design closer to
how an actual black-box study is built (a handful of types, a handful of
firearms per type, many comparisons per firearm).

# Next steps

- Swap the updated numbers into `icfis/draft.qmd`'s Case Study 2 slides.
- Consider whether the "hypothetical better study" (beat 2) should use
  equal-per-type allocation or a different defensible design (e.g.,
  oversampling rare types further for power) — equal allocation was
  chosen for simplicity, not because it's the only valid choice.
- `firearm_difficulty_sd = 0.9` (in `make_firearm_population()`) is
  illustrative, not calibrated to the 0%–62% Beretta figure or anything
  else specific — worth a sensitivity check once there's time.
- If useful, add a version of beat 3 using *partial* type coverage (e.g.,
  4 of 8 types, not just the 2 easiest) to show the bias shrinking
  gradually as coverage improves, rather than only the two extremes.

# References

::: {#refs}
:::
