---
title: "Case Study: A Remediable Flaw (Inconclusive Handling)"
subtitle: "Flaw D -- recomputing error rates from the same raw responses"
bibliography: ../output/paper/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
    code-fold: true
---

# Goal

This is a first pass at the talk's "remediable flaw" case study, using
Flaw D (inconclusive responses treated as correct or ignored) rather than
Flaw E (invalid uncertainty measures / CI coverage), since Flaw D needs no
new simulation machinery — it's a pure reanalysis of the same raw
responses `icfis/analysis.qmd` already validated. If CI coverage (Flaw E)
gets built out later, this document either becomes a second case study or
stays as backup material.

The claim to demonstrate: **how a validation study handles inconclusive
responses is a reporting choice, not a design choice** — it can be redone
after the fact from the same raw data, unlike Flaw B (non-representative
sampling), which is baked into what was collected.

```{r setup}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)

source("R/simulate_dgp.R")
```

# The two ways error rates get reported

@cuellar2024methodological point out that most black-box studies handle
inconclusive responses by either excluding them or folding them into the
"correct" category — both of which understate the true error rate. They
recommend a bounding procedure instead: recompute the error rate twice,
once with every inconclusive response recoded as an identification, and
once with every inconclusive response recoded as an elimination. That
gives a defensible range instead of one falsely precise number.

Working through the algebra (see `icfis/analysis.qmd`'s benchmark
reconstruction) shows something sharper: for false positive and false
negative rates defined the way Monson, Smith, and Bajic (2023a) define
them — identifications divided by *all* different-source test cases,
inconclusives included in the denominator — the commonly **reported**
point estimate already equals one of the two bounds. It is not a neutral
middle ground.

- **False positive rate** (different-source pairs):
  - *Reported* $= \dfrac{\text{identifications}}{\text{identifications} + \text{inconclusives} + \text{eliminations}}$
    — identical to recoding every inconclusive as an *elimination*.
  - *Worst case* $= \dfrac{\text{identifications} + \text{inconclusives}}{\text{identifications} + \text{inconclusives} + \text{eliminations}}$
    — recoding every inconclusive as an *identification*.
- **False negative rate** (same-source pairs):
  - *Reported* $= \dfrac{\text{eliminations}}{\text{identifications} + \text{inconclusives} + \text{eliminations}}$
    — identical to recoding every inconclusive as an *identification*.
  - *Worst case* $= \dfrac{\text{eliminations} + \text{inconclusives}}{\text{identifications} + \text{inconclusives} + \text{eliminations}}$
    — recoding every inconclusive as an *elimination*.

Both numbers come from the same raw responses. Nothing about the study
needs to change to compute both — which is exactly what makes this flaw
remediable.

```{r define-rate-functions}
compute_reported_and_worst_case <- function(sim_data) {
  diff_data <- sim_data %>% filter(ground_truth_label == "different-source")
  same_data <- sim_data %>% filter(ground_truth_label == "same-source")

  diff_id <- sum(diff_data$response == "identification")
  diff_inc <- sum(diff_data$response == "inconclusive")
  diff_total <- nrow(diff_data)

  same_elim <- sum(same_data$response == "elimination")
  same_inc <- sum(same_data$response == "inconclusive")
  same_total <- nrow(same_data)

  tibble(
    rate = c("False positive rate", "False positive rate", "False negative rate", "False negative rate"),
    method = c("Reported", "Worst case", "Reported", "Worst case"),
    value = c(
      diff_id / diff_total,
      (diff_id + diff_inc) / diff_total,
      same_elim / same_total,
      (same_elim + same_inc) / same_total
    )
  )
}
```

# One example study

A single simulated study under the study's actual design (incomplete
assignment: 50 examiners, 30 of 300 items each) and the retuned defaults
from `icfis/analysis.qmd`.

```{r one-study}
example_study <- generate_sim_data_incomplete(
  seed = 1,
  n_examiners = incomplete_defaults$n_examiners,
  n_comparisons = incomplete_defaults$n_comparisons,
  items_per_examiner = incomplete_defaults$items_per_examiner,
  match_rate = incomplete_defaults$match_rate,
  false_positive_rate = incomplete_defaults$false_positive_rate,
  false_negative_rate = incomplete_defaults$false_negative_rate,
  inconclusive_rate = incomplete_defaults$inconclusive_rate,
  examiner_skill_sd = incomplete_defaults$examiner_skill_sd,
  examiner_inconclusive_sd = incomplete_defaults$examiner_inconclusive_sd,
  question_sd = incomplete_defaults$question_sd
)

knitr::kable(
  study_crosstab(example_study) %>%
    select(-replicate_id) %>%
    mutate(proportion = percent(proportion, accuracy = 0.1)),
  col.names = c("Ground truth", "Decision", "Count", "Proportion")
)
```

From this one study's raw responses, here are both ways of reporting the
error rate:

```{r one-study-rates}
example_rates <- compute_reported_and_worst_case(example_study) %>%
  mutate(rate = factor(rate, levels = c("False positive rate", "False negative rate")))

knitr::kable(
  example_rates %>% mutate(value = percent(value, accuracy = 0.1)),
  col.names = c("Rate", "Method", "Value")
)
```

Same raw data, same study, and the "worst case" number is dramatically
larger than what a study would typically report.

# Across 100 studies: is this a stable finding?

One study's numbers could be a fluke of that particular random draw.
Repeating the same comparison across 100 independent replicate studies
checks whether the gap between "reported" and "worst case" holds up.

```{r simulate-100-studies}
n_replications <- 100
set.seed(3031)
replicate_seeds <- sample.int(1e6, n_replications)

case_study_results <- purrr::map_dfr(seq_len(n_replications), function(r) {
  study_data <- generate_sim_data_incomplete(
    seed = replicate_seeds[r],
    n_examiners = incomplete_defaults$n_examiners,
    n_comparisons = incomplete_defaults$n_comparisons,
    items_per_examiner = incomplete_defaults$items_per_examiner,
    match_rate = incomplete_defaults$match_rate,
    false_positive_rate = incomplete_defaults$false_positive_rate,
    false_negative_rate = incomplete_defaults$false_negative_rate,
    inconclusive_rate = incomplete_defaults$inconclusive_rate,
    examiner_skill_sd = incomplete_defaults$examiner_skill_sd,
    examiner_inconclusive_sd = incomplete_defaults$examiner_inconclusive_sd,
    question_sd = incomplete_defaults$question_sd
  )

  compute_reported_and_worst_case(study_data) %>%
    mutate(replicate_id = r)
})
```

```{r summarize-100-studies}
case_study_summary <- case_study_results %>%
  group_by(rate, method) %>%
  summarise(
    mean_value = mean(value),
    sd_value = sd(value),
    .groups = "drop"
  ) %>%
  mutate(rate = factor(rate, levels = c("False positive rate", "False negative rate")))

knitr::kable(
  case_study_summary %>%
    mutate(
      mean_value = percent(mean_value, accuracy = 0.1),
      sd_value = percent(sd_value, accuracy = 0.1)
    ),
  col.names = c("Rate", "Method", "Mean across 100 studies", "SD across 100 studies")
)
```

```{r plot-100-studies, fig.width=8, fig.height=5}
case_study_plot_data <- case_study_results %>%
  mutate(
    rate = factor(rate, levels = c("False positive rate", "False negative rate")),
    method = factor(method, levels = c("Reported", "Worst case"))
  )

ggplot(case_study_plot_data, aes(x = method, y = value)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~ rate) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Reported vs. worst-case error rates across 100 simulated studies",
    subtitle = "Same underlying responses each time -- only how inconclusives are handled changes",
    x = NULL,
    y = "Rate"
  ) +
  theme_minimal()
```

The gap between "reported" and "worst case" is consistent across
replications, not an artifact of one unlucky draw — it's a structural
consequence of how large a share of responses are inconclusive at these
(retuned) defaults (roughly 37% either way; see `icfis/analysis.qmd`).

# Why this is remediable

Nothing above required a new study. The same 1,500 raw responses per
study support both numbers — the only thing that changed between them is
a reporting decision made *after* data collection. That's the definition
of remediable used throughout the talk: if the raw examiner-by-item
responses are available, a flaw in how they were originally summarized
can be fixed by whoever has the data, without going back into the field.

Contrast this with the talk's other case study (Flaw B, non-representative
sampling): no reanalysis of these same responses can produce a valid
estimate for a differently-composed population of examiners or firearms.
That's the structural difference the flowchart slide is pointing at, made
concrete with the same simulation machinery on both sides.

# Next steps

- Swap this into `icfis/draft.qmd`'s Case Study 1 slides in place of the
  CI-coverage placeholder — the "One example study" table and the
  "Reported vs. worst-case" boxplot are the two candidates for the
  Results slide.
- Consider whether to also show the *inconclusive rate itself* varying
  (e.g., under the contextual-bias / Hawthorne-effect mechanism from
  `icfis/analysis.qmd`'s backup slides) to connect this case study back to
  Flaw C.
- If time allows before the talk, revisit CI coverage (Flaw E) as a second
  remediable example — the firearm-clustering extension it needs is still
  unbuilt.

# References

::: {#refs}
:::
