---
title: "Remediable and Non-Remediable Flaws in Firearms Validity Studies"
subtitle: "A response to Chin et al. (2026)"
author:
  - name: "Amanda Luby"
  - name: "Maria Cuellar"
date: last-modified
bibliography: ../output/paper/references.bib
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: true
    footer: "ICFIS 2026"
    # TODO: swap in institution logo 
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)
```

<!--
TIMING BUDGET — talk is 30 minutes, excluding Q&A. Keep this comment for
rehearsal; delete before final export if desired.

  0:00–0:01  Title / hook
  0:01–0:06  1. Flaws + literature back-and-forth      (~5 min)
  0:06–0:10  2. Flowchart: remediable vs. non-remediable (~4 min)
  0:10–0:16  3. Data-generating process / study design  (~6 min)
  0:16–0:22  4. Case study 1: remediable flaw           (~6 min)
  0:22–0:28  5. Case study 2: non-remediable flaw        (~6 min)
  0:28–0:30  Recommendations / takeaways                 (~2 min)
  ------------------------------------------------------------
  Total: 30:00

RESOLVED: Case study 1 uses inconclusive handling (Flaw D), not CI coverage
(Flaw E) -- built and verified in icfis/case_study_inconclusive.qmd, which
sources the shared DGP in icfis/R/simulate_dgp.R (canonical defaults: maximum-
likelihood fit to the raw Ames II cartridge-case responses, with separate
same-/different-source inconclusive baselines -- see icfis/fit_ames2.qmd; the
earlier shared-baseline calibration is icfis/analysis.qmd). CI coverage
would need a nested firearm random effect the DGP doesn't have yet (see the
"Appendix: planned firearm-clustering extension" backup slide) -- revisit
only if there's time for a second remediable-flaw example.
-->

# Motivation

## Why this talk

- NAS (2009) and PCAST (2016): scientific validity of firearms examination has not been established
- Daubert Standard: Is there a **known error rate** for the technique?
- Recent large-scale black-box validation studies have attempted to remedy this (Ames I and II, Hicklin et al)
- @cuellar2024methodological: literature search of 28 black-box validation studies
  - Found six flaws (A–F) present, with minor exceptions, in **every** study
  - Concluded: none of the existing studies can establish scientific validity
- @chin2026towards responded: a flaw in one study doesn't doom the whole evidence base
  - Draws on evidence-based medicine / systematic-review practice
  - Argues for a **cumulative** forensic science: openness, registration, meta-analysis


## The six flaws {.smaller}

| Flaw | Description | Cuellar et al. framing |
|---|---|---|
| A | Inadequate sample size | Exacerbates the other flaws |
| B | Non-representative sample (examiners, firearms, ammunition) | Damages the data |
| C | Non-representative testing conditions (contextual bias, known-test effects) | Damages the data |
| D | Inconclusives treated as correct or ignored | Fixable if raw responses available |
| E | Invalid or nonexistent uncertainty measures | Fixable if raw responses available |
| F | Missing data handled inappropriately | Fixable if missingness can be characterized |


## Where we agree with Chin et al.

- No single study is perfect
- Not every flaw has the same consequences
- Analytic shortcomings may be **repairable after the fact** if the raw
  responses were recorded
- Imperfect studies are still useful!
- We should not discourage data collection just because a study has flaws 

## Main point of disagreement: 

> The problem is not simply that the wrong quantity was reported. The problem is that the data needed to estimate the relevant quantity were never collected in the first place.

- Some flaws (B, C) operate on **study design and sampling**, not analysis
- No amount of reanalysis recovers information that was never observed
- **Claim of this talk:** the six flaws split into two qualitatively different
  buckets — remediable (fixable via reanalysis) and non-remediable (fixable
  only with new data)


# The Remediable / Non-Remediable Distinction

## What "remediable" means here

Define the target quantity (the **estimand**):

> the population-average error rate (FPR / FNR), under casework-representative firearms, ammunition, examiners, and testing conditions

- **Remediable**: given the raw examiner-by-item responses, the estimand can
  still be validly estimated (possibly with extra assumptions / new analysis)
- **Non-remediable**: the sampled data cannot support the estimand, *regardless*
  of what analysis is applied to them


## Flowchart: sorting the six flaws

<!-- TODO: replace  -->

![img/flowchart.png]

## Remainder of the talk

1. Remediable flaw example

2. Non-remediable flaw example

## Approach: simulation study

- We can't go back and observe the "true" error rate for the 28 studies
- A simulation with a **known estimand** lets us show what each flaw does to a study's conclusions
- Same generative model powers both case studies



## Data generating model

- Each item $i$ has a latent difficulty $d_i \sim N(0, \sigma_d^2)$
- Each examiner $j$ has a latent skill $s_j \sim N(0, \sigma_s^2)$ **and** a latent tendency to call items inconclusive $\tau_j \sim N(0, \sigma_\tau^2)$
- Pairing item $i$ with examiner $j$ creates a decision challenge: $c_{ij} = d_i - s_j$
    - this becomes an "offset" from the overall estimand


## Response probabilities {.smaller}

Target estimand: overall population FPR or FNR or IR

A harder item or a weaker examiner makes the challenge larger, which pushes up both the chance of an error and the chance of an inconclusive call on the log-odds scale: 

$\pi_{ij}^{err} = P(\text{error}) = \text{expit}[\text{logit}(\text{FPR or FNR}) + c_{ij}]$

$\pi_{ij}^{inc} = P(\text{Inconclusive}) = \text{expit}[\text{logit}(\text{IR}_{G_i}) + 0.5 \cdot c_{ij} + \tau_j]$

The baseline inconclusive rate $\text{IR}_{G_i}$ can differ for same-source and different-source items.


## The response, formally {.smaller}

$e(\cdot)$ denotes an error and $c(\cdot)$ denotes a correct response. $G_i$ indicates ground truth (SS or DS)

$$
R_{ij} \mid G_i, c_{ij}, \tau_j \;\sim\; 
\begin{cases}
e(G_i) & \text{w.p. } \pi^{err}_{ij} \\
\text{inconclusive} & \text{w.p. } \pi^{inc}_{ij} \\
c(G_i) & \text{w.p. } 1 - \pi^{err}_{ij} - \pi^{inc}_{ij}
\end{cases}
$$

independently across $(i,j)$ given the latent variables above.


::: {.notes}
~1.5 min. This slide is the payoff of the notation build-up — say the plain-
English version too: "an error, an inconclusive, or a correct call, with
probabilities that shift based on how hard the item is and how skilled the
examiner is." Don't let the audience get lost in the categorical notation.
:::

## Calibrating to real studies

- Probability model above *cannot* capture every effect that arises in real data
- Goal is to create simulated studies that resemble previously-collected results, not to perfectly replicate a single study
- Some benchmarks that we used:  
    - Ames II cartridge cases (raw responses): baseline FPR, FNR, and inconclusive rates, fit by maximum likelihood
        - same-source and different-source items get *separate* inconclusive baselines: about 17% vs. 50%
    - Scurich et al (2025) blind-vs-known test inconclusive rates: 39.3% (blind) vs 56.4% (known)
    - Ames II: range of 0-62% "hard" comparison range across individual Beretta pistols

::: callout-warning
Disagree with calibrating to these numbers? We have an app for that! [add link]
:::

<!--
Previous calibration (before the raw Ames II data were available): a single
inconclusive baseline shared by same- and different-source items, tuned
against a benchmark reconstructed from @cuellar2024methodological's bounding
algebra for Monson, Smith, and Bajic (2023a) -- see icfis/analysis.qmd. That
version could not reproduce the same-source vs. different-source
inconclusive asymmetry (a ~13-point residual gap in opposite directions);
the maximum-likelihood fit to the raw data (icfis/fit_ames2.qmd) removes it.
-->

::: {.notes}
~1.5 min. This slide doubles as a credibility signal — show the audience
you're not hand-waving the parameters, *and* that you know exactly where
the model still falls short. The baselines are fit by maximum likelihood
to the raw Ames II cartridge-case responses (icfis/fit_ames2.qmd); simulated
studies at these defaults match all six cells of the real decision table to
within about a point, and the SD across examiners of inconclusive rates
(0.24 real, 0.24 simulated).

If pressed on "how well calibrated is this, really?": good on the response
distribution and examiner spread. Known misses: (1) firearm-level variation
-- real same-source inconclusive rates range from 3% to 70% across the 38
firearms, more than the model produces (backup slide candidate); (2) an
examiner's inconclusive tendency behaves differently on same- vs.
different-source items in the real data; (3) item and examiner-skill SD are
NOT identified by this study (each item was seen by one examiner), so those
two defaults are fixed at plausible values (0.8, 0.5), not estimated.
:::

## Example: 

```{r}
ames2 <- readxl::read_xlsx(here::here("real-data/ames2.xlsx"), sheet = 3) |>
  mutate(
    response_3 = case_when(stringr::str_detect(`Examiner's Analysis`, "Inc") ~ "Inconclusive",
    stringr::str_detect(`Examiner's Analysis`, "ID") ~ "Identification",
    stringr::str_detect(`Examiner's Analysis`, "Elim") ~ "Elimination"
    )
  )

ames2 |>
  select(c("M_NM", "response_3")) |>
  table() |>
  prop.table(1) |>
  round(2)
```

column 1: 1 study

[table]

column 2: 100 studies

[table]

# Case Study 1: A Remediable Flaw

## Setup: inconclusive handling (Flaw D)

- Black-box studies handle inconclusive responses inconsistently — excluded,
  folded into "correct," or otherwise treated as if they don't affect the
  error rate
- @cuellar2024methodological recommend a bounding procedure instead: recode
  every inconclusive as an identification, then as an elimination, and
  report **both** recomputed error rates as a range
- Claim: this is *remediable* — both numbers come from the same raw
  responses, recomputed after the fact


## Simulation design

- Same DGP as earlier, at the Ames II-calibrated defaults, under an incomplete design (50 examiners, 30 of 300 items each)
- For each simulated study, compute both numbers from the same raw
  responses:
  - **Reported**: identifications / all different-source cases (and the
    same-source analog for the false negative rate)
  - **Worst case**: recode every inconclusive as the error-favoring label,
    recompute
- Repeat across 100 replicate studies

::: {.notes}
~1.5 min.
:::

## Results

- One example study: FPR **0.6% → 51.1%**, FNR **1.8% → 23.8%** — same raw
  responses, same study, wildly different numbers depending on a reporting
  choice made after data collection
- Across 100 replicate studies (means): FPR **0.9% → 50.7%**, FNR
  **1.7% → 25.4%** — the gap holds up, it's not one unlucky draw
- The real Ames II cartridge-case responses give the same picture: FPR
  **0.9% → 51.5%**, FNR **1.8% → 25.7%**

::: {.notes}
~2 min. Lead with the one-study numbers (concrete, easy to say out loud),
then the 100-study means as the "this isn't a fluke" follow-up. The boxplot
figure should show both distributions barely overlapping.

The real-data bullet is a realism check, not independent evidence: the
simulation's defaults were fit to those responses. The point is that the
simulated gap is the size of the gap in an actual published study. The
asymmetry (larger gap for false positives) comes from inconclusives being much
more common on different-source items (about 50%) than same-source (about 23%).
:::

## Results

[add simulation showing the bounds under different inconclusive tendencies]

## Why this counts as remediable

- The fix required **no new data collection** — the same raw examiner-by-item
  responses already support both numbers; only how they're summarized
  changed
- This is exactly the kind of flaw Chin et al. are right about: a study with
  this flaw alone can still contribute to a cumulative evidence base, *once
  reanalyzed*
- On the other hand, these intervals are so large that this isn't really a "solution" to the error rate problem

# Case Study 2: A Non-Remediable Flaw

## Setup: non-representative sampling (Flaw B)

- Precondition for validity: firearms, ammunition, and examiners must
  represent the **full spectrum of real casework** [@cuellar2024methodological]
- None of the 28 studies demonstrate this
- Chin et al.'s response: non-representative samples can still contribute to
  a cumulative literature if combined with other studies 


## Ames II (bullets)

- 11 Ruger and 27 Beretta firearms were used to generate test cases
- [how were they paired?]

```{r}
# facet by ruger/beretta
# bars for each firearm
# fill by decision
```

## Two levels of variability

- Different firearm makes/models have different baseline difficulties
    - **Simulation:** choose 8 types

- Individual firearms *within* a make/model also have different difficulties
    - **Simulation:** assume a difficulty distribution for each firearm type; draw 5 firearms from that distribution

- 8 types $\times$ 5 of each type = 40 individual firearms used to generate test cases
    - Ames II: used ~30

## Three different "analysis" approaches:

1. **Biased estimate.** A convenience sample restricted to only two types (21% of casework by weight), 5 firearms each 
2. **The fix that could work.** A hypothetical better study covering all
   eight types (5 firearms each, not in population-correct proportions)
   with type recorded as a covariate — reweighting by the true population
   distribution afterward — Chin et al.'s "moderator" / big-team-science
   suggestion.
3. **The fix that doesn't exist.** Applying that same reweighting to the
   *actual* convenience sample from (1), which has zero data on the six
   unsampled types — 79% of the population's weight.

::: {.notes}
~2 min. This is the slide to get right — beat 2 is what keeps this honest
(we're not claiming reweighting is impossible in principle), beat 3 is what
makes it non-remediable *for the existing studies*. The two-level design
(types, then firearms within type) mirrors how an actual black-box study
is built.
:::

## Results

500 replicate studies per scenario (checked for Monte Carlo stability
against the first 100 replicates — the estimates moved by at most 0.03
percentage points, so 100 would have been enough; 500 is used here as the
final, most precise number). True population rate: **FPR 1.25%,
FNR 2.12%**.

- Convenience sample (naive): FPR **0.22%**, FNR **0.39%** — roughly
  **5–6x too low** on both
- Convenience sample, "reweighted": FPR **0.24%**, FNR **0.43%** — still
  roughly **5x too low** — barely moved
- Full-coverage study, reweighted: FPR **1.31%**, FNR **2.22%** — within
  ~5% of truth on both — reweighting works, even with only 5 firearms
  sampled per type

::: {.notes}
~2 min. The punchline is the third bullet: the same statistically valid
reweighting procedure that worked in beat 2 does almost nothing here. Not
because it was done wrong — because there's nothing to weight. If asked
about the firearm-level noise, it's a secondary point: even a
well-designed study (full type coverage) still needs enough firearms
*per* type, echoing case study 1's sample-size point one level down.
:::

## Why this counts as non-remediable

- No statistical procedure applied to *this* dataset can recover the
  population-average estimand — reweighting is valid and
  works, even accounting for within-model firearm sampling noise; the same
  procedure applied to the convenience sample fails, not from a math
  error, but because 79% of the population was never observed at the type
  level
- Fixing it requires **new data** collected under a design that
  characterizes the target population — not a different analysis of the
  same responses

::: {.notes}
~1 min.
:::

# Conclusion

## Takeaways

- Not all six flaws are alike — sorting them by *remediable vs. non-remediable*
  clarifies exactly where Chin et al.'s cumulative-science argument applies
- **Remediable** (A, D, E, and F with assumptions): reanalysis of existing
  data can recover a valid estimate 
- **Non-remediable** (B, C): the target quantity was never within reach of
  the sampled data — no reanalysis fixes this

::: {.notes}
~1 min.
:::

## Recommendations

- Report raw, examiner-by-item responses (not just summary error/inconclusive rates) 
    - enables post-hoc correction of remediable flaws
- Pre-register sampling frames
    - characterize the target population of firearms / examiners *before* data collection
- Use cluster-aware / MRMC statistical methods designed for the dependence structure of validation studies


::: {.notes}
~1.5 min.
:::

## Thank you

- Amanda Luby 
- Maria Cuellar
- Code and simulations: `github.com/mariacuellar/validity-firearms`

<!-- TODO: confirm repo URL / visibility before the talk -->

::: {.notes}
~0.5 min. Leave this slide up through Q&A.
:::

## References {.smaller}

::: {#refs}
:::

# Backup Slides

<!-- Not in the 30-minute budget — pull these up only if asked. -->

## Appendix: contextual-bias extension (Flaw C) {.smaller}

Hawthorne shift, calibrated to Scurich et al. (2025) blind vs. known-test
inconclusive rates at the Houston Forensic Science Center:

$$
\delta_H = \operatorname{logit}(0.564) - \operatorname{logit}(0.393)
$$

Under the "known test" condition, the inconclusive-probability equation
gains one term (error probability is unaffected):

$$
\pi^{inc}_{ij}(\text{condition}) = \operatorname{expit}\Big[\operatorname{logit}(p_{IC}) + 0.5\,c_{ij} + \tau_j + \delta_H \cdot \mathbb{1}\{\text{condition} = \text{"known test"}\}\Big]
$$

::: {.notes}
Backup only. Use if asked why the FPR/FNR barely moved between conditions
in the case-study-2 exploratory run — the shift is currently uniform across
easy and hard items, which is exactly why it doesn't move conclusive-only
error rates much. That's a modeling choice worth revisiting, not a result.
:::

## Appendix: missingness extension (Flaw F) {.smaller}

$$
\text{is\_error}_{ij} = \mathbb{1}\{R_{ij} = e(G_i)\}
$$

$$
\pi^{miss}_{ij} = \operatorname{expit}\Big[\operatorname{logit}(0.05) + 0.8\,c_{ij} + 1.2\cdot\text{is\_error}_{ij} + 0.6\cdot\mathbb{1}\{R_{ij}=\text{"inconclusive"}\}\Big]
$$

$$
M_{ij} \sim \text{Bernoulli}(\pi^{miss}_{ij}), \qquad
R^{obs}_{ij} = \begin{cases} R_{ij} & M_{ij} = 0 \\ \texttt{NA} & M_{ij} = 1 \end{cases}
$$

A missing-not-at-random mechanism: harder pairs, errors, and inconclusives
are all more likely to drop out, which is what biases naive (listwise-
deletion) error rates downward.

::: {.notes}
Backup only.
:::

## Appendix: what's left for CI coverage (Flaw E) {.smaller}

The individual-firearm random effect itself is **already built and in
use** — `make_firearm_population()` in `icfis/R/simulate_dgp_types.R`
gives each type $K$ individual firearms $k = 1, \dots, K$, each with
$f_k \sim N(0, \sigma_F^2)$, and case study 2 draws from it (5 firearms
per included type). What's *not* built is using that structure for CI
coverage:

- Case study 2 uses the firearm layer for a **bias** demonstration
  (does reweighting recover the population mean?) — it never constructs
  a confidence interval
- A Flaw E case study needs the same nested structure feeding an actual
  **interval-coverage** comparison: simulate many studies from a known
  true rate, compute a 95% CI three ways per study (naive Clopper–Pearson,
  cluster-robust, Gallas et al. 2009-style MRMC), and check what fraction
  of intervals actually contain the truth
- Responses for items sharing a firearm are **dependent** given $f_k$ —
  exactly what a naive interval ignores and a cluster-robust one accounts
  for; this is a different question from case study 2's, even though the
  generative structure is shared

::: {.notes}
Backup only — this is the concrete build item to report progress on if
asked "what's left to do." Don't say "not implemented" for the firearm
effect itself — it is; what's missing is the CI-coverage simulation loop
on top of it.
:::

## Appendix: is the "5x" just an artifact of our parameters? {.smaller}

Case study 2's `type_difficulty_sd`, `firearm_difficulty_sd`, and
`examiner_skill_sd` are uncalibrated. Checked a 3×3 grid (difficulty
heterogeneity × examiner heterogeneity, each Low/Medium/High; 50 replicate
studies per cell) — see `icfis/sensitivity_nonrepresentative.qmd`.

- **Bias ratio scales strongly with difficulty heterogeneity**: convenience
  sample is ~2–3x too low at Low, ~5x at Medium (the main case study's
  defaults), ~12–14x at High
- **Examiner (skill) heterogeneity has no consistent effect** — much smaller
  than the range difficulty heterogeneity produces, and not always in the
  same direction
- **The qualitative story holds at all 9 settings tested**: full-coverage
  reweighted stays within 0.81x–1.22x of truth everywhere; convenience
  "reweighted" never climbs out of 0.08x–0.53x
- **Conclusion**: the specific multiplier ("5x") is not an empirical claim
  — it moves with unvalidated assumptions. The *direction* and the
  reweighting-works/fails distinction don't.

::: {.notes}
Backup only. Pull this up if asked "isn't 5x just an artifact of your
parameter choices?" — yes, the number is; the direction and mechanism
aren't. That's the whole slide in one sentence if time is short.
:::
