---
title: "Fitting the DGP to the Ames II cartridge-case data"
subtitle: "How well does the talk's data-generating process describe a real black-box study?"
bibliography: ../output/paper/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
    code-fold: true
---

# Goal

`icfis/analysis.qmd` calibrated the DGP's three baseline rates to a
*reconstructed* benchmark (Cuellar et al.'s bounding numbers for Monson,
Smith, and Bajic 2023a). Now that the raw Ames II cartridge-case responses
are in `real-data/ames2.xlsx`, this document goes further: it **fits the DGP
in `icfis/R/simulate_dgp.R` to the raw responses by maximum likelihood** and
asks

1. Which parameter values does the data prefer, and how do they compare to
   the defaults used in the talk?
2. How well does the fitted model reproduce features of the real study it
   was *not* fit to (examiner-to-examiner spread, firearm-to-firearm spread,
   error counts, ...)?
3. What goes wrong — parameters the data cannot pin down, or features the
   model structurally cannot produce?

**Terminology.** "Previous talk defaults" below means the defaults the talk used
before this analysis: one inconclusive baseline shared by same- and
different-source items (36%), FPR 1%, FNR 2%, examiner inconclusive-tendency SD
0.35. They are frozen in this document as `legacy_defaults`. After this analysis,
`R/simulate_dgp.R` was updated to values based on model M1 (see "Implications for
the talk").

The fits below are implemented in this document's own likelihood, so they do
not depend on any change to the shared file. (Since this document was written,
`simulate_dgp.R` gained an optional `inconclusive_rate_same` argument that
implements the separate-baseline extension, M1, below; the default reproduces
the original model exactly.)

```{r setup}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)
library(tibble)

source(here::here("icfis/R/simulate_dgp.R"))

n_cores <- max(1, min(4, parallel::detectCores() - 1))
```

# The data

```{r load-data}
ames_raw <- readxl::read_xlsx(here::here("real-data/ames2.xlsx"), sheet = "Cartridge Cases")
stopifnot(ncol(ames_raw) == 24)

# Columns are renamed by position (two of the original names are long/truncated).
names(ames_raw) <- c(
  "orig_grp", "exam_id", "mailing_round", "packet", "set_number", "n_knowns",
  "analysis", "score", "pair", "anal_type", "gun_k1", "lot1", "gun_k2", "lot2",
  "gun_q", "lot3", "bullet_case", "m_nm", "ber", "firearm", "gun_2",
  "exam_id_2", "exam_id_3", "round"
)

# Same coding as icfis/analysis.qmd. "Unsuitable" and "did not provide answer"
# match none of the patterns and become NA.
ames_all <- ames_raw |>
  mutate(
    resp = case_when(
      stringr::str_detect(analysis, "Inc") ~ "inconclusive",
      stringr::str_detect(analysis, "ID") ~ "identification",
      stringr::str_detect(analysis, "Elim") ~ "elimination"
    ),
    truth = as.integer(m_nm == "M"),
    ground_truth_label = if_else(truth == 1, "same-source", "different-source"),
    nonber = as.integer(ber == "Non_Ber")
  )

n_dropped <- sum(is.na(ames_all$resp))

ames <- ames_all |>
  filter(!is.na(resp)) |>
  mutate(outcome = case_when(
    resp == "inconclusive" ~ "inc",
    truth == 1 & resp == "elimination" ~ "err",
    truth == 0 & resp == "identification" ~ "err",
    TRUE ~ "cor"
  ))
```

`r nrow(ames_all)` responses were recorded; `r n_dropped` are "unsuitable" or
"did not provide answer" and are dropped here (`r nrow(ames)` remain).
Dropping them is itself a Flaw-F-style choice (1.4% of responses), but the
published rates reproduce exactly with these rows excluded (next check), and it
barely matters at that size.

## Sanity check: does this reproduce the benchmark we reconstructed?

```{r benchmark-check}
raw_rates <- ames |>
  group_by(ground_truth_label) |>
  summarise(
    n = n(),
    id = mean(resp == "identification"),
    inc = mean(resp == "inconclusive"),
    elim = mean(resp == "elimination"),
    .groups = "drop"
  )

check <- tribble(
  ~quantity, ~from_raw_data, ~cuellar_2024,
  "FPR, inconclusive -> elimination", raw_rates$id[raw_rates$ground_truth_label == "different-source"], 0.0092,
  "FPR, inconclusive -> identification", with(raw_rates[raw_rates$ground_truth_label == "different-source", ], id + inc), 0.514,
  "FNR, inconclusive -> identification", raw_rates$elim[raw_rates$ground_truth_label == "same-source"], 0.0176,
  "FNR, inconclusive -> elimination", with(raw_rates[raw_rates$ground_truth_label == "same-source", ], elim + inc), 0.256
)

knitr::kable(
  check |> mutate(across(c(from_raw_data, cuellar_2024), ~ percent(.x, accuracy = 0.01))),
  col.names = c("Quantity", "From raw data", "Cuellar et al. (2024)")
)
```

The raw data reproduce all four bounding numbers, so the benchmark
reconstructed in `analysis.qmd` was exact, and this really is the study
those numbers came from.

## Structure of the study

```{r design-facts}
design_facts <- tibble(
  fact = c(
    "Responses (after dropping unsuitable / no answer)",
    "Examiners",
    "Responses per examiner (min - max)",
    "Packets of 15 comparisons",
    "Times any one (packet, item) is evaluated",
    "Share of responses that are same-source",
    "Share of responses on Beretta (vs. non-Beretta) firearms",
    "Distinct individual firearms recorded"
  ),
  value = c(
    nrow(ames),
    n_distinct(ames$exam_id),
    paste(range(count(ames, exam_id)$n), collapse = " - "),
    n_distinct(ames$orig_grp),
    paste(unique(count(ames_raw, orig_grp, set_number)$n), collapse = ", "),
    percent(mean(ames$truth == 1), accuracy = 0.1),
    percent(mean(ames$nonber == 0), accuracy = 0.1),
    n_distinct(ames$firearm)
  )
)
knitr::kable(design_facts, col.names = c("", ""))
```

Three features of the design matter for fitting:

- **Every item is seen exactly once.** Unlike the DGP's fully-crossed
  simulations, examiners are not crossed with items, so item difficulty is
  not separately observable by comparing examiners on the same item. Item
  effects can only be inferred from the *shape* of the response distribution.
- **Only 33% of comparisons are same-source**, not the DGP default of 50%.
  Ground truth is observed, so the likelihood below conditions on it and
  `match_rate` is not a parameter — but simulated studies used for goodness
  of fit reuse the real study's ground-truth pattern.
- **Firearm identity is recorded** (27 Beretta pistols, 11 non-Beretta), even
  though the DGP has no firearm effect. That lets us check a
  feature the model omits entirely.

```{r crosstab}
knitr::kable(
  ames |>
    count(ground_truth_label, resp) |>
    group_by(ground_truth_label) |>
    mutate(proportion = percent(n / sum(n), accuracy = 0.1)) |>
    ungroup() |>
    mutate(resp = factor(resp, levels = c("identification", "inconclusive", "elimination"))) |>
    arrange(ground_truth_label, resp),
  col.names = c("Ground truth", "Decision", "Count", "Proportion")
)
```

# The likelihood

Because each item appears once, the DGP's latent variables are independent
across responses *except* through the examiner. For examiner $j$ with
responses $R_{ij}$ on items of known ground truth $G_i$, the marginal
likelihood is

$$
L_j = \int\!\!\int \prod_{i \in j} \left[ \int P(R_{ij} \mid G_i, c = d - s, \tau) \, \phi_{\sigma_d}(d)\, \mathrm{d}d \right] \phi_{\sigma_s}(s)\, \phi_{\sigma_\tau}(\tau)\, \mathrm{d}s\, \mathrm{d}\tau ,
$$

where $P(R \mid G, c, \tau)$ is exactly the three-way response model in
`simulate_response()`, *including* its 0.95 cap on the combined error and
inconclusive probability. The inner integral over item difficulty is
one-dimensional; the outer two are examiner-level. All three are done by
Gauss–Hermite quadrature. Since the response probabilities depend on a
response only through its (ground truth, error / inconclusive / correct)
category, the inner integrals are computed once per category and each
examiner's log-likelihood is a matrix product of their category counts with
those values — so a likelihood evaluation takes milliseconds.

```{r likelihood}
# Gauss-Hermite nodes/weights for a standard normal (Golub-Welsch); weights sum to 1.
gh_normal <- function(n) {
  i <- seq_len(n - 1)
  J <- matrix(0, n, n)
  J[cbind(i, i + 1)] <- sqrt(i)
  J[cbind(i + 1, i)] <- sqrt(i)
  e <- eigen(J, symmetric = TRUE)
  list(z = e$values, w = e$vectors[1, ]^2)
}

# P(error), P(inconclusive), P(correct) for challenge `c` and examiner
# inconclusive tendency `tau`. Mirrors simulate_response() in R/simulate_dgp.R
# (checked by simulation below). `a_err` / `a_inc` are baseline logits.
response_probs <- function(c, tau, a_err, a_inc) {
  err <- plogis(a_err + c)
  inc <- plogis(a_inc + 0.5 * c + tau)
  total <- err + inc
  scale <- ifelse(total >= 0.95, 0.95 / total, 1)
  list(err = err * scale, inc = inc * scale, cor = 1 - (err + inc) * scale)
}
```

## Check: the likelihood's response model is the shared DGP's

Before trusting the likelihood, confirm `response_probs()` agrees with what
`simulate_response()` actually does, at a handful of (ground truth,
challenge, tendency) points — including one where the 0.95 cap binds.

```{r check-response-probs, cache=TRUE}
set.seed(11)
check_points <- tribble(
  ~truth, ~c,   ~tau,
  0,      0,    0,
  0,      1.5,  0.5,
  1,     -1,    1,
  1,      4,    3
)

fpr <- 0.02; fnr <- 0.05; ir <- 0.30
n_draws <- 20000

check_result <- pmap_dfr(check_points, function(truth, c, tau) {
  draws <- purrr::map_chr(seq_len(n_draws), ~ simulate_response(
    truth, c, tau, false_positive_rate = fpr, false_negative_rate = fnr, inconclusive_rate = ir
  ))
  p <- response_probs(c, tau, qlogis(if (truth == 1) fnr else fpr), qlogis(ir))
  error_label <- if (truth == 1) "elimination" else "identification"
  tibble(
    truth = truth, c = c, tau = tau,
    err_sim = mean(draws == error_label), err_formula = p$err,
    inc_sim = mean(draws == "inconclusive"), inc_formula = p$inc
  )
})

knitr::kable(check_result, digits = 3)
```

Simulated frequencies (`n = 20,000` draws each) match the formula to within
Monte Carlo error at every point.

## Fitting machinery

Three models, fit by box-constrained maximum likelihood (`L-BFGS-B`, several
starting points each):

- **M0 — the DGP as it stands.** Six parameters: baseline logit FPR, FNR, and
  a *single* inconclusive rate shared by same- and different-source items,
  plus the three heterogeneity SDs $\sigma_d$ (item), $\sigma_s$ (examiner
  skill), $\sigma_\tau$ (examiner inconclusive tendency).
- **M1 — separate inconclusive baselines.** Same, but same-source and
  different-source items get their own baseline inconclusive rate. This is the
  extension `analysis.qmd` said the benchmark asymmetry called for.
- **M2 — M1 plus a firearm-type shift.** A single additive shift $\delta$ to
  the decision challenge for non-Beretta items, to see whether a
  make/model-level difficulty difference (as in case study 2) improves the fit.

SDs are bounded to $[0.02, 3]$. The upper bound is not arbitrary: it is 1.5x
the app's slider maximum and is where the quadrature is still accurate
(verified at 120 nodes below). Whether the fits *reach* the bound is itself a
diagnostic.

```{r fit-functions}
# Category columns: truth x nonber x outcome, outcome varying fastest.
cell_grid <- expand.grid(outcome = c("err", "inc", "cor"), nonber = 0:1, truth = 0:1, stringsAsFactors = FALSE)
cell_names <- with(cell_grid, paste(truth, nonber, outcome))

count_matrix <- ames |>
  mutate(cell = paste(truth, nonber, outcome)) |>
  count(exam_id, cell) |>
  pivot_wider(names_from = cell, values_from = n, values_fill = 0)
N <- as.matrix(count_matrix[, cell_names])

# theta layouts (unconstrained: logits for rates, log for SDs):
#   M0: fpr, fnr, ir,          log sd_d, log sd_s, log sd_t
#   M1: fpr, fnr, ir_D, ir_S,  log sd_d, log sd_s, log sd_t
#   M2: fpr, fnr, ir_D, ir_S, delta, log sd_d, log sd_s, log sd_t
unpack <- function(theta, model) {
  switch(model,
    M0 = list(fpr = theta[1], fnr = theta[2], irD = theta[3], irS = theta[3], delta = 0, sd = exp(theta[4:6])),
    M1 = list(fpr = theta[1], fnr = theta[2], irD = theta[3], irS = theta[4], delta = 0, sd = exp(theta[5:7])),
    M2 = list(fpr = theta[1], fnr = theta[2], irD = theta[3], irS = theta[4], delta = theta[5], sd = exp(theta[6:8]))
  )
}

make_loglik <- function(Q) {
  gs <- gh_normal(Q); gt <- gh_normal(Q); gd <- gh_normal(Q)
  log_w <- as.vector(outer(log(gs$w), log(gt$w), "+"))

  # (err, inc, cor) probabilities, integrated over item difficulty, on the
  # (examiner skill, examiner tendency) grid, for one ground-truth x type cell.
  cell_probs <- function(a_err, a_inc, shift, sd) {
    c_mat <- shift + outer(-sd[2] * gs$z, sd[1] * gd$z, "+")   # skill x item
    out <- matrix(0, 3, Q * Q)
    for (b in seq_len(Q)) {
      p <- response_probs(c_mat, sd[3] * gt$z[b], a_err, a_inc)
      idx <- (b - 1) * Q + seq_len(Q)
      out[1, idx] <- p$err %*% gd$w
      out[2, idx] <- p$inc %*% gd$w
      out[3, idx] <- p$cor %*% gd$w
    }
    out
  }

  function(theta, model) {
    p <- unpack(theta, model)
    probs <- matrix(NA_real_, 12, Q * Q)
    for (tr in 0:1) for (nb in 0:1) {
      probs[tr * 6 + nb * 3 + 1:3, ] <- cell_probs(
        a_err = if (tr == 1) p$fnr else p$fpr,
        a_inc = if (tr == 1) p$irS else p$irD,
        shift = nb * p$delta,
        sd = p$sd
      )
    }
    ll <- sweep(N %*% log(probs), 2, log_w, "+")
    m <- apply(ll, 1, max)
    sum(m + log(rowSums(exp(ll - m))))
  }
}

loglik <- make_loglik(40)
loglik_fine <- make_loglik(120)

bounds <- function(model) {
  n_ir <- if (model == "M0") 1 else 2
  list(
    lower = c(rep(qlogis(1e-4), 2), rep(qlogis(1e-3), n_ir), if (model == "M2") -4, rep(log(0.02), 3)),
    upper = c(rep(qlogis(0.9), 2), rep(qlogis(0.95), n_ir), if (model == "M2") 4, rep(log(3), 3))
  )
}

fit_once <- function(start, model, ll = loglik) {
  b <- bounds(model)
  start <- pmin(pmax(start, b$lower), b$upper)
  optim(start, function(p) -ll(p, model), method = "L-BFGS-B",
        lower = b$lower, upper = b$upper, control = list(maxit = 500, factr = 1e7))
}

fit_multistart <- function(model, start, n_starts = 4, seed = 1) {
  set.seed(seed)
  starts <- c(list(start), replicate(n_starts - 1, start + rnorm(length(start), 0, 0.8), simplify = FALSE))
  fits <- parallel::mclapply(starts, fit_once, model = model, mc.cores = n_cores)
  best <- fits[[which.min(map_dbl(fits, "value"))]]
  best$all_ll <- -map_dbl(fits, "value")
  best
}
```

```{r fit-models, cache=TRUE}
# The talk's defaults *before* this fit (a single inconclusive baseline shared by
# same- and different-source items). Frozen here so this document's comparisons
# stay true even though R/simulate_dgp.R has since been updated to the M1-based
# values. Also used as the starting point for the fits.
legacy_defaults <- list(
  false_positive_rate = 0.01, false_negative_rate = 0.02, inconclusive_rate = 0.36,
  question_sd = 0.8, examiner_skill_sd = 0.7, examiner_inconclusive_sd = 0.35
)

talk_theta <- c(
  fpr = qlogis(legacy_defaults$false_positive_rate),
  fnr = qlogis(legacy_defaults$false_negative_rate),
  ir = qlogis(legacy_defaults$inconclusive_rate),
  sd_d = log(legacy_defaults$question_sd),
  sd_s = log(legacy_defaults$examiner_skill_sd),
  sd_t = log(legacy_defaults$examiner_inconclusive_sd)
)

fit0 <- fit_multistart("M0", unname(talk_theta))
fit1 <- fit_multistart("M1", unname(c(fit0$par[1:3], fit0$par[3], fit0$par[4:6])))
fit2 <- fit_multistart("M2", unname(c(fit1$par[1:4], 0, fit1$par[5:7])))

fits <- list(M0 = fit0, M1 = fit1, M2 = fit2)
```

## Do the fits agree across starting points, and is the quadrature accurate?

```{r fit-diagnostics}
diag_tbl <- tibble(
  model = names(fits),
  `best log-lik` = map_dbl(fits, ~ -.x$value),
  `spread across starts` = map_dbl(fits, ~ diff(range(.x$all_ll))),
  `log-lik at 120 nodes` = map2_dbl(fits, names(fits), ~ loglik_fine(.x$par, .y))
)
knitr::kable(diag_tbl, digits = 3)
```

For M1 and M2, random restarts all land on the same optimum, and 40- vs.
120-node quadrature agree to a few thousandths of a log-likelihood unit, so
those fits reflect the model, not the numerics. **M0 is different**: its
restarts disagree by several log-likelihood units, i.e. its likelihood has
multiple modes or a long ridge and the reported optimum is the best of a
badly-behaved set. That is a result about M0, discussed below, not an
optimizer failure — M1 is well-behaved under identical machinery.

# Results

## Fitted parameters

```{r param-table}
talk_p <- unpack(unname(c(talk_theta[1:3], talk_theta[3], talk_theta[4:6])), "M1")

param_row <- function(p, label, k, ll) {
  tibble(
    Model = label,
    FPR = plogis(p$fpr), FNR = plogis(p$fnr),
    IR_diff = plogis(p$irD), IR_same = plogis(p$irS),
    delta = p$delta,
    sd_item = p$sd[1], sd_skill = p$sd[2], sd_incon = p$sd[3],
    k = k, loglik = ll
  )
}

param_tbl <- bind_rows(
  param_row(talk_p, "Previous talk defaults (fixed)", 0, loglik(unname(c(talk_theta[1:3], talk_theta[4:6])), "M0")),
  param_row(unpack(fit0$par, "M0"), "M0: DGP as-is", 6, -fit0$value),
  param_row(unpack(fit1$par, "M1"), "M1: separate inconclusive baselines", 7, -fit1$value),
  param_row(unpack(fit2$par, "M2"), "M2: M1 + non-Beretta shift", 8, -fit2$value)
) |>
  mutate(AIC = 2 * k - 2 * loglik)

knitr::kable(
  param_tbl |>
    mutate(
      across(c(FPR, FNR), ~ percent(.x, accuracy = 0.01)),
      across(c(IR_diff, IR_same), ~ percent(.x, accuracy = 0.1)),
      across(c(delta, sd_item, sd_skill, sd_incon, loglik, AIC), ~ round(.x, 2))
    ),
  col.names = c("Model", "Baseline FPR", "Baseline FNR", "Inconclusive (diff.-source)", "Inconclusive (same-source)",
                "Non-Beretta shift", "SD item", "SD examiner skill", "SD examiner inconclusive", "Params", "Log-lik", "AIC")
)
```

Baseline parameters are on the DGP's logit scale and are *not* the realized
marginal rates (see Jensen's-inequality note in `analysis.qmd`); realized
rates appear in the goodness-of-fit section.

```{r unpack-best, include=FALSE}
b0 <- unpack(fit0$par, "M0")
b1 <- unpack(fit1$par, "M1")
```

Reading the table:

- **The previous talk defaults are far from the data's preference.** Holding
  everything else fixed, the defaults have a log-likelihood of
  `r round(param_tbl$loglik[1], 1)` against `r round(param_tbl$loglik[2], 1)` for the
  best-fitting version of the *same* model — a difference of
  `r round(param_tbl$loglik[2] - param_tbl$loglik[1], 1)` log-likelihood units on six
  parameters. The biggest single gap is examiner-level inconclusive heterogeneity:
  the data want $\sigma_\tau$ around `r round(b1$sd[3], 1)`, versus the 0.35 used in the talk.
- **M0 (the DGP as it stands) does not settle on an interior solution.** Its
  item SD runs to the upper bound (`r round(b0$sd[1], 2)`) and its baseline FPR / FNR
  collapse to `r percent(plogis(b0$fpr), accuracy = 0.01)` /
  `r percent(plogis(b0$fnr), accuracy = 0.01)` — which looks like the fit reaching for a very
  heavy-tailed item-difficulty distribution to generate rare errors, rather than using a baseline
  rate. Its restarts also disagree (previous section). Read this as the likelihood straining
  against a structural limit of M0 (next section), not as a believable parameter set.
- **M1 fixes most of the problem with one extra parameter.** Allowing
  same- and different-source items their own inconclusive baseline improves the
  log-likelihood by `r round(-fit1$value + fit0$value, 1)` (AIC drops by
  `r round(param_tbl$AIC[2] - param_tbl$AIC[3], 0)`) and yields an interior, sensible
  solution: baseline FPR `r percent(plogis(b1$fpr), accuracy = 0.1)`, FNR
  `r percent(plogis(b1$fnr), accuracy = 0.1)`, and inconclusive baselines of
  `r percent(plogis(b1$irD), accuracy = 1)` (different-source) vs.
  `r percent(plogis(b1$irS), accuracy = 1)` (same-source). The 13-point residual gap
  that `analysis.qmd` flagged as "known and accepted" is not a small detail — it is the single
  largest structural misfit.
- **A firearm-type shift (M2) adds almost nothing.** The non-Beretta shift
  improves the log-likelihood by only `r round(-fit2$value + fit1$value, 2)` —
  not worth its parameter by AIC — even though non-Beretta same-source items really are
  harder (next section). One shift on a single difficulty axis cannot describe a
  difference that goes in *opposite directions* for same- and different-source items.

## Which parameters does the data actually determine? Profile likelihoods

For each heterogeneity SD, this fixes it at a grid of values, re-optimizes
everything else, and plots the drop in log-likelihood from the best fit.
Values within 1.92 (dashed line) of the maximum are inside an approximate 95%
interval. A flat curve means the data cannot distinguish those values.

```{r profiles, cache=TRUE}
profile_sd <- function(model, which_sd, grid, fit) {
  n_theta <- length(fit$par)
  sd_index <- n_theta - 3 + which_sd
  b <- bounds(model)

  parallel::mclapply(grid, function(v) {
    f <- optim(
      fit$par[-sd_index],
      function(p) {
        th <- numeric(n_theta); th[sd_index] <- log(v); th[-sd_index] <- p
        -loglik(th, model)
      },
      method = "L-BFGS-B", lower = b$lower[-sd_index], upper = b$upper[-sd_index],
      control = list(maxit = 300, factr = 1e7)
    )
    tibble(model = model, parameter = c("item", "examiner skill", "examiner inconclusive")[which_sd],
           value = v, loglik = -f$value)
  }, mc.cores = n_cores) |> bind_rows()
}

sd_grid <- c(0.02, 0.25, 0.5, 0.8, 1.2, 1.8, 3)

profile_results <- bind_rows(
  map_dfr(1:3, ~ profile_sd("M0", .x, sd_grid, fit0)),
  map_dfr(1:3, ~ profile_sd("M1", .x, sd_grid, fit1))
) |>
  left_join(tibble(model = c("M0", "M1"), max_ll = c(-fit0$value, -fit1$value)), by = "model") |>
  mutate(
    drop = pmin(loglik - max_ll, 0),
    parameter = factor(parameter, levels = c("item", "examiner skill", "examiner inconclusive")),
    model = recode(model, M0 = "M0: DGP as-is", M1 = "M1: separate inconclusive baselines")
  )
```

```{r plot-profiles, fig.width=10, fig.height=5}
ggplot(profile_results, aes(x = value, y = drop)) +
  geom_hline(yintercept = -1.92, linetype = "dashed", color = "grey50") +
  geom_line() +
  geom_point() +
  facet_grid(model ~ parameter, scales = "free_y") +
  labs(
    title = "Profile log-likelihood for each heterogeneity SD",
    subtitle = "Drop from the maximum; dashed line = approximate 95% cutoff",
    x = "Fixed value of the SD (latent logit scale)",
    y = "Log-likelihood minus maximum"
  ) +
  theme_minimal()
```

# Goodness of fit

A good likelihood is not a good model. To ask what the fitted models get
wrong, simulate *new studies with the real study's design* — the same 173
examiners with their actual numbers of responses, and the same ground truth,
firearm type, and firearm for every response — from each fitted parameter
set, using the shared `simulate_response()`. Then compare summaries the
likelihood was never asked to match against the same summaries computed on the
real data. (Plug-in parameters, so this ignores parameter uncertainty; for
features off by many multiples that does not change the conclusion.)

Three parameter sets are compared: the **previous talk defaults**, **M0**, and **M1**.

```{r gof-functions}
gof_design <- ames |> select(exam_id, truth, nonber, firearm, resp)

simulate_fitted_study <- function(p, design, seed) {
  set.seed(seed)
  examiners <- distinct(design, exam_id) |>
    mutate(skill = rnorm(n(), 0, p$sd[2]), tendency = rnorm(n(), 0, p$sd[3]))

  sim <- design |>
    select(-resp) |>
    left_join(examiners, by = "exam_id") |>
    mutate(challenge = nonber * p$delta + rnorm(n(), 0, p$sd[1]) - skill)

  sim$resp <- purrr::pmap_chr(
    list(
      ground_truth = sim$truth,
      decision_challenge = sim$challenge,
      examiner_inconclusive_tendency = sim$tendency,
      inconclusive_rate = if_else(sim$truth == 1, plogis(p$irS), plogis(p$irD))
    ),
    simulate_response,
    false_positive_rate = plogis(p$fpr),
    false_negative_rate = plogis(p$fnr)
  )
  sim
}

# Summaries that the likelihood was not directly fit to.
gof_stats <- function(df) {
  df <- df |> mutate(is_err = (truth == 1 & resp == "elimination") | (truth == 0 & resp == "identification"),
                     is_inc = resp == "inconclusive")

  by_ex <- df |>
    group_by(exam_id) |>
    summarise(
      inc = mean(is_inc), err = mean(is_err),
      inc_ss = mean(is_inc[truth == 1]), inc_ds = mean(is_inc[truth == 0]),
      n_ss = sum(truth == 1), n_ds = sum(truth == 0), any_err = any(is_err),
      .groups = "drop"
    )
  by_fw <- df |> filter(truth == 1) |> group_by(firearm) |> summarise(inc = mean(is_inc), .groups = "drop")
  both <- by_ex |> filter(n_ss >= 3, n_ds >= 3)

  cells <- df |>
    count(truth, resp) |>
    group_by(truth) |>
    mutate(p = n / sum(n)) |>
    ungroup()
  cell_p <- setNames(cells$p, paste0("cell_", if_else(cells$truth == 1, "same", "diff"), "_", cells$resp))

  type_inc <- df |> group_by(truth, nonber) |> summarise(inc = mean(is_inc), .groups = "drop")
  type_p <- setNames(type_inc$inc, paste0("inc_", if_else(type_inc$truth == 1, "same", "diff"), "_", if_else(type_inc$nonber == 1, "nonber", "ber")))

  scalars <- c(
    cell_p, type_p,
    n_errors = sum(df$is_err),
    n_examiners_with_error = sum(by_ex$any_err),
    sd_examiner_inc = sd(by_ex$inc),
    cor_examiner_inc_same_vs_diff = cor(both$inc_ss, both$inc_ds),
    cor_examiner_inc_vs_err = cor(by_ex$inc, by_ex$err),
    sd_firearm_inc_same = sd(by_fw$inc)
  )
  list(scalars = scalars, examiner_sorted = sort(by_ex$inc), firearm_sorted = sort(by_fw$inc))
}

n_sims <- 200

gof_params <- list(
  "Previous talk defaults" = talk_p,
  "M0: DGP as-is" = unpack(fit0$par, "M0"),
  "M1: separate inconclusive baselines" = unpack(fit1$par, "M1")
)
```

```{r run-gof, cache=TRUE}
gof_seeds <- 9000 + seq_len(n_sims)

gof_sims <- imap(gof_params, function(p, label) {
  parallel::mclapply(gof_seeds, function(sd) gof_stats(simulate_fitted_study(p, gof_design, sd)), mc.cores = n_cores)
})
gof_obs <- gof_stats(gof_design)
```

## Summary statistics: real study vs. fitted models

Each row is a summary of the real study; columns show the range spanned by the
middle 95% of the `r n_sims` simulated studies from each parameter set. A
"p" is the two-sided Monte Carlo tail probability of a simulated value at
least as extreme as the real one; `< .01` means that in none (or almost none) of the
`r n_sims` simulated studies did the model produce something as extreme as
the real data.

```{r gof-table}
stat_labels <- c(
  cell_diff_identification = "Different-source: % identification (FPR)",
  cell_diff_inconclusive = "Different-source: % inconclusive",
  cell_same_inconclusive = "Same-source: % inconclusive",
  cell_same_elimination = "Same-source: % elimination (FNR)",
  n_errors = "Total errors",
  n_examiners_with_error = "Examiners with at least one error",
  sd_examiner_inc = "SD across examiners of inconclusive rate",
  cor_examiner_inc_same_vs_diff = "Across examiners: cor(inconclusive rate on same- vs. different-source items)",
  cor_examiner_inc_vs_err = "Across examiners: cor(inconclusive rate, error rate)",
  sd_firearm_inc_same = "SD across firearms of same-source inconclusive rate",
  inc_same_ber = "Same-source, Beretta: % inconclusive",
  inc_same_nonber = "Same-source, non-Beretta: % inconclusive",
  inc_diff_ber = "Different-source, Beretta: % inconclusive",
  inc_diff_nonber = "Different-source, non-Beretta: % inconclusive"
)

is_pct <- function(nm) grepl("^(cell_|inc_)", nm)
fmt <- function(x, nm) if (is_pct(nm)) percent(x, accuracy = 0.1) else if (grepl("^n_", nm)) as.character(round(x)) else sprintf("%.2f", x)

gof_table <- map_dfr(names(stat_labels), function(nm) {
  obs <- gof_obs$scalars[[nm]]
  row <- tibble(Statistic = stat_labels[[nm]], Observed = fmt(obs, nm))
  for (label in names(gof_sims)) {
    sims <- map_dbl(gof_sims[[label]], ~ .x$scalars[[nm]])
    sims <- sims[is.finite(sims)]
    p <- min(1, 2 * min(mean(sims >= obs), mean(sims <= obs)))
    range_txt <- paste0(fmt(quantile(sims, 0.025), nm), " to ", fmt(quantile(sims, 0.975), nm))
    row[[label]] <- paste0(range_txt, if (p < 0.01) " (p < .01)" else sprintf(" (p = %.2f)", p))
  }
  row
})

knitr::kable(gof_table)
```

## Examiner-to-examiner spread in inconclusive rates

For each model, the sorted inconclusive rate of the 173 examiners in the real
study (points) against the middle 95% envelope of the same order statistics in
simulated studies (band).

```{r plot-examiner-spread, fig.width=10, fig.height=4.5}
envelope <- function(key) {
  imap_dfr(gof_sims, function(sims, label) {
    mat <- do.call(rbind, map(sims, key))
    tibble(model = label, rank = seq_len(ncol(mat)),
           lo = apply(mat, 2, quantile, 0.025), hi = apply(mat, 2, quantile, 0.975))
  })
}

obs_examiner <- tibble(rank = seq_along(gof_obs$examiner_sorted), rate = gof_obs$examiner_sorted)

ggplot(envelope(~ .x$examiner_sorted), aes(x = rank)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey75") +
  geom_point(data = obs_examiner, aes(y = rate), size = 0.8, color = "#D55E00") +
  facet_wrap(~ model) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Examiner-level inconclusive rates: real study vs. simulated studies",
    subtitle = "Points: 173 real examiners, sorted. Band: 95% envelope across simulated studies",
    x = "Examiner (sorted by inconclusive rate)", y = "Inconclusive rate"
  ) +
  theme_minimal()
```

## Firearm-to-firearm spread in same-source inconclusive rates

The DGP has no firearm effect. If firearms really differ in difficulty, real
firearm-level inconclusive rates should be more spread out than any of these
models can produce, even though each model was fit to (and reproduces) the
overall inconclusive rate.

```{r plot-firearm-spread, fig.width=10, fig.height=4.5}
obs_firearm <- tibble(rank = seq_along(gof_obs$firearm_sorted), rate = gof_obs$firearm_sorted)

ggplot(envelope(~ .x$firearm_sorted), aes(x = rank)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey75") +
  geom_point(data = obs_firearm, aes(y = rate), size = 1.2, color = "#D55E00") +
  facet_wrap(~ model) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Firearm-level same-source inconclusive rates: real study vs. simulated studies",
    subtitle = "Points: 38 real firearms (about 30 same-source comparisons each), sorted. Band: 95% envelope across simulated studies",
    x = "Firearm (sorted by inconclusive rate)", y = "Inconclusive rate on same-source comparisons"
  ) +
  theme_minimal()
```

# What we learn

## What the DGP gets right

Once same- and different-source items are allowed separate inconclusive
baselines (M1), the fitted model reproduces the real study's headline
numbers and one structural feature it was not directly fit to:

- All six cells of the ground truth x decision crosstab, including the
  realized FPR (real `r percent(gof_obs$scalars[["cell_diff_identification"]], accuracy = 0.1)`)
  and FNR (real `r percent(gof_obs$scalars[["cell_same_elimination"]], accuracy = 0.1)`).
- The number of errors (real `r gof_obs$scalars[["n_errors"]]`) and of examiners who made at least one
  (real `r gof_obs$scalars[["n_examiners_with_error"]]`).
- **The examiner-to-examiner spread in inconclusive rates** (SD across examiners:
  real `r sprintf("%.2f", gof_obs$scalars[["sd_examiner_inc"]])`), even though the
  likelihood never targeted that summary directly. The examiner-level plot above
  shows the real distribution sitting essentially inside the simulated envelope from one end to the other.

## What goes wrong

1. **The current DGP cannot match both inconclusive rates.** M0's single shared
   inconclusive baseline lands at about 41% for both ground-truth categories, against
   `r percent(gof_obs$scalars[["cell_diff_inconclusive"]], accuracy = 1)` (different-source) and
   `r percent(gof_obs$scalars[["cell_same_inconclusive"]], accuracy = 1)` (same-source) in the
   real data — a ~27-point gap between the two that no shared baseline can produce. Freeing
   the two baselines (M1) is worth `r round(-fit1$value + fit0$value, 0)` log-likelihood units and
   makes M0's odd parameter values (item SD at its bound, near-zero baseline error rates) go
   away. In `analysis.qmd` this asymmetry was labeled a small residual to be accepted; on the
   raw data it is the dominant structural misfit.
2. **The previous talk defaults badly understate examiner heterogeneity, and overshoot errors.**
   With $\sigma_\tau = 0.35$ the simulated SD of examiner inconclusive rates is about
   0.13–0.16 against a real 0.24, and the examiner-level plot shows the real spread
   extending well outside the envelope on both ends. The data want $\sigma_\tau \approx$
   `r round(b1$sd[3], 1)`. On this study's design, the defaults also produce 76–117 errors (real: 51)
   — i.e., FPR/FNR roughly 1.5–2x too high, consistent with the ~2x that `analysis.qmd` accepted, but
   now shown not to be sampling noise.
3. **Item and examiner-skill heterogeneity are not identified by this study.**
   The profile likelihoods are flat over most of the range for $\sigma_d$ (item) and
   $\sigma_s$ (skill) under M1: values from ~0 up to about 1 are all
   within the 95% cutoff for skill, and item SD is best at its lower bound and
   only mildly penalized out to ~1.2. Only $\sigma_\tau$ is sharply determined.
   That is what the design implies: every item is seen by one examiner, so
   difficulty and skill can only enter through the (rare — 51 in 4,258 responses)
   errors and through a half-weight on inconclusives. Under M0 the item SD is instead
   pulled to its upper bound and the restarts disagree. **Consequence:** these
   sliders in `app.R` cannot be set from this study, and any conclusion
   that depends on them should be shown across a range (as
   `sensitivity_nonrepresentative.qmd` already does for case study 2), not at a fixed default.
4. **Examiner inconclusive tendency is not one number per examiner.** The DGP gives each
   examiner a single $\tau_j$ applied to same- and different-source items alike. In the
   real data, an examiner's inconclusive rate on same-source items is only weakly
   correlated with their rate on different-source items
   (real `r sprintf("%.2f", gof_obs$scalars[["cor_examiner_inc_same_vs_diff"]])`), while
   the fitted models predict about 0.6–0.8. Examiners appear to behave differently on
   the two kinds of comparison — not just more or less inconclusive overall.
5. **Firearms differ far more than the model allows.** Across the 38 firearms, the SD of the
   same-source inconclusive rate is `r sprintf("%.2f", gof_obs$scalars[["sd_firearm_inc_same"]])`
   in the real data vs. 0.06–0.10 in every fitted model; the firearm-level plot shows the
   real firearms at both extremes (3% to 70%) outside the envelope. This is direct empirical
   support for the firearm random effect already used in case study 2
   (`make_firearm_population()`), and is consistent with the 0%–62% range across Beretta
   pistols that Cuellar et al. cite.
6. **A single "difficulty" axis cannot describe make/model differences.** Non-Beretta
   same-source items are harder (more inconclusive: 32% vs. 18%), but non-Beretta
   *different*-source items are *less* often inconclusive (44% vs. 53%). A shift that
   makes items harder for both truth values (M2) fits neither well; its
   improvement over M1 is `r round(-fit2$value + fit1$value, 2)` log-likelihood units. Difficulty here is
   not one dimension shared by same- and different-source comparisons.
7. **Errors and inconclusives are (weakly) negatively related across examiners in the
   real data** (real correlation `r sprintf("%.2f", gof_obs$scalars[["cor_examiner_inc_vs_err"]])`),
   i.e., more cautious examiners err less — a trade-off. In the DGP a harder challenge raises
   both error and inconclusive probabilities, so examiners with low skill are
   simultaneously more error-prone and more inconclusive, which pushes the
   correlation the other way. The evidence here is thin (51 errors) and the real
   value sits at the edge of the simulated envelope, so treat it as a
   flag rather than a finding.

## Implications for the talk

- **Case study 1 gets stronger, not weaker.** On the raw Ames II responses,
  `r percent(with(raw_rates[raw_rates$ground_truth_label == "different-source", ], inc), accuracy = 0.1)`
  of different-source and
  `r percent(with(raw_rates[raw_rates$ground_truth_label == "same-source", ], inc), accuracy = 0.1)`
  of same-source responses are inconclusive; the false positive rate is
  `r percent(check$from_raw_data[1], accuracy = 0.1)` if inconclusives are counted as eliminations and
  `r percent(check$from_raw_data[2], accuracy = 0.1)` if counted as identifications. That is a real study,
  not a simulation. The simulated case study, run at the talk's symmetric 36%
  inconclusive rate, gave worst-case FPR and FNR of similar size (about 39% and 40%); the real
  study is asymmetric (about 52% and 26%).
- **The defaults have been updated accordingly.** `R/simulate_dgp.R` now holds values based on M1:
  inconclusive baselines of 50% (different-source) and 17% (same-source), examiner
  inconclusive-tendency SD 1.3, examiner skill SD 0.5, item SD 0.8, and baseline FPR / FNR of
  0.6% / 1.1%. Because item and skill SD are not identified by this study (issue 3), they were
  fixed at values inside their profile-likelihood ranges and the identified parameters were re-estimated
  conditional on them (log-likelihood within 1 unit of the unconstrained M1 fit). With those defaults, the shared
  generator reproduces all six cells of the real crosstab to within about a point and the examiner-level SD
  of inconclusive rates (0.24 real vs. 0.24 simulated).
- **Suggested wording for the calibration slide:** "Fit by maximum likelihood to the
  Ames II cartridge-case responses. Matches the response distribution and examiner-to-examiner
  spread; does not capture firearm-level differences" — with the firearm-level plot as a backup
  slide. That is honest about both the fit and its limits.

## Caveats

- Goodness-of-fit simulations use plug-in parameter values and ignore parameter
  uncertainty; for features off by several multiples that does not change the conclusion,
  but it would matter for borderline ones (the error–inconclusive correlation, some Beretta-type rows).
- The likelihood treats items as independent given the examiner (the DGP's assumption) even
  though items share firearms. That is the very dependence issue 5 documents, so fitted SDs
  partly absorb firearm-level variation and should not be read as pure item or examiner effects.
- SDs are bounded at $[0.02, 3]$. Point estimates at a bound (M0 item SD, M1/M2 item SD) come with
  no meaningful standard error; profile likelihoods are used instead of Wald intervals.
- One study, one evidence type (cartridge cases; the workbook's `Bullets` sheet is untouched).
  Ames II examiners were not sampled from the population of casework examiners, so fitted rates describe
  this sample of examiners, not that population (Flaw B, again).
- 62 unsuitable / unanswered responses (1.4%) were dropped, treated as ignorable.

## Next steps

- ~~Extend `simulate_response()` with truth-specific inconclusive baselines~~ (done: `inconclusive_rate_same`).
  Still open: consider truth-specific examiner tendencies ($\tau_j$ for same-source vs.
  different-source) to address issue 4.
- Add a firearm random effect to the likelihood-based fit. Items share firearms and firearms
  are crossed with examiners, so the exact likelihood is no longer a small quadrature; a
  simulation-based fit (matching examiner- and firearm-level summaries) is the practical route. The observed
  firearm-level SD of same-source inconclusive rates (0.14) is a direct target for `firearm_difficulty_sd`.
- Repeat on the `Bullets` sheet and, if available, on Ames I / Baldwin et al., to see which fitted
  parameters are stable across studies and which are study-specific.

# References

::: {#refs}
:::
