---
title: "Sensitivity Analysis: Non-Representative Sampling Case Study"
subtitle: "How much does the bias magnitude depend on uncalibrated heterogeneity parameters?"
bibliography: ../output/paper/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
    code-fold: true
---

# Goal

`icfis/case_study_nonrepresentative.qmd`'s headline numbers (convenience
sample FPR/FNR roughly 7--8x too low) depend on three uncalibrated
heterogeneity parameters: `type_difficulty_sd` (between-model, currently
1.2), `firearm_difficulty_sd` (within-model, currently 0.9), and
`examiner_skill_sd` (currently 0.5, shared with the rest of the talk; the
Ames II fit cannot identify it).
None of these are fit to real data — see that case study's own "Next
steps."

This grid matches that case study's current design: firearm counts held
equal (80 total) across the convenience and full-coverage scenarios rather
than firearms-per-type held equal, a 4,000-firearm world population per
type, and an incomplete design (each examiner sees only a subset of the
study's items, matching case study 1's mechanism).

This document checks how much of the three-beat story is **structural**
(should hold for any reasonable parameter choice) versus **magnitude**
(specific to the chosen values), by re-running the same three-beat
comparison across a small grid of parameter settings.

```{r setup}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)

source("R/simulate_dgp.R")
source("R/simulate_dgp_types.R")
```

# Design matrix

Two factors, three levels each:

- **Difficulty heterogeneity** (Low / Medium / High): scales both
  `type_difficulty_sd` and `firearm_difficulty_sd` together by 0.5x / 1x
  (the case study's current values) / 1.5x. Medium approximately reproduces the
  case study's numbers (a smaller reference simulation and 200 rather than 500
  replicates, so not exactly).
- **Examiner heterogeneity** (Low / Medium / High): scales
  `examiner_skill_sd` by 0.5x / 1x (the talk's canonical default) / 2x.

```{r design-constants}
# Matches icfis/case_study_nonrepresentative.qmd's current design.
world_firearms_per_type <- 4000
total_firearms <- 80
n_firearms_convenience <- total_firearms / 2   # 2 easiest types
n_firearms_full_coverage <- total_firearms / 8 # all 8 types
items_per_examiner <- 30

# Replicates per cell for the grid below. Raised from an earlier 50 (enough
# for stable means, per the main case study's own Monte Carlo check) because
# percentile intervals need more replicates per cell to not be noisy at the
# extremes.
n_replications_grid <- 200
```

```{r design-matrix}
design_matrix <- crossing(
  difficulty_scenario = c("Low", "Medium", "High"),
  examiner_scenario = c("Low", "Medium", "High")
) %>%
  mutate(
    difficulty_multiplier = case_when(
      difficulty_scenario == "Low" ~ 0.5,
      difficulty_scenario == "Medium" ~ 1,
      difficulty_scenario == "High" ~ 1.5
    ),
    type_difficulty_sd = difficulty_multiplier * 1.2,
    firearm_difficulty_sd = difficulty_multiplier * 0.9,
    examiner_multiplier = case_when(
      examiner_scenario == "Low" ~ 0.5,
      examiner_scenario == "Medium" ~ 1,
      examiner_scenario == "High" ~ 2
    ),
    examiner_skill_sd = examiner_multiplier * dgp_defaults$examiner_skill_sd,
    difficulty_scenario = factor(difficulty_scenario, levels = c("Low", "Medium", "High")),
    examiner_scenario = factor(examiner_scenario, levels = c("Low", "Medium", "High"))
  )

knitr::kable(
  design_matrix %>%
    select(difficulty_scenario, examiner_scenario, type_difficulty_sd, firearm_difficulty_sd, examiner_skill_sd) %>%
    mutate(across(c(type_difficulty_sd, firearm_difficulty_sd, examiner_skill_sd), ~ round(.x, 2))),
  col.names = c("Difficulty", "Examiner", "type_difficulty_sd", "firearm_difficulty_sd", "examiner_skill_sd")
)
```

Scaling `type_difficulty_sd` preserves *which* two types are easiest
(confirmed separately: the relative ordering of the eight types'
difficulties doesn't change under rescaling, since it's the same fixed
random draw just multiplied) — so the convenience sample is always
restricted to the same two types across every cell of this grid, only
their difficulty *magnitude* changes.

# Running the grid

For each of the 9 cells, this builds a fresh type/firearm population at
that cell's heterogeneity settings, computes the true population-average
rate, and runs the same design as the main case study — equal 80 total
firearms per scenario, an incomplete design (`r items_per_examiner`
items/examiner), and a `r comma(world_firearms_per_type)`-firearm world
population per type. Each cell now runs **five** scenarios (up from three):
the original always-easiest-two convenience sample (naive and
"reweighted") and full-coverage (reweighted), plus two new ones matching
`case_study_nonrepresentative.qmd`'s "Robustness check" section --
**a convenience-sized (2-type) sample where the two included types are
drawn uniformly at random each replicate**, naive and "reweighted".

Bumped from 50 to `r n_replications_grid` replicate studies per cell:
50 was enough to see the *mean* pattern (as the main case study's own
Monte Carlo check confirms), but percentile intervals need more replicates
per cell to be anything but noisy at the extremes.

```{r run-grid-functions}
run_cell <- function(type_difficulty_sd, firearm_difficulty_sd, examiner_skill_sd,
                     n_replications, seed_base) {
  type_population <- make_type_population(type_difficulty_sd = type_difficulty_sd)
  firearm_population <- make_firearm_population(
    type_population, firearms_per_type = world_firearms_per_type, firearm_difficulty_sd = firearm_difficulty_sd
  )

  easiest_types <- type_population %>% arrange(type_difficulty) %>% slice(1:2) %>% pull(type_id)
  conv_weights <- if_else(type_population$type_id %in% easiest_types, 1, 0)
  better_weights <- rep(1, nrow(type_population))

  true_reference_data <- generate_population_reference_by_firearm(
    seed = seed_base,
    type_population = type_population,
    firearm_population = firearm_population,
    sampling_weights = type_population$population_weight,
    n_items_total = 1000,
    n_examiners = dgp_defaults$n_examiners,
    match_rate = dgp_defaults$match_rate,
    false_positive_rate = dgp_defaults$false_positive_rate,
    false_negative_rate = dgp_defaults$false_negative_rate,
    inconclusive_rate = dgp_defaults$inconclusive_rate,
    inconclusive_rate_same = dgp_defaults$inconclusive_rate_same,
    examiner_skill_sd = examiner_skill_sd,
    examiner_inconclusive_sd = dgp_defaults$examiner_inconclusive_sd,
    question_sd = dgp_defaults$question_sd
  )
  true_rates <- compute_pooled_rates(true_reference_data)

  seeds <- seed_base + seq_len(n_replications)

  simulate_one <- function(seed, sampling_weights, n_firearms_per_type, items_per_firearm) {
    generate_sim_data_by_type_and_firearm_incomplete(
      seed = seed,
      type_population = type_population,
      firearm_population = firearm_population,
      sampling_weights = sampling_weights,
      n_firearms_per_type = n_firearms_per_type,
      items_per_firearm = items_per_firearm,
      items_per_examiner = items_per_examiner,
      n_examiners = dgp_defaults$n_examiners,
      match_rate = dgp_defaults$match_rate,
      false_positive_rate = dgp_defaults$false_positive_rate,
      false_negative_rate = dgp_defaults$false_negative_rate,
      inconclusive_rate = dgp_defaults$inconclusive_rate,
      inconclusive_rate_same = dgp_defaults$inconclusive_rate_same,
      examiner_skill_sd = examiner_skill_sd,
      examiner_inconclusive_sd = dgp_defaults$examiner_inconclusive_sd,
      question_sd = dgp_defaults$question_sd
    )
  }

  # Full coverage doesn't depend on which 2 types a convenience sample would
  # pick, so it's computed once per cell and shared by both convenience
  # designs below (rather than duplicated, as an earlier version did).
  full_coverage_results <- map_dfr(seeds, function(s) {
    reweight_by_population(compute_type_level_rates(simulate_one(s, better_weights, n_firearms_full_coverage, 4)), type_population) %>%
      rename(rate = reweighted_rate) %>%
      mutate(scenario = "Full-coverage study (reweighted)")
  })

  # Always the two easiest types (the main case study's design).
  easiest_naive <- map_dfr(seeds, function(s) {
    compute_pooled_rates(simulate_one(s, conv_weights, n_firearms_convenience, 10)) %>%
      mutate(scenario = "Convenience sample (naive)")
  })
  easiest_reweighted <- map_dfr(seeds, function(s) {
    reweight_by_population(compute_type_level_rates(simulate_one(s, conv_weights, n_firearms_convenience, 10)), type_population) %>%
      rename(rate = reweighted_rate) %>%
      mutate(scenario = "Convenience sample (\"reweighted\")")
  })

  # A different random pair of the two included types each replicate,
  # independent of difficulty -- see case_study_nonrepresentative.qmd's
  # "Robustness check" section. A separate seed stream (offset by 1e7) so it
  # never overlaps `seeds`, which still governs everything about a given
  # replicate's examiners/items/responses.
  pair_seeds <- (seed_base + 1e7) + seq_len(n_replications)
  random_pairs <- map(pair_seeds, function(s) {
    set.seed(s)
    sample(type_population$type_id, 2)
  })
  random_pair_weights <- map(random_pairs, function(t) if_else(type_population$type_id %in% t, 1, 0))

  random_pair_results <- map_dfr(seq_len(n_replications), function(i) {
    study_data <- simulate_one(seeds[i], random_pair_weights[[i]], n_firearms_convenience, 10)
    naive <- compute_pooled_rates(study_data) %>% mutate(scenario = "Random pair (naive)")
    reweighted <- reweight_by_population(compute_type_level_rates(study_data), type_population) %>%
      rename(rate = reweighted_rate) %>%
      mutate(scenario = "Random pair (\"reweighted\")")
    bind_rows(naive, reweighted)
  })

  bind_rows(full_coverage_results, easiest_naive, easiest_reweighted, random_pair_results) %>%
    group_by(metric, scenario) %>%
    summarise(
      mean_rate = mean(rate),
      sd_rate = sd(rate),
      q025 = quantile(rate, 0.025),
      q975 = quantile(rate, 0.975),
      .groups = "drop"
    ) %>%
    left_join(true_rates %>% rename(true_rate = rate), by = "metric")
}
```

```{r run-grid, cache=TRUE}
grid_results <- design_matrix %>%
  mutate(
    row_id = row_number(),
    results = pmap(
      list(type_difficulty_sd, firearm_difficulty_sd, examiner_skill_sd, row_id),
      function(td, fd, es, rid) run_cell(td, fd, es, n_replications = n_replications_grid, seed_base = 5000 * rid)
    )
  ) %>%
  select(difficulty_scenario, examiner_scenario, results) %>%
  unnest(results) %>%
  mutate(
    scenario = factor(
      scenario,
      levels = c(
        "Convenience sample (naive)",
        "Full-coverage study (reweighted)",
        "Convenience sample (\"reweighted\")",
        "Random pair (naive)",
        "Random pair (\"reweighted\")"
      )
    ),
    metric = factor(metric, levels = c("FPR", "FNR")),
    ratio_to_truth = mean_rate / true_rate
  )
```

# Results

## Full table

```{r results-table}
knitr::kable(
  grid_results %>%
    mutate(
      mean_rate = percent(mean_rate, accuracy = 0.01),
      q025 = percent(q025, accuracy = 0.01),
      q975 = percent(q975, accuracy = 0.01),
      true_rate = percent(true_rate, accuracy = 0.01),
      ratio_to_truth = sprintf("%.2fx", ratio_to_truth)
    ) %>%
    select(difficulty_scenario, examiner_scenario, scenario, metric, mean_rate, q025, q975, true_rate, ratio_to_truth) %>%
    arrange(metric, scenario, difficulty_scenario, examiner_scenario),
  col.names = c("Difficulty", "Examiner", "Scenario", "Metric", "Mean rate", "2.5th pctile", "97.5th pctile", "True rate", "Ratio to truth")
)
```

## Does the bias ratio depend on difficulty heterogeneity or examiner heterogeneity?

```{r heatmap-plot, fig.width=10, fig.height=6}
convenience_ratios <- grid_results %>%
  filter(scenario == "Convenience sample (naive)")

ggplot(convenience_ratios, aes(x = examiner_scenario, y = difficulty_scenario, fill = ratio_to_truth)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1fx", ratio_to_truth)), color = "white", size = 5) +
  facet_wrap(~ metric) +
  scale_fill_viridis_c(name = "Naive rate /\ntrue rate", option = "magma", direction = -1) +
  labs(
    title = "How much does the convenience sample's bias ratio move across the grid?",
    subtitle = sprintf("Each cell: mean naive rate / true population rate, %d replicate studies", n_replications_grid),
    x = "Examiner heterogeneity",
    y = "Difficulty heterogeneity (type + firearm, scaled together)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

## Does reweighting still work (beat 2) and still fail (beat 3) across the whole grid?

```{r reweighting-robustness-plot, fig.width=10, fig.height=6}
# Keep this to the original two lines (full-coverage vs. always-easiest-two
# reweighted) -- the new random-pair scenarios get their own dedicated plot
# below rather than silently showing up here too.
reweighting_check <- grid_results %>%
  filter(scenario %in% c("Full-coverage study (reweighted)", "Convenience sample (\"reweighted\")"))

ggplot(reweighting_check, aes(x = examiner_scenario, y = ratio_to_truth, color = scenario, group = scenario)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  geom_line() +
  geom_point(size = 2) +
  facet_grid(metric ~ difficulty_scenario, labeller = labeller(difficulty_scenario = function(x) paste("Difficulty:", x))) +
  scale_y_continuous(labels = label_number(suffix = "x")) +
  labs(
    title = "Reweighted estimate / true rate, across the grid",
    subtitle = "Dashed line at 1x = perfect recovery. Full-coverage (reweighted) should stay near 1x everywhere; convenience (\"reweighted\") should not.",
    x = "Examiner heterogeneity",
    y = "Ratio to true rate",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Bias vs. variance, with 95% intervals: the biased sample vs. a random-but-convenience sample

The plots above show only means. These two add a **95% interval** (the
2.5th-97.5th percentile of `rate - true_rate` across the
`r n_replications_grid` replicate studies in each cell) around each point
-- one for the always-easiest-two convenience sample ("the biased sample"),
one for the random-pair design ("the random but convenience sample"),
both against the same full-coverage reference.

```{r bias-interval-colors, include=FALSE}
# One shared color per scenario label, used by both plots below, so (e.g.)
# "Full-coverage study (reweighted)" is the same color in both -- ggplot's
# default discrete palette would otherwise reassign colors by whichever
# levels survive each plot's own filter, making green mean different things
# in the two plots.
scenario_colors <- c(
  "Convenience sample (naive)"          = "#D55E00",
  "Convenience sample (\"reweighted\")" = "#0072B2",
  "Full-coverage study (reweighted)"    = "#009E73",
  "Random pair (naive)"                 = "#CC79A7",
  "Random pair (\"reweighted\")"        = "#E69F00"
)
```

```{r bias-interval-plot-easiest, fig.width=10, fig.height=7}
easiest_plot_data <- grid_results %>%
  filter(scenario %in% c(
    "Convenience sample (naive)",
    "Full-coverage study (reweighted)",
    "Convenience sample (\"reweighted\")"
  )) %>%
  mutate(diff = mean_rate - true_rate, diff_lo = q025 - true_rate, diff_hi = q975 - true_rate)

ggplot(easiest_plot_data, aes(x = examiner_scenario, y = diff, color = scenario)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = diff_lo, ymax = diff_hi), position = position_dodge(width = 0.5), size = 0.4) +
  facet_grid(metric ~ difficulty_scenario) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = scenario_colors) +
  labs(
    title = "The biased sample: always the two easiest types",
    subtitle = sprintf("Point: mean rate - true rate, %d replicate studies. Bars: 95%% interval of individual studies", n_replications_grid),
    x = "Examiner heterogeneity", y = "Mean rate - true rate", color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r bias-interval-plot-random, fig.width=10, fig.height=7}
random_plot_data <- grid_results %>%
  filter(scenario %in% c(
    "Random pair (naive)",
    "Full-coverage study (reweighted)",
    "Random pair (\"reweighted\")"
  )) %>%
  mutate(diff = mean_rate - true_rate, diff_lo = q025 - true_rate, diff_hi = q975 - true_rate)

ggplot(random_plot_data, aes(x = examiner_scenario, y = diff, color = scenario)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = diff_lo, ymax = diff_hi), position = position_dodge(width = 0.5), size = 0.4) +
  facet_grid(metric ~ difficulty_scenario) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = scenario_colors) +
  labs(
    title = "The random-but-convenience sample: 2 random types, not always the easiest",
    subtitle = sprintf("Point: mean rate - true rate, %d replicate studies. Wider bars than above, centered nearer zero", n_replications_grid),
    x = "Examiner heterogeneity", y = "Mean rate - true rate", color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The contrast between the two plots is the point this pair of figures is
for: the biased-sample points sit consistently below the dashed
zero line with narrow bars (**precisely wrong**), while the random-pair
points scatter around zero with bars several times wider (**imprecisely
close to right**) -- across every cell of the grid, not just the one
setting the main case study and `case_study_nonrepresentative.qmd`'s
"Robustness check" section report.

*(Faceted by both metrics here; for a slide, filter `grid_results` to
`metric == "FPR"` (or `"FNR"`) first to match a single-row layout like the
one sketched for this request.)*

# Interpretation

```{r interp-helpers, include=FALSE}
naive <- grid_results %>% filter(scenario == "Convenience sample (naive)")
naive_range <- function(m, lvl) { r <- naive %>% filter(metric == m, difficulty_scenario == lvl) %>% pull(ratio_to_truth); c(min(r), max(r)) }
fmt_ratio <- function(r) sprintf("%.2f--%.2fx", r[1], r[2])
fmt_too_low <- function(r) { lo <- round(1 / r[2]); hi <- round(1 / r[1]); if (lo == hi) sprintf("~%.0fx too low", lo) else sprintf("~%.0f--%.0fx too low", lo, hi) }

med_row <- naive %>% filter(metric == "FPR", difficulty_scenario == "Medium")
ex_effect_txt <- paste(sprintf("%.2fx", med_row$ratio_to_truth[order(med_row$examiner_scenario)]), collapse = " -> ")

full_rng <- grid_results %>% filter(scenario == "Full-coverage study (reweighted)") %>% pull(ratio_to_truth) %>% range()
fix_rng <- grid_results %>% filter(scenario == "Convenience sample (\"reweighted\")") %>% pull(ratio_to_truth) %>% range()
all_naive <- range(naive$ratio_to_truth)

# The main case study's headline figure is this grid's Medium-difficulty,
# Medium-examiner cell; computed once here and reused below and in
# "How to use this in the talk" so it can't drift out of sync with a re-run.
medium_medium <- naive %>% filter(metric == "FPR", difficulty_scenario == "Medium", examiner_scenario == "Medium")
headline_x <- round(1 / medium_medium$ratio_to_truth)
```

- **The bias ratio scales strongly and monotonically with difficulty
  heterogeneity.** For FPR, the convenience sample's naive rate is roughly
  `r fmt_ratio(naive_range("FPR", "Low"))` the true rate at Low difficulty heterogeneity
  (`r fmt_too_low(naive_range("FPR", "Low"))`), `r fmt_ratio(naive_range("FPR", "Medium"))` at Medium
  (`r fmt_too_low(naive_range("FPR", "Medium"))` — this is where the main case study's "`r headline_x`x" comes from), and
  `r fmt_ratio(naive_range("FPR", "High"))` at High (`r fmt_too_low(naive_range("FPR", "High"))`). FNR
  shows the same pattern (Low `r fmt_ratio(naive_range("FNR", "Low"))`, Medium `r fmt_ratio(naive_range("FNR", "Medium"))`,
  High `r fmt_ratio(naive_range("FNR", "High"))`). More between-model difficulty spread makes the
  two "easiest" types more extreme relative to the population average, so
  the bias gets *worse*, not better, as this parameter grows.
- **Examiner (skill) heterogeneity has no consistent effect.** Within each difficulty row the
  ratio moves by much less than it does across rows and not always in the same direction
  (e.g., FPR at Medium difficulty: `r ex_effect_txt` going Low -> Medium -> High examiner heterogeneity),
  so treat it as small next to what difficulty heterogeneity produces on its own. (Only examiner
  *skill* is varied here; the examiner inconclusive-tendency SD, which the Ames II data do identify,
  is held at the talk's default.)
- **The qualitative story holds at every one of the 9 grid points
  tested.** Full-coverage (reweighted) stays within
  `r fmt_ratio(full_rng)` of the true rate everywhere (consistent with sampling noise at only 50
  replicates per cell, not a systematic miss). Convenience ("reweighted")
  stays within `r fmt_ratio(range(fix_rng))` — badly biased everywhere, never approaching
  1x. Reweighting recovering the truth with coverage, and failing without
  it, does not depend on the specific heterogeneity magnitudes used in
  the main case study.
- **The specific "`r headline_x`x" figure in the main case study is one
  point in this grid (Medium difficulty, Medium examiner)** — real,
  reproducible from the same code, but not privileged over any other
  cell; the true bias could plausibly be reported as anywhere from about
  `r sprintf("%.0f", 1 / max(all_naive))`x to `r sprintf("%.0f", 1 / min(all_naive))`x too low depending on how much between-model
  heterogeneity actually exists in real casework, which nothing here is calibrated to (see the
  main case study's "Next steps"). Only the *direction* and the
  reweighting-works/fails distinction should be treated as a claim; the
  specific multiplier should not be quoted as an empirical estimate about
  real firearms examination.

# How to use this in the talk

If asked "isn't your `r headline_x`x just an artifact of your parameter choices?" —
yes, the *number* is; the *direction and mechanism* aren't. This document
is the receipt: the bias ratio moves with difficulty heterogeneity as
expected, is roughly flat across examiner heterogeneity, and reweighting's
success/failure pattern (recovers truth with coverage, doesn't without)
holds at every one of the 9 settings tested here, not just the one
reported in the main case study.

# References

::: {#refs}
:::
