---
title: "Sensitivity Analysis: Non-Representative Sampling Case Study"
subtitle: "How much does the bias magnitude depend on uncalibrated heterogeneity parameters?"
bibliography: ../output/paper/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
    code-fold: true
---

# Goal

`icfis/case_study_nonrepresentative.qmd`'s headline numbers (convenience
sample FPR/FNR roughly 5x and 4.5x too low) depend on three uncalibrated
heterogeneity parameters: `type_difficulty_sd` (between-model, currently
1.2), `firearm_difficulty_sd` (within-model, currently 0.9), and
`examiner_skill_sd` (currently 0.7, shared with the rest of the talk).
None of these are fit to real data — see that case study's own "Next
steps."

This document checks how much of the three-beat story is **structural**
(should hold for any reasonable parameter choice) versus **magnitude**
(specific to the chosen values), by re-running the same three-beat
comparison across a small grid of parameter settings.

```{r setup}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)

source("R/simulate_dgp.R")
source("R/simulate_dgp_types.R")
```

# Design matrix

Two factors, three levels each:

- **Difficulty heterogeneity** (Low / Medium / High): scales both
  `type_difficulty_sd` and `firearm_difficulty_sd` together by 0.5x / 1x
  (the case study's current values) / 1.5x. Medium reproduces the case
  study's numbers exactly.
- **Examiner heterogeneity** (Low / Medium / High): scales
  `examiner_skill_sd` by 0.5x / 1x (the talk's canonical default) / 2x.

```{r design-matrix}
design_matrix <- crossing(
  difficulty_scenario = c("Low", "Medium", "High"),
  examiner_scenario = c("Low", "Medium", "High")
) %>%
  mutate(
    difficulty_multiplier = case_when(
      difficulty_scenario == "Low" ~ 0.5,
      difficulty_scenario == "Medium" ~ 1,
      difficulty_scenario == "High" ~ 1.5
    ),
    type_difficulty_sd = difficulty_multiplier * 1.2,
    firearm_difficulty_sd = difficulty_multiplier * 0.9,
    examiner_multiplier = case_when(
      examiner_scenario == "Low" ~ 0.5,
      examiner_scenario == "Medium" ~ 1,
      examiner_scenario == "High" ~ 2
    ),
    examiner_skill_sd = examiner_multiplier * dgp_defaults$examiner_skill_sd,
    difficulty_scenario = factor(difficulty_scenario, levels = c("Low", "Medium", "High")),
    examiner_scenario = factor(examiner_scenario, levels = c("Low", "Medium", "High"))
  )

knitr::kable(
  design_matrix %>%
    select(difficulty_scenario, examiner_scenario, type_difficulty_sd, firearm_difficulty_sd, examiner_skill_sd) %>%
    mutate(across(c(type_difficulty_sd, firearm_difficulty_sd, examiner_skill_sd), ~ round(.x, 2))),
  col.names = c("Difficulty", "Examiner", "type_difficulty_sd", "firearm_difficulty_sd", "examiner_skill_sd")
)
```

Scaling `type_difficulty_sd` preserves *which* two types are easiest
(confirmed separately: the relative ordering of the eight types'
difficulties doesn't change under rescaling, since it's the same fixed
random draw just multiplied) — so the convenience sample is always
restricted to the same two types across every cell of this grid, only
their difficulty *magnitude* changes.

# Running the grid

For each of the 9 cells, this builds a fresh type/firearm population at
that cell's heterogeneity settings, computes the true population-average
rate, and runs the same three-beat comparison as the main case study — at
50 replicate studies per beat (this is a sensitivity survey, not a final
number; the Monte Carlo error section in `case_study_nonrepresentative.qmd`
already showed 100 replicates barely differs from 500 for the mean, so 50
is enough to see the pattern here).

```{r run-grid-functions}
run_cell <- function(type_difficulty_sd, firearm_difficulty_sd, examiner_skill_sd,
                     n_replications = 50, seed_base) {
  type_population <- make_type_population(type_difficulty_sd = type_difficulty_sd)
  firearm_population <- make_firearm_population(type_population, firearm_difficulty_sd = firearm_difficulty_sd)

  easiest_types <- type_population %>% arrange(type_difficulty) %>% slice(1:2) %>% pull(type_id)
  conv_weights <- if_else(type_population$type_id %in% easiest_types, 1, 0)
  better_weights <- rep(1, nrow(type_population))

  true_reference_data <- generate_population_reference_by_firearm(
    seed = seed_base,
    type_population = type_population,
    firearm_population = firearm_population,
    sampling_weights = type_population$population_weight,
    n_items_total = 1000,
    n_examiners = dgp_defaults$n_examiners,
    match_rate = dgp_defaults$match_rate,
    false_positive_rate = dgp_defaults$false_positive_rate,
    false_negative_rate = dgp_defaults$false_negative_rate,
    inconclusive_rate = dgp_defaults$inconclusive_rate,
    examiner_skill_sd = examiner_skill_sd,
    examiner_inconclusive_sd = dgp_defaults$examiner_inconclusive_sd,
    question_sd = dgp_defaults$question_sd
  )
  true_rates <- compute_pooled_rates(true_reference_data)

  seeds <- seed_base + seq_len(n_replications)

  simulate_one <- function(seed, sampling_weights, items_per_firearm) {
    generate_sim_data_by_type_and_firearm(
      seed = seed,
      type_population = type_population,
      firearm_population = firearm_population,
      sampling_weights = sampling_weights,
      n_firearms_per_type = 5,
      items_per_firearm = items_per_firearm,
      n_examiners = dgp_defaults$n_examiners,
      match_rate = dgp_defaults$match_rate,
      false_positive_rate = dgp_defaults$false_positive_rate,
      false_negative_rate = dgp_defaults$false_negative_rate,
      inconclusive_rate = dgp_defaults$inconclusive_rate,
      examiner_skill_sd = examiner_skill_sd,
      examiner_inconclusive_sd = dgp_defaults$examiner_inconclusive_sd,
      question_sd = dgp_defaults$question_sd
    )
  }

  convenience_results <- map_dfr(seeds, function(s) {
    compute_pooled_rates(simulate_one(s, conv_weights, 10)) %>%
      mutate(scenario = "Convenience sample (naive)")
  })
  better_results <- map_dfr(seeds, function(s) {
    reweight_by_population(compute_type_level_rates(simulate_one(s, better_weights, 4)), type_population) %>%
      rename(rate = reweighted_rate) %>%
      mutate(scenario = "Full-coverage study (reweighted)")
  })
  convenience_fixed_results <- map_dfr(seeds, function(s) {
    reweight_by_population(compute_type_level_rates(simulate_one(s, conv_weights, 10)), type_population) %>%
      rename(rate = reweighted_rate) %>%
      mutate(scenario = "Convenience sample (\"reweighted\")")
  })

  bind_rows(convenience_results, better_results, convenience_fixed_results) %>%
    group_by(metric, scenario) %>%
    summarise(mean_rate = mean(rate), .groups = "drop") %>%
    left_join(true_rates %>% rename(true_rate = rate), by = "metric")
}
```

```{r run-grid, cache=TRUE}
grid_results <- design_matrix %>%
  mutate(
    row_id = row_number(),
    results = pmap(
      list(type_difficulty_sd, firearm_difficulty_sd, examiner_skill_sd, row_id),
      function(td, fd, es, rid) run_cell(td, fd, es, n_replications = 50, seed_base = 5000 * rid)
    )
  ) %>%
  select(difficulty_scenario, examiner_scenario, results) %>%
  unnest(results) %>%
  mutate(
    scenario = factor(
      scenario,
      levels = c(
        "Convenience sample (naive)",
        "Full-coverage study (reweighted)",
        "Convenience sample (\"reweighted\")"
      )
    ),
    metric = factor(metric, levels = c("FPR", "FNR")),
    ratio_to_truth = mean_rate / true_rate
  )
```

# Results

## Full table

```{r results-table}
knitr::kable(
  grid_results %>%
    mutate(
      mean_rate = percent(mean_rate, accuracy = 0.01),
      true_rate = percent(true_rate, accuracy = 0.01),
      ratio_to_truth = sprintf("%.2fx", ratio_to_truth)
    ) %>%
    arrange(metric, scenario, difficulty_scenario, examiner_scenario),
  col.names = c("Difficulty", "Examiner", "Scenario", "Metric", "Mean rate", "True rate", "Ratio to truth")
)
```

## Does the bias ratio depend on difficulty heterogeneity or examiner heterogeneity?

```{r heatmap-plot, fig.width=10, fig.height=6}
convenience_ratios <- grid_results %>%
  filter(scenario == "Convenience sample (naive)")

ggplot(convenience_ratios, aes(x = examiner_scenario, y = difficulty_scenario, fill = ratio_to_truth)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1fx", ratio_to_truth)), color = "white", size = 5) +
  facet_wrap(~ metric) +
  scale_fill_viridis_c(name = "Naive rate /\ntrue rate", option = "magma", direction = -1) +
  labs(
    title = "How much does the convenience sample's bias ratio move across the grid?",
    subtitle = "Each cell: mean naive rate / true population rate, 50 replicate studies",
    x = "Examiner heterogeneity",
    y = "Difficulty heterogeneity (type + firearm, scaled together)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

## Does reweighting still work (beat 2) and still fail (beat 3) across the whole grid?

```{r reweighting-robustness-plot, fig.width=10, fig.height=6}
reweighting_check <- grid_results %>%
  filter(scenario != "Convenience sample (naive)")

ggplot(reweighting_check, aes(x = examiner_scenario, y = ratio_to_truth, color = scenario, group = scenario)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  geom_line() +
  geom_point(size = 2) +
  facet_grid(metric ~ difficulty_scenario, labeller = labeller(difficulty_scenario = function(x) paste("Difficulty:", x))) +
  scale_y_continuous(labels = label_number(suffix = "x")) +
  labs(
    title = "Reweighted estimate / true rate, across the grid",
    subtitle = "Dashed line at 1x = perfect recovery. Full-coverage (reweighted) should stay near 1x everywhere; convenience (\"reweighted\") should not.",
    x = "Examiner heterogeneity",
    y = "Ratio to true rate",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Interpretation

- **The bias ratio scales strongly and monotonically with difficulty
  heterogeneity.** For FPR, the convenience sample's naive rate is roughly
  0.4–0.56x the true rate at Low difficulty heterogeneity (~2–2.5x too
  low), 0.20–0.24x at Medium (~4–5x too low — this is where the main case
  study's "5x" comes from), and 0.09–0.11x at High (~9–11x too low). FNR
  shows the same pattern. More between-model difficulty spread makes the
  two "easiest" types more extreme relative to the population average, so
  the bias gets *worse*, not better, as this parameter grows.
- **Examiner heterogeneity has a real but much smaller effect.** Within
  each difficulty row, moving from Low to High examiner heterogeneity
  nudges the ratio up modestly (e.g., FPR at Medium difficulty: 0.20x →
  0.21x → 0.24x going Low → Medium → High examiner) — a mild dilution
  effect, not the near-total insensitivity originally hypothesized, but
  small next to the ~5x range difficulty heterogeneity produces on its
  own.
- **The qualitative story holds at every one of the 9 grid points
  tested.** Full-coverage (reweighted) stays within 0.85x–1.31x of the
  true rate everywhere (consistent with sampling noise at only 50
  replicates per cell, not a systematic miss). Convenience ("reweighted")
  stays within 0.10x–0.62x — badly biased everywhere, never approaching
  1x. Reweighting recovering the truth with coverage, and failing without
  it, does not depend on the specific heterogeneity magnitudes used in
  the main case study.
- **The specific "5x" and "4.5x" figures in the main case study are one
  point in this grid (Medium difficulty, Medium examiner)** — real,
  reproducible from the same code, but not privileged over any other
  cell; the true bias could plausibly be reported as anywhere from ~2x to
  ~11x too low depending on how much between-model heterogeneity actually
  exists in real casework, which nothing here is calibrated to (see the
  main case study's "Next steps"). Only the *direction* and the
  reweighting-works/fails distinction should be treated as a claim; the
  specific multiplier should not be quoted as an empirical estimate about
  real firearms examination.

# How to use this in the talk

If asked "isn't your 5x just an artifact of your parameter choices?" —
yes, the *number* is; the *direction and mechanism* aren't. This document
is the receipt: the bias ratio moves with difficulty heterogeneity as
expected, is roughly flat across examiner heterogeneity, and reweighting's
success/failure pattern (recovers truth with coverage, doesn't without)
holds at every one of the 9 settings tested here, not just the one
reported in the main case study.

# References

::: {#refs}
:::
