---
title: "Sensitivity Analysis: Inconclusive-Handling Case Study"
subtitle: "Can a high inconclusive rate hide a high true error rate -- and does that depend on how we model inconclusives, examiners, and items?"
bibliography: ../output/paper/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
    code-fold: true
---

# Goal

`icfis/case_study_inconclusive.qmd` shows that the same raw responses give
very different "error rates" depending on how inconclusives are handled
(reported vs. worst case). That case study uses one set of DGP defaults: a
single inconclusive baseline shared by same- and different-source items and
fixed levels of examiner and item heterogeneity. This document asks how much
of the story depends on those choices, by re-running it over a grid:

- **(a) Inconclusive baselines:** one rate shared by same- and different-source
  comparisons (the talk's *previous* default, 36%) vs. separate rates for each (the talk's current default, from what the
  Ames II data prefer; see `icfis/fit_ames2.qmd`).
- **(b) Examiner heterogeneity:** low / medium / high.
- **(c) Item heterogeneity:** low / medium / high.
- **(d) Testing condition:** blind (casework-like) vs. a known test, in which
  examiners' inconclusive probability is shifted up by the Hawthorne effect
  estimated from Scurich et al. (2025).

The claim to check is the one the talk wants to make: **a study's reported error
rate can be very low while the true error rate is high, with the difference
hidden inside a large inconclusive rate — and a known-test study's
inconclusive rate need not resemble casework's at all.**

```{r setup}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(scales)
library(tibble)

source(here::here("icfis/R/simulate_dgp.R"))
```

# What "true error rate" means here

In the shared DGP, an examiner's response is an error, an inconclusive, or a
correct call, and the study's reported error rate
(identifications / all different-source comparisons, and the same-source
analog) is exactly the probability of the "error" response. Nothing in that
model says what an inconclusive *would have been* had the examiner been
required to decide, so with the model alone the reported rate cannot be shown
to be wrong. It is only bounded: recoding every inconclusive as an error gives
the worst case.

To ask "what if the inconclusives are hiding errors?" we add one scenario
parameter on top of the DGP, **without changing any of the simulated
responses**: $h$, the share of inconclusive responses that would have been
errors had the examiner been forced to make a call. Then, for either rate,

$$
\text{true (forced-decision) rate}(h) \;=\; \text{reported} \;+\; h \times \text{inconclusive share},
$$

with $h = 0$ reproducing the reported rate and $h = 1$ the worst case. **$h$
cannot be estimated from a study's responses** — that is the point — so results
are shown for $h \in \{0.1, 0.25, 0.5\}$ alongside the two extremes. This is a
scenario, not a calibrated quantity; nothing in the Ames II data pins it down.

# The grid

```{r grid-definition}
# (a) Inconclusive baselines. The separate values are the talk's current defaults,
# from the M1 fit to the Ames II cartridge-case data (icfis/fit_ames2.qmd): about
# 50% for different-source and 17% for same-source. The shared value is the
# talk's previous default (icfis/analysis.qmd).
ir_structures <- tribble(
  ~ir_structure,            ~ir_diff, ~ir_same,
  "Shared (previous default)",   0.36,     0.36,
  "Separate (current default)",  0.50,     0.17
)

# (b) Examiner heterogeneity: skill SD and inconclusive-tendency SD move together.
# "High" is the talk's current default (tendency SD 1.3 from the Ames II fit, skill
# SD 0.5); "Low" uses the previous default's tendency SD (0.35). Skill SD is not
# identified by Ames II, so these values are only illustrative.
examiner_levels <- tribble(
  ~examiner_level, ~sd_skill, ~sd_incon,
  "Low",           0.25,      0.35,
  "Medium",        0.35,      0.80,
  "High",          0.50,      1.30
)

# (c) Item heterogeneity: 0.8 is the talk's default (current and previous); Ames II cannot distinguish
# values from ~0 to about 1.2 (see fit_ames2.qmd's profile likelihoods).
item_levels <- tribble(
  ~item_level, ~sd_item,
  "Low",       0.4,
  "Medium",    0.8,
  "High",      1.2
)

# (d) Known-test (Hawthorne) shift on the inconclusive logit, from Scurich et al.
# (2025): 39.3% inconclusive on blind vs. 56.4% on known-test items.
hawthorne_shift <- qlogis(0.564) - qlogis(0.393)

conditions <- tribble(
  ~condition,     ~shift,
  "Blind",        0,
  "Known test",   hawthorne_shift
)

# Held fixed: the talk's current error baselines and the incomplete design (50 examiners,
# 30 of 300 items each) used throughout the case studies.
fixed <- list(
  fpr = dgp_defaults$false_positive_rate,
  fnr = dgp_defaults$false_negative_rate,
  match_rate = incomplete_defaults$match_rate,
  n_examiners = incomplete_defaults$n_examiners,
  n_comparisons = incomplete_defaults$n_comparisons,
  items_per_examiner = incomplete_defaults$items_per_examiner
)

h_values <- c(0.1, 0.25, 0.5)
n_reps <- 500

design_matrix <- crossing(
  examiner_level = examiner_levels$examiner_level,
  item_level = item_levels$item_level
) |>
  left_join(examiner_levels, by = "examiner_level") |>
  left_join(item_levels, by = "item_level") |>
  mutate(
    examiner_level = factor(examiner_level, levels = c("Low", "Medium", "High")),
    item_level = factor(item_level, levels = c("Low", "Medium", "High"))
  )

knitr::kable(
  bind_rows(
    ir_structures |> transmute(Factor = "(a) Inconclusive baselines", Level = ir_structure,
                               Values = sprintf("different-source %s, same-source %s", percent(ir_diff), percent(ir_same))),
    examiner_levels |> transmute(Factor = "(b) Examiner heterogeneity", Level = examiner_level,
                                 Values = sprintf("skill SD %.2f, inconclusive-tendency SD %.2f", sd_skill, sd_incon)),
    item_levels |> transmute(Factor = "(c) Item heterogeneity", Level = item_level,
                             Values = sprintf("item SD %.1f", sd_item)),
    conditions |> transmute(Factor = "(d) Testing condition", Level = condition,
                            Values = sprintf("inconclusive logit shift %+.2f", shift))
  )
)
```

The grid is 2 x 3 x 3 x 2 = 36 combinations, each simulated
`r n_reps` times. Within a replicate, all four combinations of (a) and (d)
reuse the same random draws (same examiners, items, ground truth, and
uniform draws for each response), so differences between them are due to the
factor, not noise.

# Simulation

The shared `generate_sim_data_incomplete()` supports separate baselines via its
`inconclusive_rate_same` argument, but simulates one response at a time. With
36 grid cells x 500 replicates x 4 (a) x (d) combinations, this document uses
a vectorized re-implementation of the same response model instead (which also
lets the four combinations share their random draws). The response
probabilities are the same as `simulate_response()`'s, including the 0.95 cap;
the next section checks that the two implementations agree for both levels of (a).

```{r simulator}
# Mirrors simulate_response() in R/simulate_dgp.R, vectorized. `a_err` and
# `a_inc` are baseline logits; `challenge` and `tendency` are per response.
response_probs <- function(challenge, tendency, a_err, a_inc) {
  err <- plogis(a_err + challenge)
  inc <- plogis(a_inc + 0.5 * challenge + tendency)
  total <- err + inc
  scale <- ifelse(total >= 0.95, 0.95 / total, 1)
  list(err = err * scale, inc = inc * scale)
}

# All random draws for one simulated study (incomplete design). Drawn once per
# seed and reused across the (a) x (d) combinations.
draw_latents <- function(seed, sd_skill, sd_incon, sd_item, fixed) {
  set.seed(seed)
  items <- tibble(
    item = seq_len(fixed$n_comparisons),
    truth = rbinom(fixed$n_comparisons, 1, fixed$match_rate),
    difficulty = rnorm(fixed$n_comparisons, 0, sd_item)
  )
  examiners <- tibble(
    examiner = seq_len(fixed$n_examiners),
    skill = rnorm(fixed$n_examiners, 0, sd_skill),
    tendency = rnorm(fixed$n_examiners, 0, sd_incon)
  )
  assignment <- tibble(
    examiner = rep(examiners$examiner, each = fixed$items_per_examiner),
    item = unlist(replicate(fixed$n_examiners, sample.int(fixed$n_comparisons, fixed$items_per_examiner), simplify = FALSE))
  )
  assignment |>
    left_join(examiners, by = "examiner") |>
    left_join(items, by = "item") |>
    mutate(challenge = difficulty - skill, u = runif(n()))
}

# Response for each row: "err", "inc", or "cor".
respond <- function(lat, ir_diff, ir_same, shift, fixed) {
  a_err <- if_else(lat$truth == 1, safe_logit(fixed$fnr), safe_logit(fixed$fpr))
  a_inc <- if_else(lat$truth == 1, safe_logit(ir_same), safe_logit(ir_diff)) + shift
  p <- response_probs(lat$challenge, lat$tendency, a_err, a_inc)
  case_when(lat$u < p$err ~ "err", lat$u < p$err + p$inc ~ "inc", TRUE ~ "cor")
}

# Reported, worst-case, and inconclusive share for FPR and FNR.
summarise_study <- function(lat, resp) {
  diff_rows <- lat$truth == 0
  same_rows <- lat$truth == 1
  tibble(
    rate = c("FPR", "FNR"),
    reported = c(mean(resp[diff_rows] == "err"), mean(resp[same_rows] == "err")),
    inc_share = c(mean(resp[diff_rows] == "inc"), mean(resp[same_rows] == "inc"))
  ) |>
    mutate(worst_case = reported + inc_share)
}
```

## Check: the vectorized simulator matches the shared generator

For each level of (a), blind, the two implementations should give the same
ground truth x decision proportions up to Monte Carlo error. (The separate-baseline
row also exercises the shared generator's `inconclusive_rate_same` argument.)

```{r check-simulator, cache=TRUE}
check_reps <- 100
canonical <- dgp_defaults

check_one <- function(ir_diff, ir_same) {
  shared_gen <- map_dfr(seq_len(check_reps), function(r) {
    generate_sim_data_incomplete(
      seed = 700 + r,
      n_examiners = incomplete_defaults$n_examiners, n_comparisons = incomplete_defaults$n_comparisons,
      items_per_examiner = incomplete_defaults$items_per_examiner, match_rate = incomplete_defaults$match_rate,
      false_positive_rate = canonical$false_positive_rate, false_negative_rate = canonical$false_negative_rate,
      inconclusive_rate = ir_diff, inconclusive_rate_same = ir_same,
      examiner_skill_sd = canonical$examiner_skill_sd,
      examiner_inconclusive_sd = canonical$examiner_inconclusive_sd, question_sd = canonical$question_sd
    ) |> study_crosstab(replicate_id = r)
  }) |>
    group_by(ground_truth_label, response) |>
    summarise(shared_generator = mean(proportion), .groups = "drop")

  local_gen <- map_dfr(seq_len(check_reps), function(r) {
    lat <- draw_latents(700 + r, canonical$examiner_skill_sd, canonical$examiner_inconclusive_sd, canonical$question_sd, fixed)
    resp <- respond(lat, ir_diff, ir_same, 0, fixed)
    tibble(truth = lat$truth, resp = resp) |>
      count(truth, resp) |>
      group_by(truth) |>
      mutate(proportion = n / sum(n)) |>
      ungroup()
  }) |>
    mutate(
      ground_truth_label = if_else(truth == 1, "same-source", "different-source"),
      response = case_when(
        resp == "inc" ~ "inconclusive",
        resp == "err" & truth == 1 ~ "elimination",
        resp == "err" & truth == 0 ~ "identification",
        resp == "cor" & truth == 1 ~ "identification",
        TRUE ~ "elimination"
      )
    ) |>
    group_by(ground_truth_label, response) |>
    summarise(vectorized_simulator = sum(proportion) / check_reps, .groups = "drop")

  left_join(shared_gen, local_gen, by = c("ground_truth_label", "response"))
}

simulator_check <- ir_structures |>
  mutate(result = map2(ir_diff, ir_same, check_one)) |>
  select(ir_structure, result) |>
  unnest(result)

knitr::kable(
  simulator_check |>
    mutate(across(c(shared_generator, vectorized_simulator), ~ percent(.x, accuracy = 0.1))),
  col.names = c("Baselines", "Ground truth", "Decision", "Shared generator (mean of 100 studies)", "Vectorized simulator (mean of 100 studies)")
)
```

Agreement is within what 100 replicates of 1,500 responses can resolve
(differences of a few tenths of a percentage point on the larger cells), for both
levels of (a).

# Running the grid

```{r run-grid, cache=TRUE, cache.extra=list(n_reps, hawthorne_shift)}
run_cell <- function(cell_id, sd_skill, sd_incon, sd_item) {
  map_dfr(seq_len(n_reps), function(r) {
    lat <- draw_latents(10000 * cell_id + r, sd_skill, sd_incon, sd_item, fixed)

    crossing(ir_structures, conditions) |>
      pmap_dfr(function(ir_structure, ir_diff, ir_same, condition, shift) {
        summarise_study(lat, respond(lat, ir_diff, ir_same, shift, fixed)) |>
          mutate(ir_structure = ir_structure, condition = condition, replicate = r)
      })
  })
}

study_results <- design_matrix |>
  mutate(cell_id = row_number()) |>
  mutate(results = pmap(list(cell_id, sd_skill, sd_incon, sd_item), run_cell)) |>
  select(examiner_level, item_level, results) |>
  unnest(results) |>
  mutate(
    ir_structure = factor(ir_structure, levels = ir_structures$ir_structure),
    condition = factor(condition, levels = conditions$condition),
    rate = factor(rate, levels = c("FPR", "FNR"))
  )

grid_summary <- study_results |>
  group_by(ir_structure, condition, examiner_level, item_level, rate) |>
  summarise(
    # SDs first: later arguments in summarise() see columns created earlier.
    sd_reported = sd(reported), sd_worst_case = sd(worst_case),
    reported = mean(reported), inc_share = mean(inc_share), worst_case = mean(worst_case),
    .groups = "drop"
  ) |>
  mutate(
    truth_h10 = reported + 0.10 * inc_share,
    truth_h25 = reported + 0.25 * inc_share,
    truth_h50 = reported + 0.50 * inc_share
  )
```

# Results

## The headline picture: one plausible cell, four ways

The cell below combines the Ames-II-like settings from the fit (separate
baselines, high examiner heterogeneity) with the talk's default item
heterogeneity. That is exactly the talk's current default parameter set. The
contrast is the previous shared-baseline default. Each row
shows what a study would *report* (filled dot), the *worst case* (open dot),
and where the true forced-decision rate would sit if 10%, 25%, or 50% of
inconclusives were hidden errors (diamonds).

```{r plot-headline, fig.width=10, fig.height=4.5}
headline <- grid_summary |>
  filter(examiner_level == "High", item_level == "Medium") |>
  mutate(row = paste(ir_structure, condition, sep = " | "))

headline_long <- headline |>
  select(row, rate, truth_h10, truth_h25, truth_h50) |>
  pivot_longer(starts_with("truth_"), names_to = "h", values_to = "value") |>
  mutate(h = recode(h, truth_h10 = "10% of inconclusives", truth_h25 = "25%", truth_h50 = "50%"))

ggplot(headline, aes(y = row)) +
  geom_segment(aes(x = reported, xend = worst_case, yend = row), color = "grey70", linewidth = 2) +
  geom_point(data = headline_long, aes(x = value, shape = h), size = 2.5, color = "#D55E00") +
  geom_point(aes(x = worst_case), shape = 1, size = 3) +
  geom_point(aes(x = reported), size = 3) +
  facet_wrap(~ rate) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_shape_manual(values = c("10% of inconclusives" = 5, "25%" = 18, "50%" = 9)) +
  labs(
    title = "Reported vs. worst-case error rate, with plausible true rates in between",
    subtitle = "Filled dot: reported. Open dot: worst case. Diamonds: true rate if a share of inconclusives were errors",
    x = "Error rate", y = NULL, shape = "Share of inconclusives that\nwould have been errors"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Full table

```{r results-table}
knitr::kable(
  grid_summary |>
    arrange(rate, ir_structure, condition, examiner_level, item_level) |>
    mutate(across(c(reported, inc_share, worst_case, truth_h10, truth_h25, truth_h50, sd_reported, sd_worst_case), ~ percent(.x, accuracy = 0.1))) |>
    select(rate, ir_structure, condition, examiner_level, item_level, reported, inc_share, truth_h10, truth_h25, truth_h50, worst_case, sd_reported, sd_worst_case),
  col.names = c("Rate", "Baselines", "Condition", "Examiner het.", "Item het.", "Reported", "Inconclusive share",
                "True (h = 10%)", "True (h = 25%)", "True (h = 50%)", "Worst case", "SD reported", "SD worst case")
)
```

## How far above the reported rate could the truth be?

Each tile is the true rate at $h = 25\%$ divided by the reported rate — how many
times larger the true forced-decision error rate is than what the study
reports if a quarter of its inconclusives were errors.

```{r plot-ratio-heatmap, fig.width=11, fig.height=8}
ggplot(grid_summary, aes(x = examiner_level, y = item_level, fill = truth_h25 / reported)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.0fx", truth_h25 / reported)), color = "white", size = 4) +
  facet_grid(ir_structure + condition ~ rate) +
  scale_fill_viridis_c(name = "True (h = 25%) /\nreported", option = "magma", direction = -1) +
  labs(
    title = "True forced-decision error rate relative to the reported rate",
    subtitle = "If 25% of inconclusives would have been errors. Each cell: mean over simulated studies",
    x = "Examiner heterogeneity", y = "Item heterogeneity"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

## Does the reported rate itself move with heterogeneity?

```{r plot-reported-heatmap, fig.width=11, fig.height=8}
ggplot(grid_summary, aes(x = examiner_level, y = item_level, fill = reported)) +
  geom_tile(color = "white") +
  geom_text(aes(label = percent(reported, accuracy = 0.1)), color = "white", size = 4) +
  facet_grid(ir_structure + condition ~ rate) +
  scale_fill_viridis_c(name = "Reported rate", option = "magma", direction = -1, labels = percent_format(accuracy = 1)) +
  labs(
    title = "What a study would report, across the grid",
    subtitle = "Each cell: mean over simulated studies",
    x = "Examiner heterogeneity", y = "Item heterogeneity"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

## The Hawthorne effect: what a known test changes

Comparing the blind and known-test rows within each cell (paired: same
simulated examiners, items, and uniform draws):

```{r hawthorne-table}
hawthorne <- grid_summary |>
  select(ir_structure, examiner_level, item_level, rate, condition, reported, inc_share, worst_case) |>
  pivot_wider(names_from = condition, values_from = c(reported, inc_share, worst_case)) |>
  mutate(
    d_reported = `reported_Known test` - reported_Blind,
    d_inc_share = `inc_share_Known test` - inc_share_Blind,
    d_worst_case = `worst_case_Known test` - worst_case_Blind
  ) |>
  group_by(ir_structure, rate) |>
  summarise(
    across(c(d_reported, d_inc_share, d_worst_case), list(mean = mean, min = min, max = max)),
    .groups = "drop"
  )

pp <- function(x) sprintf("%+.1f pp", 100 * x)
knitr::kable(
  hawthorne |>
    transmute(
      Baselines = ir_structure, Rate = rate,
      `Change in reported rate` = sprintf("%s (range %s to %s)", pp(d_reported_mean), pp(d_reported_min), pp(d_reported_max)),
      `Change in inconclusive share` = sprintf("%s (range %s to %s)", pp(d_inc_share_mean), pp(d_inc_share_min), pp(d_inc_share_max)),
      `Change in worst case` = sprintf("%s (range %s to %s)", pp(d_worst_case_mean), pp(d_worst_case_min), pp(d_worst_case_max))
    )
)
```

## Study-to-study spread

Under high examiner heterogeneity, a single study's inconclusive share — and
therefore its worst-case bound — depends heavily on which 50 examiners happened
to take part.

```{r plot-spread, fig.width=10, fig.height=5}
spread_data <- study_results |>
  filter(item_level == "Medium", condition == "Blind") |>
  select(ir_structure, examiner_level, rate, replicate, reported, worst_case) |>
  pivot_longer(c(reported, worst_case), names_to = "measure", values_to = "value") |>
  mutate(measure = recode(measure, reported = "Reported", worst_case = "Worst case"))

ggplot(spread_data, aes(x = examiner_level, y = value, fill = measure)) +
  geom_boxplot(outlier.size = 0.4, position = position_dodge(width = 0.8)) +
  facet_grid(rate ~ ir_structure, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("Reported" = "grey85", "Worst case" = "#E69F00")) +
  labs(
    title = "Single-study reported and worst-case rates, by examiner heterogeneity",
    subtitle = "Blind condition, medium item heterogeneity; each box: 500 simulated studies",
    x = "Examiner heterogeneity", y = "Rate", fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Interpretation

```{r interp-numbers, include=FALSE}
rng <- function(x, digits = 1) {
  lo <- percent(min(x), accuracy = 10^-digits); hi <- percent(max(x), accuracy = 10^-digits)
  if (lo == hi) lo else paste0(lo, "--", hi)
}
rng_pp <- function(x) sprintf("%.0f--%.0f", 100 * min(x), 100 * max(x))

gs <- grid_summary
ratio_range <- function(d) sprintf("%.0f--%.0fx", min(d$truth_h25 / d$reported), max(d$truth_h25 / d$reported))

blind_sep <- gs |> filter(condition == "Blind", ir_structure == "Separate (current default)")
blind_shared <- gs |> filter(condition == "Blind", ir_structure == "Shared (previous default)")
known_sep <- gs |> filter(condition == "Known test", ir_structure == "Separate (current default)")

ex_effect <- gs |>
  filter(item_level == "Medium", condition == "Blind") |>
  group_by(examiner_level) |>
  summarise(inc = mean(inc_share), sd_worst = mean(sd_worst_case), .groups = "drop")
item_effect <- gs |>
  filter(examiner_level == "Medium", condition == "Blind", rate == "FPR") |>
  group_by(item_level) |>
  summarise(reported = mean(reported), .groups = "drop")
```

- **Reported rates stay small and the worst case is huge, everywhere in the grid.**
  Across all 36 combinations, the reported FPR is `r rng(gs$reported[gs$rate == "FPR"])` and the
  reported FNR `r rng(gs$reported[gs$rate == "FNR"])`, while the worst case ranges from
  `r rng(gs$worst_case)`. The interval between them is exactly the inconclusive share, and no setting of
  examiner or item heterogeneity brings it near zero. The reported number by itself says
  almost nothing about how large the true error rate could be.
- **Whether the true rate is "high" depends entirely on $h$, which the study cannot tell us.**
  If a quarter of inconclusives would have been errors, the forced-decision rate is
  `r ratio_range(gs |> filter(rate == "FPR"))` the reported FPR and
  `r ratio_range(gs |> filter(rate == "FNR"))` the reported FNR. False positive rates are affected most
  because their reported rate starts lowest. At $h = 10\%$ the truth is already several times the
  reported rate; nothing in the responses distinguishes $h = 0$ from $h = 1$.
- **(a) Separate inconclusive baselines change *where* the hidden error sits, not whether it exists.**
  With the previous shared 36% baseline, the blind worst cases are nearly symmetric
  (FPR `r rng(blind_shared$worst_case[blind_shared$rate == "FPR"], 0)`, FNR
  `r rng(blind_shared$worst_case[blind_shared$rate == "FNR"], 0)`). With the Ames II-style
  separate baselines they are not: the worst-case FPR is
  `r rng(blind_sep$worst_case[blind_sep$rate == "FPR"], 0)` but the worst-case FNR only
  `r rng(blind_sep$worst_case[blind_sep$rate == "FNR"], 0)`. The shared-baseline simulation
  in `case_study_inconclusive.qmd` therefore *understates* the ambiguity in false positives and *overstates* it in false negatives, relative to what the
  real study's inconclusive pattern implies.
- **(b) Examiner heterogeneity barely moves the average bounds, but changes how much a single study can be trusted.**
  Going from low to high examiner heterogeneity moves the mean inconclusive share by only
  `r sprintf("%.0f", 100 * (ex_effect$inc[ex_effect$examiner_level == "High"] - ex_effect$inc[ex_effect$examiner_level == "Low"]))`
  points on average in this cut, but the study-to-study SD of the worst case grows from about
  `r percent(ex_effect$sd_worst[ex_effect$examiner_level == "Low"], accuracy = 0.1)` to
  `r percent(ex_effect$sd_worst[ex_effect$examiner_level == "High"], accuracy = 0.1)` (see the spread plot).
  At the current default (Ames II-like) level of examiner heterogeneity, which 50 examiners take part
  visibly changes how wide a single study's interval is.
- **(c) Item heterogeneity mostly changes the *reported* rate, in the opposite direction from what one might hope.**
  Reported FPR rises from `r percent(item_effect$reported[item_effect$item_level == "Low"], accuracy = 0.1)` to
  `r percent(item_effect$reported[item_effect$item_level == "High"], accuracy = 0.1)` as item heterogeneity goes
  from low to high (a logit-scale Jensen effect noted in `analysis.qmd`), while the inconclusive share is essentially
  unchanged. So the *ratio* of the forced-decision rate to the reported rate is smaller when items are
  more heterogeneous, but the *level* of the true rate at a given $h$ is nearly the same.
- **(d) A known test widens the interval by about 10--16 points; in this model it does not lower the reported rate.**
  The Hawthorne shift adds
  `r sprintf("%.0f to %.0f", 100 * min(hawthorne$d_inc_share_min), 100 * max(hawthorne$d_inc_share_max))` points of inconclusive share (see the table above) and moves the
  reported rate by 0.0 points. The second half is built into the DGP
  (the shift acts only on the inconclusive probability), not a finding: it is the same simplification the
  contextual-bias backup slide notes. Two consequences follow. First, what the known-test study shows
  about casework is *different in kind* from what a blind study of the same examiners would show:
  its unresolved share is 
  `r sprintf("%.0f", 100 * mean(known_sep$inc_share[known_sep$rate == "FPR"]))`% for false positives against
  `r sprintf("%.0f", 100 * mean(blind_sep$inc_share[blind_sep$rate == "FPR"]))`% under blind conditions,
  so its worst-case bound is not the casework bound. Second, the mechanism the talk may want to invoke — examiners
  moving would-be errors into "inconclusive" when they know they are tested, so that the *reported* rate
  falls — is **not** in this model; if the extra inconclusives under a known test come disproportionately from would-be errors, the reported rate would drop and the hidden share would
  grow. That would need a selective mechanism (e.g., the Hawthorne shift scaling with item difficulty)
  and is not supported or ruled out here.

## Caveats

- $h$ is a scenario parameter with no empirical anchor; the grid shows what follows for
  $h \in \{0.1, 0.25, 0.5\}$, not which is right.
- All levels in the grid are illustrative. Ames II pins down the inconclusive baselines and
  the examiner inconclusive-tendency SD but not the item or skill SDs (see `fit_ames2.qmd`), and it also shows
  structure the DGP lacks (firearm-level variation, examiner behaviour that differs by ground truth) that
  would widen the study-to-study spread beyond what is shown here.
- The error baselines (FPR `r percent(fixed$fpr, accuracy = 0.1)`, FNR `r percent(fixed$fnr, accuracy = 0.1)`),
  the 50-examiner, 30-of-300-item design, and the 50/50 match rate
  are fixed at the talk's current defaults; the real study is one-third same-source.

# How to use this in the talk

- **The message this supports:** a validity study's reported error rate and its
  worst case can be an order of magnitude or more apart, that gap is exactly the inconclusive share, and the
  study's own responses cannot narrow it. It does not depend on how heterogeneous examiners or items are, and it
  survives switching between the previous shared baseline and the current Ames II-style separate baselines.
- **The headline plot** (Ames II-like settings, four rows) is the candidate slide: filled dots stay near
  zero while the open dots (worst case) sit at roughly 25--65%.
- **On the Hawthorne effect, be precise about what is shown:** a known test raises the inconclusive share by
  about 10--16 points, so the study's unresolved mass no longer matches casework's. This model does *not* show a
  known test lowering the reported error rate; do not say it does unless a selective mechanism is added.
- **If asked "isn't $h$ made up?":** yes, and that is the point — the study cannot estimate it. The
  claim is not that the true error rate *is* (say) 13%, it is that a reported false positive rate of about 1% is compatible with anything up to
  about 50%.

# References

::: {#refs}
:::
