---
title: "Data Simulation for Firearms Validity Studies"
format: html
editor: source
---

## Goal

The purpose of this study is to simulate what could happen due to the flaws outlined in this article:

Cuellar, M., Vanderplas, S., Luby, A., & Rosenblum, M. (2024). Methodological problems in every black-box study of forensic firearm comparisons. *Law, Probability and Risk*, 23(1), mgae015.

## Data generation

### Number of examiners and comparisons

First, install the packages used in the analysis if they are not already available on the machine.

```{r load-packages}
# Preamble
list.of.packages <- c("tidyverse", "ggthemes")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
```

Now define the main simulation settings, including the study size, the baseline response rates, and the amount of examiner and question heterogeneity.

```{r set-simulation-parameters}
set.seed(123)

# Parameters
n_examiners <- 50
n_comparisons <- 100
match_rate <- 0.5  # 50% same-source comparisons

# Baseline response probabilities
false_positive_rate <- 0.02
false_negative_rate <- 0.05
inconclusive_rate <- 0.1

# Heterogeneity parameters
examiner_sd <- 0.7
question_sd <- 0.8
```

### Question difficulty

Now generate the set of underlying comparisons. Each question has a fixed truth label and a difficulty value that is shared across examiners.

```{r generate-comparison-set}
comparison_set <- tibble(
  question_id = 1:n_comparisons,
  ground_truth = rbinom(n_comparisons, 1, match_rate),
  question_difficulty = rnorm(n_comparisons, mean = 0, sd = question_sd)
)
```

Now generate the examiner panel. Each examiner gets a latent skill value and a separate tendency to call a comparison inconclusive.

### Examiner performance

```{r generate-examiner-panel}
examiner_panel <- tibble(
  examiner_id = paste0("E", 1:n_examiners),
  examiner_skill = rnorm(n_examiners, mean = 0, sd = examiner_sd),
  examiner_inconclusive_tendency = rnorm(n_examiners, mean = 0, sd = examiner_sd / 2)
)
```

Next, combine the shared question set with the examiner panel so that every examiner sees the same questions. Then compute a challenge score for each examiner-question pair.

### Assigning questions to examiners

```{r assign-questions-to-examiners}
sim_test <- tidyr::crossing(examiner_panel, comparison_set) %>%
  mutate(
    ground_truth_label = if_else(ground_truth == 1, "same-source", "different-source"),
    decision_challenge = question_difficulty - examiner_skill
  ) %>%
  arrange(examiner_id, question_id)
```

At this point, inspect the structure of the simulated test before any responses are generated.

```{r preview-simulated-test}
head(sim_test)
dim(sim_test) # there are 50 examiners, with 100 questions each, for 5000 total responses
```

### Simulated responses

Now define the response model for a single examiner-question pair. The baseline false positive, false negative, and inconclusive rates are treated as average rates in the population. For each examiner-question pair, those baseline probabilities are moved up or down on the log-odds scale according to the challenge of that pairing, so harder questions and weaker examiners lead to higher error probabilities while easier questions and stronger examiners lead to lower ones.

```{r define-response-model}
# Response model for a single examiner-comparison pair.
simulate_response <- function(ground_truth,
                              decision_challenge,
                              examiner_inconclusive_tendency,
                              false_positive_rate,
                              false_negative_rate,
                              inconclusive_rate) {
  if (ground_truth == 1) {
    error_probability <- plogis(qlogis(false_negative_rate) + decision_challenge)
  } else {
    error_probability <- plogis(qlogis(false_positive_rate) + decision_challenge)
  }

  inconclusive_probability <- plogis(
    qlogis(inconclusive_rate) +
      0.5 * decision_challenge +
      examiner_inconclusive_tendency
  )

  total_probability <- error_probability + inconclusive_probability

  if (total_probability >= 0.95) {
    scaling_factor <- 0.95 / total_probability
    error_probability <- error_probability * scaling_factor
    inconclusive_probability <- inconclusive_probability * scaling_factor
  }

  p <- runif(1)

  if (ground_truth == 1) {
    if (p < error_probability) {
      "elimination"
    } else if (p < error_probability + inconclusive_probability) {
      "inconclusive"
    } else {
      "identification"
    }
  } else {
    if (p < error_probability) {
      "identification"
    } else if (p < error_probability + inconclusive_probability) {
      "inconclusive"
    } else {
      "elimination"
    }
  }
}
```

Next, apply the response model to every examiner-question pair and generate the simulated examiner decisions.

```{r generate-simulated-responses}
sim_data <- sim_test %>%
  mutate(
    response = purrr::pmap_chr(
      list(
        ground_truth = ground_truth,
        decision_challenge = decision_challenge,
        examiner_inconclusive_tendency = examiner_inconclusive_tendency
      ),
      simulate_response,
      false_positive_rate = false_positive_rate,
      false_negative_rate = false_negative_rate,
      inconclusive_rate = inconclusive_rate
    )
  )
```

Then inspect a small, readable slice of the simulated responses to check whether the generated data look sensible. The table below shows the first several rows from the first few examiners.

```{r preview-simulated-responses}
sim_data_preview <- sim_data %>%
  filter(examiner_id %in% paste0("E", 1:3)) %>%
  group_by(examiner_id) %>%
  slice_head(n = 6) %>%
  ungroup() %>%
  select(
    examiner_id,
    question_id,
    ground_truth_label,
    question_difficulty,
    examiner_skill,
    decision_challenge,
    response
  ) %>%
  mutate(
    question_difficulty = round(question_difficulty, 2),
    examiner_skill = round(examiner_skill, 2),
    decision_challenge = round(decision_challenge, 2)
  )

knitr::kable(
  sim_data_preview,
  col.names = c(
    "Examiner",
    "Question",
    "Truth",
    "Question difficulty",
    "Examiner skill",
    "Decision challenge",
    "Response"
  ),
  align = c("l", "r", "l", "r", "r", "r", "l")
)
```

Each row in this table is one examiner-question pair. The columns identify the examiner, the question, the true status of the comparison, the latent difficulty of the question, the latent skill and inconclusive tendency of the examiner, the overall challenge of that pairing, and the simulated response that the examiner gave.

### What the simulated data look like

These summary statistics describe the simulated dataset as a whole, including the number of rows, the number of unique examiners and questions, and the distribution of truth labels and responses.

```{r summarize-simulated-data}
sim_data_summary <- tibble(
  quantity = c(
    "Rows in simulated dataset",
    "Examiners",
    "Questions",
    "Same-source comparisons",
    "Different-source comparisons",
    "Identification responses",
    "Elimination responses",
    "Inconclusive responses",
    "Mean examiner skill",
    "Mean question difficulty",
    "Mean decision challenge"
  ),
  value = c(
    nrow(sim_data),
    n_distinct(sim_data$examiner_id),
    n_distinct(sim_data$question_id),
    mean(sim_data$ground_truth == 1),
    mean(sim_data$ground_truth == 0),
    mean(sim_data$response == "identification"),
    mean(sim_data$response == "elimination"),
    mean(sim_data$response == "inconclusive"),
    mean(sim_data$examiner_skill),
    mean(sim_data$question_difficulty),
    mean(sim_data$decision_challenge)
  )
) %>%
  mutate(
    value = case_when(
      quantity %in% c(
        "Rows in simulated dataset",
        "Examiners",
        "Questions"
      ) ~ as.character(round(value, 0)),
      grepl("Mean", quantity) ~ sprintf("%.2f", value),
      TRUE ~ scales::percent(value, accuracy = 0.1)
    )
  )

sim_data_summary
```

These plots give a quick view of the simulated dataset. The first two plots show latent simulation quantities rather than directly observed measurements: examiner skill is a relative skill parameter, question difficulty is a relative difficulty parameter, and both are plotted on the latent scale used to generate the response probabilities. The third plot shows the mix of simulated response types.

```{r plot-examiner-skill}
ggplot(examiner_panel, aes(x = examiner_skill)) +
  geom_histogram(bins = 15) +
  labs(
    title = "Distribution of examiner skill",
    x = "Relative examiner skill (latent scale)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r plot-question-difficulty}
ggplot(comparison_set, aes(x = question_difficulty)) +
  geom_histogram(bins = 15) +
  labs(
    title = "Distribution of question difficulty",
    x = "Relative question difficulty (latent scale)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r plot-response-distribution}
sim_data %>%
  count(response) %>%
  ggplot(aes(x = response, y = n, fill = response)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Distribution of simulated responses",
    x = "Response",
    y = "Count"
  ) +
  theme_minimal()
```

## Evaluation of performance

Now convert the raw responses into performance indicators at the comparison level, including whether the response was correct, false positive, false negative, true positive, true negative, or inconclusive.

```{r create-question-level-metrics}
question_level_metrics <- sim_data %>%
  mutate(
    correct = case_when(
      ground_truth == 1 & response == "identification" ~ TRUE,
      ground_truth == 0 & response == "elimination" ~ TRUE,
      response == "inconclusive" ~ NA,
      TRUE ~ FALSE
    ),
    is_inconclusive = response == "inconclusive",
    is_fp = response == "identification" & ground_truth == 0,
    is_fn = response == "elimination" & ground_truth == 1,
    is_tp = response == "identification" & ground_truth == 1,
    is_tn = response == "elimination" & ground_truth == 0
  )

# Preview of metrics for each comparison
head(question_level_metrics)

```

### Performance by examiner

Next, summarize those comparison-level outcomes for each examiner so you can compare performance across the panel.

```{r summarize-performance-by-examiner}
# Summarizing performance by examiner.
examiner_performance <- question_level_metrics %>%
  group_by(examiner_id) %>%
  summarise(
    n_comparisons = n(),
    accuracy = mean(correct, na.rm = TRUE),
    inconclusive_rate = mean(is_inconclusive, na.rm = TRUE),
    false_positive_rate = mean(is_fp, na.rm = TRUE),
    false_negative_rate = mean(is_fn, na.rm = TRUE),
    true_positive_rate = mean(is_tp, na.rm = TRUE),
    true_negative_rate = mean(is_tn, na.rm = TRUE),
    .groups = "drop"
  )

# Preview the summarized data by examiner
head(examiner_performance)
```

**NOTE**: We should probably provide a likelihood ratio here instead of the FPR, FNR, etc. ALso, note how the FPR and FNR are being calculated regarding the inconclusives.

Now show performance across all examiners.

### Performance across examiners

Finally, average the examiner-level summaries to produce an overall description of the simulated study.

```{r summarize-overall-performance}
# Summarizing performance across all examiners
overall_summary <- examiner_performance %>%
  summarise(
    mean_accuracy = mean(accuracy, na.rm = TRUE),
    mean_inconclusive_rate = mean(inconclusive_rate, na.rm = TRUE),
    mean_false_positive_rate = mean(false_positive_rate, na.rm = TRUE),
    mean_false_negative_rate = mean(false_negative_rate, na.rm = TRUE),
    mean_true_positive_rate = mean(true_positive_rate, na.rm = TRUE),
    mean_true_negative_rate = mean(true_negative_rate, na.rm = TRUE)
  )

# Preview the overall summary
overall_summary
```

# Consequences of having flaws A-F from Cuellar et al. 2024

# A. Inadequate sample size

Small sample size matters because it does not merely make error-rate estimates less precise. It can also make them look more reassuring than the design justifies. With too few examiners or too few comparisons, a study can easily observe very few errors, or even no errors at all, simply by chance.

To illustrate that problem, the next analysis repeats the entire study many times at three different sizes. For each simulated study, it calculates pooled false positive and false negative rates, approximate confidence interval widths, and whether the study observed zero false positives. The point is not to recover one correct estimate, but to show how unstable the reported estimates are when the study is too small.

### Repeated simulated studies at different sample sizes

First, define a function that simulates one complete study at a given number of examiners and comparisons, using the same data-generating process introduced above.

```{r define-study-simulation-function}
simulate_study <- function(n_examiners, n_comparisons) {
  comparison_set <- tibble(
    question_id = 1:n_comparisons,
    ground_truth = rbinom(n_comparisons, 1, match_rate),
    question_difficulty = rnorm(n_comparisons, mean = 0, sd = question_sd)
  )

  examiner_panel <- tibble(
    examiner_id = paste0("E", 1:n_examiners),
    examiner_skill = rnorm(n_examiners, mean = 0, sd = examiner_sd),
    examiner_inconclusive_tendency = rnorm(n_examiners, mean = 0, sd = examiner_sd / 2)
  )

  sim_test <- tidyr::crossing(examiner_panel, comparison_set) %>%
    mutate(
      decision_challenge = question_difficulty - examiner_skill
    )

  sim_test %>%
    mutate(
      response = purrr::pmap_chr(
        list(
          ground_truth = ground_truth,
          decision_challenge = decision_challenge,
          examiner_inconclusive_tendency = examiner_inconclusive_tendency
        ),
        simulate_response,
        false_positive_rate = false_positive_rate,
        false_negative_rate = false_negative_rate,
        inconclusive_rate = inconclusive_rate
      )
    )
}
```

Now define a function that summarizes one simulated study. Here the false positive and false negative rates are calculated with the appropriate denominators: non-matches for the false positive rate and matches for the false negative rate.

```{r define-study-summary-function}
summarize_study <- function(study_data) {
  nonmatch_total <- sum(study_data$ground_truth == 0)
  match_total <- sum(study_data$ground_truth == 1)

  false_positives <- sum(study_data$response == "identification" & study_data$ground_truth == 0)
  false_negatives <- sum(study_data$response == "elimination" & study_data$ground_truth == 1)
  inconclusives <- sum(study_data$response == "inconclusive")

  fpr <- false_positives / nonmatch_total
  fnr <- false_negatives / match_total
  inconclusive_rate_study <- inconclusives / nrow(study_data)

  fpr_ci <- prop.test(false_positives, nonmatch_total)$conf.int
  fnr_ci <- prop.test(false_negatives, match_total)$conf.int

  tibble(
    fpr = fpr,
    fnr = fnr,
    inconclusive_rate = inconclusive_rate_study,
    fpr_ci_width = diff(fpr_ci),
    fnr_ci_width = diff(fnr_ci),
    zero_false_positives = false_positives == 0,
    nonmatch_total = nonmatch_total,
    match_total = match_total
  )
}
```

Now specify three study sizes. The small study has few examiners and few comparisons, the medium study is larger but still limited, and the large study is closer to the baseline design used above.

```{r define-sample-size-scenarios}
sample_size_scenarios <- tribble(
  ~scenario, ~n_examiners, ~n_comparisons,
  "Small", 10, 20,
  "Medium", 25, 50,
  "Large", 50, 100
)
```

Now simulate many studies under each scenario. Repeating the study many times shows the range of error-rate estimates that one could easily obtain from the same underlying process.

```{r run-sample-size-simulations}
set.seed(456)

n_replications <- 200

sample_size_results <- sample_size_scenarios %>%
  mutate(
    results = purrr::map2(
      n_examiners,
      n_comparisons,
      function(n_examiners, n_comparisons) {
        purrr::map_dfr(
          1:n_replications,
          function(replicate_id) {
            simulate_study(n_examiners, n_comparisons) %>%
              summarize_study() %>%
              mutate(replicate_id = replicate_id)
          }
        )
      }
    )
  ) %>%
  select(scenario, n_examiners, n_comparisons, results) %>%
  tidyr::unnest(results)
```

These summary statistics show how the estimated error rates and uncertainty behave across repeated studies at each sample size.

```{r summarize-sample-size-results}
sample_size_summary <- sample_size_results %>%
  group_by(scenario, n_examiners, n_comparisons) %>%
  summarise(
    mean_fpr = mean(fpr),
    sd_fpr = sd(fpr),
    mean_fnr = mean(fnr),
    sd_fnr = sd(fnr),
    mean_fpr_ci_width = mean(fpr_ci_width),
    mean_fnr_ci_width = mean(fnr_ci_width),
    proportion_zero_false_positives = mean(zero_false_positives),
    .groups = "drop"
  ) %>%
  mutate(
    mean_fpr = scales::percent(mean_fpr, accuracy = 0.1),
    sd_fpr = scales::percent(sd_fpr, accuracy = 0.1),
    mean_fnr = scales::percent(mean_fnr, accuracy = 0.1),
    sd_fnr = scales::percent(sd_fnr, accuracy = 0.1),
    mean_fpr_ci_width = scales::percent(mean_fpr_ci_width, accuracy = 0.1),
    mean_fnr_ci_width = scales::percent(mean_fnr_ci_width, accuracy = 0.1),
    proportion_zero_false_positives = scales::percent(proportion_zero_false_positives, accuracy = 0.1)
  )

knitr::kable(
  sample_size_summary,
  col.names = c(
    "Scenario",
    "Examiners",
    "Comparisons",
    "Mean FPR",
    "SD of FPR",
    "Mean FNR",
    "SD of FNR",
    "Mean FPR CI width",
    "Mean FNR CI width",
    "Studies with zero observed false positives"
  )
)
```

The next plot shows the distribution of false positive rate estimates across the repeated studies. Small studies produce much more variable estimates, including many studies that report no false positives at all.

```{r plot-fpr-by-sample-size}
ggplot(sample_size_results, aes(x = scenario, y = fpr)) +
  geom_boxplot() +
  labs(
    title = "Estimated false positive rates across repeated studies",
    x = "Study size scenario",
    y = "Estimated false positive rate"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal()
```

This plot shows the same pattern for false negative rates. Again, the smallest studies produce the most unstable estimates.

```{r plot-fnr-by-sample-size}
ggplot(sample_size_results, aes(x = scenario, y = fnr)) +
  geom_boxplot() +
  labs(
    title = "Estimated false negative rates across repeated studies",
    x = "Study size scenario",
    y = "Estimated false negative rate"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal()
```

This final plot shows how the average width of the confidence intervals shrinks as the study gets larger. Small studies do not simply have noisier estimates; they also carry much greater uncertainty.

```{r plot-ci-width-by-sample-size}
sample_size_results %>%
  select(scenario, fpr_ci_width, fnr_ci_width) %>%
  pivot_longer(
    cols = c(fpr_ci_width, fnr_ci_width),
    names_to = "metric",
    values_to = "ci_width"
  ) %>%
  mutate(
    metric = recode(
      metric,
      fpr_ci_width = "False positive rate",
      fnr_ci_width = "False negative rate"
    )
  ) %>%
  ggplot(aes(x = scenario, y = ci_width)) +
  geom_boxplot() +
  facet_wrap(~ metric) +
  labs(
    title = "Confidence interval widths across repeated studies",
    x = "Study size scenario",
    y = "Confidence interval width"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal()
```

Taken together, these simulations show that inadequate sample size can produce error-rate estimates that appear reassuring simply because the study is too small to reveal the underlying variability. A small study may report very low observed error rates, or even zero observed false positives, without providing strong evidence that the true error rate is comparably low.

B. Non-representative sample AMANDA

C. Non-representative testing conditions and environment (Contextual bias) MARIA

Use Scurich and Dror's numbers to show bias here.

D. Inconclusive responses are treated as correct or ignored e.g., If you treat them as correct the FPR is 0.1%, if you treat them as incorrec it's 90% etc.

E. Invalid or nonexistent uncertainty measures for error rates This is just a statistical requirement. No need for a simulation.

Note: Look at Hicklin 2024 bootstrap sample (still invalid because non-representative sample, but we can use it to compare).

F. Missing data MARIA

A correctly done study
