---
title: "Data Simulation for Firearms Validity Studies"
format: html
editor: visual
---


## Goal

The purpose of this study is to simulate what could happen due to the flaws outlined in this article:

Cuellar, M., Vanderplas, S., Luby, A., & Rosenblum, M. (2024). Methodological problems in every black-box study of forensic firearm comparisons. *Law, Probability and Risk*, 23(1), mgae015.

## Data generation


```{r}
# Preamble
list.of.packages <- c("tidyverse", "ggthemes")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
```


**First**, let there be a fixed number of examiners (50) and comparisons (100), and let the proporion of same-source comparisons be 0.5.

```{r}

set.seed(123)

# Parameters
n_examiners <- 50
n_comparisons <- 100
match_rate <- 0.5  # 50% same-source comparisons

# Error probabilities
false_positive_rate <- 0.02
false_negative_rate <- 0.05
inconclusive_rate <- 0.1

# Create empty data frame
library(dplyr)

```


Now design the test given to examiners.




```{r}


# Simulate the test given to examiners, with ground truth.
sim_test <- expand.grid(
  examiner_id = paste0("E", 1:n_examiners),
  question_id = 1:n_comparisons
) %>%
  mutate(
    ground_truth = rbinom(n(), 1, match_rate),  # 1 = same-source, 0 = different-source
    response = NA_character_
  ) %>%
  arrange(examiner_id)  # Sorting by examiner_id

# View a sample
head(sim_test)
dim(sim_test) # there are 50 examiners, with 100 questions each, for 5000 total responses

```


Now design the responses given by examiners.

```{r}

# Simulate the answers given by examiners.
sim_data <- sim_test %>%
  rowwise() %>%
  mutate(response = {
    if (ground_truth == 1) {
      # True match
      p <- runif(1)
      if (p < false_negative_rate) "elimination"
      else if (p < false_negative_rate + inconclusive_rate) "inconclusive"
      else "identification"
    } else {
      # True non-match
      p <- runif(1)
      if (p < false_positive_rate) "identification"
      else if (p < false_positive_rate + inconclusive_rate) "inconclusive"
      else "elimination"
    }
  })

head(sim_data)

```


## Evaluation of performance

First let's look at the responses for all examiners. Were they correct (TRUE) or incorrect (FALSE)?

```{r}


question_level_metrics <- sim_data %>%
  mutate(
    correct = case_when(
      ground_truth == 1 & response == "identification" ~ TRUE,
      ground_truth == 0 & response == "elimination" ~ TRUE,
      response == "inconclusive" ~ NA,
      TRUE ~ FALSE
    ),
    is_inconclusive = response == "inconclusive",
    is_fp = response == "identification" & ground_truth == 0,
    is_fn = response == "elimination" & ground_truth == 1,
    is_tp = response == "identification" & ground_truth == 1,
    is_tn = response == "elimination" & ground_truth == 0
  )

# Preview of metrics for each comparison
head(question_level_metrics)

```



### Performance by examiner


Now let's evaluate the performance for each examiner.

```{r}
# Summarizing performance by examiner.
examiner_performance <- question_level_metrics %>%
  group_by(examiner_id) %>%
  summarise(
    n_comparisons = n(),
    accuracy = mean(correct, na.rm = TRUE),
    inconclusive_rate = mean(is_inconclusive, na.rm = TRUE),
    false_positive_rate = mean(is_fp, na.rm = TRUE),
    false_negative_rate = mean(is_fn, na.rm = TRUE),
    true_positive_rate = mean(is_tp, na.rm = TRUE),
    true_negative_rate = mean(is_tn, na.rm = TRUE),
    .groups = "drop"
  )

# Preview the summarized data by examiner
head(examiner_performance)
```

**NOTE**: We should probably provide a likelihood ratio here instead of the FPR, FNR, etc. ALso, note how the FPR and FNR are being calculated regarding the inconclusives.

Now show performance across all examiners.


### Performance across examiners


We can see the performance overall if we average across examiners.

```{r}

# Summarizing performance across all examiners
overall_summary <- examiner_performance %>%
  summarise(
    mean_accuracy = mean(accuracy, na.rm = TRUE),
    mean_inconclusive_rate = mean(inconclusive_rate, na.rm = TRUE),
    mean_false_positive_rate = mean(false_positive_rate, na.rm = TRUE),
    mean_false_negative_rate = mean(false_negative_rate, na.rm = TRUE),
    mean_true_positive_rate = mean(true_positive_rate, na.rm = TRUE),
    mean_true_negative_rate = mean(true_negative_rate, na.rm = TRUE)
  )

# Preview the overall summary
overall_summary
```


# Consequences of having flaws A-F from Cuellar et al. 2024


# A. Inadequate sample size

Below we show how the accuracy measures vary with varying sample size.

Idea: Do several simulations to show how even if you have many studies (10, 50, 1000!), if they all have small sample size, they will give you incorrect results.

Sample size for examiners AND bullet/cartridge samples.


### An example sample size estimation

```{r}

n_small=10
n_medium=100
n_large=500


p=0.95 # sensitivity
E=0.03 # desired margin of error
Z=1.96 # confidence level: 95%

n1=(Z^2 * p*(1-p))/ E^2
n1

p2=0.85 # specificity
n2=(Z^2 * p2*(1-p2))/ E^2
n2

# Adjusting for Examiner Clustering (Design Effect)

rho=0.05 # intra-class correlation
m=20 # number of trials per examiner

DE = 1+(m-1)*rho # desired effect
DE

matches = n1*DE
nonmatches = n2*DE
trials=matches*2

matches
nonmatches
trials

n_examiners = trials/20
n_examiners

#Summary
# Estimated sample size: ~400 known match comparisons and ~1,100 known non-match comparisons
# Assuming 20 trials per examiner, recruit ~40–60 examiners
# Design accounts for examiner clustering and ±3% margin of error



```




B. Non-representative sample
AMANDA

C. Non-representative testing conditions and environment (Contextual bias)
MARIA

Use Scurich and Dror's numbers to show bias here.

D. Inconclusive responses are treated as correct or ignored
e.g., If you treat them as correct the FPR is 0.1%, if you treat them as incorrec it's 90% etc.

E. Invalid or nonexistent uncertainty measures for error rates
This is just a statistical requirement. No need for a simulation.

Note: Look at Hicklin 2024 bootstrap sample (still invalid because non-representative sample, but we can use it to compare).

F. Missing data
MARIA

A correctly done study